### 2024-11-04 README.Rmd
package files

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# canaccessR

<!-- badges: start -->
<!-- badges: end -->

Package {canaccessR} is an [Open Data Product](https://rdcu.be/dY1Ic) with travel time tables to employment and groceries in selected Canadian Metropolitan Areas.

The package was initially developed for a writing project commissioned by the [School of Cities](https://schoolofcities.utoronto.ca) of the University of Toronto, and can be used to replicate the results reported in the paper.

## Installation

You can install the development version of canaccessR like so:

``` r
# install.packages("remotes")
remotes::install_github("paezha/canaccessR")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r load-packages}
library(canaccessR)
library(ggplot2)
library(sf)
```

Load the travel time table to employment for Halifax:
```{r load-data}
data(travel_matrix_emp_hal)
```

The median travel time in minutes from all origins to all destinations in Halifax is:
```{r}
median(travel_matrix_emp_hal$travel_time_p50)
```

Simple features objects for cartography are also included. For example, this is Quebec:
```{r}
data("census_data_da_que")
```

Plot the population by Dissemination Area:
```{r}
ggplot() +
  geom_sf(data = census_data_da_que,
          aes(fill = Population)) +
  scale_fill_fermenter(palette = "Reds",
                       direction = 1) +
  theme_minimal()
```

### 2024-11-04 data-raw/00_census_data.qmd
package files

---
title: "00_census_data"
format: html
---

```{r}
#| label: load packages
library(cancensus) # Access, Retrieve, and Work with Canadian Census Data and Geography
library(glue) # Interpreted String Literals
library(here) # A Simpler Way to Find Your Files
library(furrr) # Apply Mapping Functions in Parallel using Futures
library(fs) # Cross-Platform File System Operations Based on 'libuv'
library(sf) # Simple Features for R
library(tidyverse) # Easily Install and Load the 'Tidyverse'

cache_path <- fs::dir_create("./cache")
data_path <- fs::dir_create("./data")
data_raw_path <- fs::dir_create("./data-raw")

# See here for instructions to obtain an API key: https://mountainmath.github.io/cancensus/index.html#api-key
options(cancensus.api_key = "YOUR_KEY")
options(cancensus.cache_path = cache_path)
```

# Census Data
## Define CMA lists for iterative getting of data

```{r}
# list of CMA IDs for cancensus
cma_geouid_list <- c(
  "Toronto" = "35535", 
  "Montréal" = "24462", 
  "Vancouver" = "59933",
  "Calgary" = "48825",
  "Ottawa" = "505",
  "Edmonton" = "48835",
  "Québec City" = "24421",
  "Winnipeg" = "46602",
  "Hamilton" = "35537",
  "Waterloo" = "35541",
  "London" = "35555",
  #"Niagara" = "35539",
  "Halifax" = "12205",
  "Oshawa" = "35532",
  "Abbotsford" = "59932")

# create list of analysis regions - collapsing some into others e.g. oshawa into toronto
study_region_names_list <- c(
  "Toronto" = "35535", 
  "Montréal" = "24462", 
  "Vancouver" = "59933",
  "Calgary" = "48825",
  "Ottawa" = "505",
  "Edmonton" = "48835",
  "Québec City" = "24421",
  "Winnipeg" = "46602",
  "Hamilton" = "35537",
  "Waterloo" = "35541",
  "London" = "35555",
  #"Niagara" = "35539",
  "Halifax" = "12205",
  "Toronto" = "35532",
  "Vancouver" = "59932") |> 
  enframe() |> 
  select(2:1) |> 
  deframe()

# ordered factor of study regions for figures and mapping
study_region_names_order_list <- list(
  "Toronto" = "gta", 
  "Montréal" = "mtl", 
  "Vancouver" = "van",
  "Calgary" = "cal",
  "Ottawa" = "ott",
  "Edmonton" = "edm",
  "Québec City" = "que",
  "Winnipeg" = "win",
  "Hamilton" = "ham",
  "Waterloo" = "wat",
  "London" = "ldn",
  "Halifax" = "hal") |> 
  enframe() |> 
  select(2:1) |> 
  deframe()

# define analysis regions - use to collapse CMAs into larger analysis regions
r5r_analysis_region_list <- c(
  "ggh" = "35535", 
  "mtl" = "24462", 
  "van" = "59933",
  "cal" = "48825",
  "ott" = "505",
  "edm" = "48835",
  "que" = "24421",
  "win" = "46602",
  "ggh" = "35537",
  "ggh" = "35541",
  "ldn" = "35555",
  "ggh" = "35539",
  "hal" = "12205",
  "ggh" = "35532",
  "van" = "59932") |> 
  enframe() |> 
  select(2:1) |> 
  deframe()

r5r_named_networks <- list(
  "ggh" = "ggh", 
  "mtl" = "mtl", 
  "van" = "van",
  "cal" = "cal",
  "ott" = "ott",
  "edm" = "edm",
  "que" = "que",
  "win" = "win",
  "ldn" = "ldn",
  "hal" = "hal")
```

## Define MJ Census Vectors for 2016 Census

```{r}
mj_vectors_2016 <- c(
  "tot_age_popn" = "v_CA16_1",
  "age_0_to_14" = "v_CA16_4",
  "age_15_to_64" = "v_CA16_61",
  "age_65_plus" = "v_CA16_244",
  "tot_cens_fam_priv_hh" = "v_CA16_484",
  "fem_parent" = "v_CA16_489",
  "LICO_AT" = "v_CA16_2555", 
  "tot_LICO_status_2015" = "v_CA16_2510",
  "knwl_off_lang" = "v_CA16_512",
  "neither_EN_FR" = "v_CA16_524",
  "Aborin_ID_priv_HH" = "v_CA16_3852",
  "Aborin_ID" = "v_CA16_3855",
  "tot_priv_hh_hsing_suit" = "v_CA16_4859",
  "not_suitable" = "v_CA16_4861",
  "dwell_condtns" = "v_CA16_4870",
  "maj_rep_needed" = "v_CA16_4872",
  "tot_shelt_cost" = "v_CA16_4886",
  "shelt_gt30pc" = "v_CA16_4888",
  "tot_tenant_HH" = "v_CA16_4897",
  "tenant_HH_subsid" = "v_CA16_4898",
  "tot_immig_stat" = "v_CA16_3405",
  "immig_stat_last5" = "v_CA16_3432",
  "tot_adm_immig_popn_1980_to_2016" = "v_CA16_3831",
  "refugees" = "v_CA16_3846",
  "visible_min" = "v_CA16_3957",
  "tot_visible_min" = "v_CA16_3954",
  "Black" = "v_CA16_3966",
  "tot_lbr_f_15_plus" = "v_CA16_5597",
  "lbr_f_unemp" = "v_CA16_5606",
  "tot_commuting_dur_lbr_f_15_plus" = "v_CA16_5813",
  "commuting_dur_60mins_plus" = "v_CA16_5828",
  "ave_HH_size" = "v_CA16_425",
  "tot_degree_age_15_plus" = "v_CA16_5051",
  "bachelor_or_higher" = "v_CA16_5078",
  "mhh_inc" = "v_CA16_2397",
  "popn_dens" = "v_CA16_406",
  "from_mode_total" = "v_CA16_5792",
  "from_mode_car" = "v_CA16_5795",
  "from_mode_pass" = "v_CA16_5798",
  "from_mode_tran" = "v_CA16_5801",
  "from_mode_walk" = "v_CA16_5804")
```

## Get CMA Data and Geometries

```{r}
census_data_cma <- map_dfr(
  cma_geouid_list, 
  ~ get_census(
  dataset = 'CA16', 
  regions = list(CMA = c(.x)),
  vectors = mj_vectors_2016,
  level = 'CMA',
  geo_format = "sf",
  use_cache = TRUE)) |> 
  mutate(
    r5r_analysis_region = as.factor(r5r_analysis_region_list[GeoUID]),
    study_region_name = as.factor(study_region_names_list[GeoUID]),
    study_region_name = fct_relevel(study_region_name, study_region_names_order_list)
  )
```

join water erased geometries

```{r}
cmas_no_water <- st_read(glue::glue(here::here(), 
                                    "/data-raw/census_data_cma_erased.shp"))

census_data_cma <- census_data_cma |> 
  st_drop_geometry() |> 
  left_join(cmas_no_water, by = "GeoUID") |> 
  st_as_sf() |> 
  st_make_valid() |> 
  st_transform(crs = 4326) 
```

## Get DA Data and Geometries

```{r}
census_data_da <- map_dfr(
  cma_geouid_list, 
  ~ get_census(
  dataset = 'CA16', 
  regions = list(CMA = c(.x)),
  vectors = mj_vectors_2016,
  level = 'DA',
  geo_format = "sf",
  use_cache = TRUE)) |> 
  mutate(
    study_region_name = study_region_names_list[CMA_UID],
    study_region_name = fct_relevel(study_region_name, study_region_names_order_list),
    r5r_analysis_region = r5r_analysis_region_list[CMA_UID]
    )
```

```{r}
das_no_water <- st_read(glue::glue(here::here(), 
                                    "/data-raw/census_data_da_erased.shp"))

# DA_EMP_2016_Canada.csv is a confidential file
da_employment <- read_csv(glue::glue(here::here(), 
                                    "/data/DA_EMP_2016_Canada.csv")) |> 
  select(GeoUID, Total_Emp) |> 
  mutate(GeoUID = as.character(GeoUID))

census_data_da <- census_data_da |> 
  left_join(da_employment |> select(GeoUID, Total_Emp), by = "GeoUID") |> 
  mutate(Total_Emp = replace_na(Total_Emp, 0)) |> 
  st_drop_geometry() |> 
  left_join(das_no_water, by = "GeoUID") |>
  st_as_sf() |> 
  st_make_valid() |> 
  st_transform(crs = 4326)
```

## Region Backgrounds for Mapping

```{r}
region_backgrounds <- st_read(glue::glue(here::here(), 
                                    "/data-raw/pr_region_backgrounds.shp")) |> 
  rename(region_name = regn_nm)
```

## Save Census Data

```{r}
# save(
#   census_data_da, 
#   , 
#   region_backgrounds, 
#   file = "./data/census_data.Rdata")
```

Regions used for routing in {r5r}:
```{r}
regions <- levels(census_data_cma$r5r_analysis_region)
```

Save CMA data by region:
```{r}

for(i in regions){
  census_data_cma |>
    rename(region = r5r_analysis_region) |>
    filter(region == i) |>
    select(Type, GeoUID, name, region) |>
    assign(x = glue::glue("census_data_cma_{i}"),
         value = _)

  save(list = glue::glue("census_data_cma_{i}"),
       file = glue::glue(here::here(),
                       "/data/census_data_cma_{i}.rda"),
     compress = "xz")
}

```

Save DA data by region:
```{r}
for(i in regions){
  census_data_da |>
    rename(region = r5r_analysis_region) |>
    filter(region == i) |>
    select(-c(`Shape Area`, `Region Name`, `Area (sq km)`, popn_dens, `study_region_name`)) |>
    select(Type, GeoUID, region, everything()) |>
    assign(x = glue::glue("census_data_da_{i}"),
         value = _)

  save(list = glue::glue("census_data_da_{i}"),
       file = glue::glue(here::here(),
                       "/data/census_data_da_{i}.rda"),
     compress = "xz")
}

```

Save backgrounds by region:
```{r}
for(i in regions){
  region_backgrounds |>
    rename(region = regn_xp) |>
    select(region) |>
    filter(region == i) |>
    assign(x = glue::glue("region_background_{i}"),
         value = _)

  save(list = glue::glue("region_background_{i}"),
       file = glue::glue(here::here(),
                       "/data/region_background_{i}.rda"),
     compress = "xz")
}

```

Save all regions in a single object. These files are pretty heavy, so we will put them in a confidential folder:
```{r}
save(
  census_data_da,
  file = glue::glue(here::here(), "/data-raw/census/census_data_da.rda"),
  compress = "xz")
```

```{r}
save(
  census_data_cma, 
  file = glue::glue(here::here(), "/data-raw/census/census_data_cma.rda"),
  compress = "xz")
```

```{r}
save(
  region_backgrounds, 
  file = glue::glue(here::here(), "/data-raw/census/region_backgrounds.rda"),
  compress = "xz")
```

### 2024-11-04 data-raw/04_accessibility_measure_calcs.qmd
package files

---
title: "05_accessibility_measure_calcs"
format: html
---

# Setup

```{r}
#| label: load packages

#library(arrow) # Integration to 'Apache' 'Arrow'
library(duckdb) # DBI Package for the DuckDB Database Management System
library(dbplyr) # A 'dplyr' Back End for Databases
library(fs) # Cross-Platform File System Operations Based on 'libuv'
library(glue) # Interpreted String Literals
library(here) # A Simpler Way to Find Your Files
library(hms) # Pretty Time of Day
library(sf) # Simple Features for R
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
#library(accessibility) 
```

# Load Data

```{r}
# Load census data. These files are confidential and were created in notebook 00_census_data.qmd
load(file = glue::glue(here::here(), 
                       "/data-raw/census/census_data_da.rda"))
load(file = glue::glue(here::here(), 
                       "/data-raw/census/census_data_cma.rda"))
load(file = glue::glue(here::here(), 
                       "/data-raw/census/region_backgrounds.rda"))

census_data_da <- census_data_da |>  
  mutate(Population = ifelse(is.na(Population), 0, Population),
          Total_Emp = ifelse(is.na(Total_Emp), 0, Total_Emp))

# Load grocers data. This is a confidential file that was created in notebook 01_cleaning_destinations.qmd
load(glue::glue(here::here(),
                       "/data-raw/DMTI_2023_EPOI/grocers.rda"))
```

We will only use the 12pm departure transit travel time matrices (already run in 03.qmd) to grocery stores (DA-to-store) and employment locations (DA-to-DAs). This will be the basis of our accessibility calculations.

Census data by DA for join
```{r}
census_data_da_for_join <- census_data_da |> 
  st_drop_geometry() |> 
  select(GeoUID, study_region_name, Population) #|>  #including POPULATION here
  #arrow::to_duckdb()
```

# Travel time to grocery stores

First, grocery store ttms:

<!--

**WARNING**: The travel time matrices for 12 pm are missing.

```{r}
#| eval: false
ttm_small_path <- fs::dir_create("./ttms_12pm")

ttm_12pm_folders <- fs::dir_ls("./ttms", 
                               type = "directory", 
                               regexp = "12_00_00", 
                               recurse = TRUE)

walk(ttm_12pm_folders, ~dir_copy(.x, fs::path(ttm_small_path, (.x)), overwrite = TRUE))
```

read in travel time matrices as a data frame:
```{r}
# preparing the data frame to only consider only the 12:00pm departure (+15 min departure window), on Saturday!
travel_matrix_grocer <- arrow::open_dataset(glue::glue(here::here(), 
                                                       "/data-raw/ttms_12pm")) |>
  collect() |>
  mutate(
    start_datetime = lubridate::ymd_hms(start_datetime)
    ) |> filter(start_datetime  == as.POSIXct("2019-04-20 12:00:00", tz="UTC") |  
         start_datetime  == as.POSIXct("2019-06-29 12:00:00", tz="UTC") |  #for quebec
           start_datetime  == as.POSIXct("2023-04-22 12:00:00", tz="UTC")) |> #keeping saturdays!

  mutate(start_year = lubridate::year(start_datetime)) |> #adding the start_year

  left_join(census_data_da_for_join |> select(c(GeoUID, study_region_name, Population)), by = c("from_id" = "GeoUID")) #joining the population to the "from_id"

```

Assigning grocery store related travel impedance weights, for now, from Kwan 1998
```{r}
b0 <- 180
travel_matrix_grocer <- travel_matrix_grocer |>  
  mutate(
    o_j = 1, # opportunity intensity of each grocery store is 1
    f = exp(-travel_time_p50 ^ 2 / b0) #informed by Kwan 1998) 
    )
```

## grocer spatial accessibility

Let's also add non-competitive grocery store spatial accessibility to the frame:
```{r}
access_grocers_2019 <- travel_matrix_grocer_2019 |> group_by(from_id) |> summarise(S_i = sum(f*o_j))
access_grocers_2023 <- travel_matrix_grocer_2023 |> group_by(from_id) |> summarise(S_i = sum(f*o_j))

#also adding per capita spatial accessibility 
spatial_mea_grocers_2019_DA <- spatial_mea_grocers_2019_DA |> left_join(access_grocers_2019, by="from_id") |>
  mutate(s_i = S_i/Population)
spatial_mea_grocers_2023_DA <- spatial_mea_grocers_2023_DA |> left_join(access_grocers_2023, by="from_id") |>
  mutate(s_i = S_i/Population)
```
```{r}
#to save space, remove ttms grocers
rm(travel_matrix_grocer, travel_matrix_grocer_2019, travel_matrix_grocer_2023)
```
-->

# Accessibility to employment

Read DA-to-DA travel time matrices as a dataframe:
```{r}
# preparing the data frame to only consider only the 12:00pm departure (+15 min departure window), on Tuesday!
travel_matrix_employ <- arrow::open_dataset(glue::glue(here::here(), 
                                                       "/data-raw/ttms_employment")) |>
  collect() |>
  mutate(
    start_datetime = lubridate::ymd_hms(start_datetime)
    ) |>
  filter(start_datetime  == as.POSIXct("2019-04-16 08:00:00", tz="UTC") |  
           start_datetime  == as.POSIXct("2019-06-25 08:00:00", tz="UTC") |  #quebec!
           start_datetime  == as.POSIXct("2023-04-18 08:00:00", tz="UTC")) |> #keeping tuesdays!

  mutate(start_year = lubridate::year(start_datetime)) |> #adding the start_year

  left_join(census_data_da_for_join |> select(c(GeoUID, study_region_name, Population)), by = c("from_id" = "GeoUID")) |> #joining the population to the "from_id"

  left_join(census_data_da |> st_drop_geometry() |> select(c(GeoUID, Total_Emp)), by = c("to_id" = "GeoUID")) #joining the JOBS to the "to_id"

```

Save travel time matrices for employment:
```{r}
regions <- levels(as.factor(travel_matrix_employ$region))

for(i in regions){
  travel_matrix_employ |>
    filter(region == i) |>
    assign(x = glue::glue("travel_matrix_emp_{i}"),
         value = _)

  save(list = glue::glue("travel_matrix_emp_{i}"),
       file = glue::glue(here::here(),
                       "/data/travel_matrix_emp_{i}.rda"),
     compress = "xz")
}

```

<!--  
Assigning employment related travel impedance weights, for now, for transit mode in Kapitllian et al. 2023 is used:

```{r}
travel_matrix_employ <- travel_matrix_employ |>  mutate(
    f = case_when(
      study_region_name == "Calgary" ~  (1 / (1 + (travel_time_p50 / 41.6) ^ 5.0643)),
      study_region_name == "Edmonton" ~  (1 / (1 + (travel_time_p50 / 53) ^ 5.0643)),
      study_region_name == "Toronto" ~ (1 / (1 + (travel_time_p50 / 49) ^ 4.4856)),
      study_region_name == "Halifax" ~  (1 / (1 + (travel_time_p50 / 49) ^ 4.2019)),
      study_region_name == "London" ~ (1 / (1 + (travel_time_p50 / 47) ^ 4.6801)),
      study_region_name == "Montréal" ~ (1 / (1 + (travel_time_p50 / 44) ^ 4.1168)),
      study_region_name == "Ottawa" ~  (1 / (1 + (travel_time_p50 / 42.2) ^ 5.0643)),
      study_region_name == "Québec City" ~ (1 / (1 + (travel_time_p50 / 49) ^ 4.1947)),
      study_region_name == "Vancouver" ~ (1 / (1 + (travel_time_p50 / 45) ^ 4.1339)),
      study_region_name == "Winnipeg" ~ (1 / (1 + (travel_time_p50 / 45) ^ 4.6869)),
      study_region_name == "Hamilton" ~ (1 / (1 + (travel_time_p50 / 50) ^ 4.6869)),
      study_region_name == "Waterloo" ~ (1 / (1 + (travel_time_p50 / 40.3) ^ 4.6801)),
      .default = 9999
    ))
```

## job spatial availability

Calculating employment spatial_availability for 2019:
```{r}
#for 2019
travel_matrix_employ_2019 <- travel_matrix_employ |> filter(start_year == "2019")

spatial_mea_employ_2019 <- 
  travel_matrix_employ_2019 |> mutate(catch = 1) |>
  sp_avail_detailed(o_id = from_id,
                    d_id = to_id, 
                    pop = Population,
                    opp = Total_Emp,
                    r = catch, #value of 1, all are in catchment.
                    f = f, 
                    alpha = 1) |>
  group_by(from_id) |>
  summarize(V_i = sum(V_ij, na.rm = TRUE))

#Checks, to make sure the total number of opportunities equals the total sum of V_i:
travel_matrix_employ_2019 |> group_by(to_id) |> summarise(jobs = first(Total_Emp)) |> select(jobs) |> sum()
spatial_mea_employ_2019$V_i |> sum()
```

for 2023:
```{r}
travel_matrix_employ_2023 <- travel_matrix_employ |> filter(start_year == "2023")

spatial_mea_employ_2023 <- 
  travel_matrix_employ_2023 |> mutate(catch = 1) |> 
  sp_avail_detailed(o_id = from_id,
                    d_id = to_id, 
                    pop = Population,
                    opp = Total_Emp,
                    r = catch, #value of 1, all are in catchment.
                    f = f, 
                    alpha = 1) |>
  group_by(from_id) |>
  summarize(V_i = sum(V_ij, na.rm = TRUE))

#Checks, to make sure the total number of opportunities equals the total sum of V_i:
travel_matrix_employ_2023 |> group_by(to_id) |> summarise(jobs = first(Total_Emp)) |> select(jobs) |> sum()
spatial_mea_employ_2023$V_i |> sum()
spatial_mea_employ_2023|> filter(V_i == 0)

#NOTE: the total number of jobs (and population) differs between 2019 and 2023 -- this is because people/jobs with OD travel times are different from the 2019 vs. 2023 travel matrices. 
```

job spatial availability per capita:
```{r}
spatial_mea_employ_2019_DA <- spatial_mea_employ_2019 |> full_join(census_data_da, by=c("from_id" = "GeoUID")) |>
  mutate(Population = ifelse(is.na(Population), 0, Population),
         v_i = V_i/Population)

spatial_mea_employ_2023_DA <- spatial_mea_employ_2023 |> full_join(census_data_da, by=c("from_id" = "GeoUID"))|>
  mutate(Population = ifelse(is.na(Population), 0, Population),
         v_i = V_i/Population)
```

## job spatial accessibility

Let's add non-competitive job spatial accessibility to the frame:
```{r}
#calculating Hansen-type spatial accessibility, sum of (f*opportunities) for each origin (from_id) to all destination (possible to_ids):
access_employ_2019 <- travel_matrix_employ_2019 |> group_by(from_id) |> summarise(S_i = sum(f*Total_Emp))
access_employ_2023 <- travel_matrix_employ_2023 |> group_by(from_id) |> summarise(S_i = sum(f*Total_Emp))

#inspect:
access_employ_2019$S_i |> summary()
access_employ_2023$S_i |> summary()

#also adding per capita spatial accessibility 
spatial_mea_employ_2019_DA <- spatial_mea_employ_2019_DA |> left_join(access_employ_2019, by="from_id") |>
  mutate(s_i = S_i/Population)
spatial_mea_employ_2023_DA <- spatial_mea_employ_2023_DA |> left_join(access_employ_2023, by="from_id") |>
  mutate(s_i = S_i/Population)
```

```{r}
#to save space, remove ttms grocers
rm(travel_matrix_employ, travel_matrix_employ_2019, travel_matrix_employ_2023)
```

Saving all files
```{r, eval=FALSE}
#to save space.. remove the census data columns
spatial_mea_employ_2019_DA <- spatial_mea_employ_2019_DA |> select(c("from_id", "study_region_name", "V_i", "v_i", "S_i", "s_i"))
spatial_mea_employ_2023_DA <- spatial_mea_employ_2023_DA |> select(c("from_id", "study_region_name", "V_i", "v_i", "S_i", "s_i"))
spatial_mea_grocers_2019_DA <- spatial_mea_grocers_2019_DA |> select(c("from_id", "study_region_name", "V_i", "v_i", "S_i", "s_i"))
spatial_mea_grocers_2023_DA <- spatial_mea_grocers_2023_DA |> select(c("from_id", "study_region_name", "V_i", "v_i", "S_i", "s_i"))

save(spatial_mea_employ_2019_DA, 
     spatial_mea_employ_2023_DA , 
     spatial_mea_grocers_2019_DA, 
     spatial_mea_grocers_2023_DA , 
     file="./data/spatial_access_measures.Rda")
```

### 2024-11-04 R/canaccessR-package.R

@@ -1,0 +1,1512 @@
+#' @keywords internal
+"_PACKAGE"
+
+## usethis namespace: start
+## usethis namespace: end
+NULL
+
+#' Travel time matrix and employment statistics for Calgary Metropolitan Area.
+#'
+#' A data frame with XX origin-destination pairs, travel time by transit, and
+#' population and employment by Census Dissemination Area. Census Dissemination
+#' Areas (DAs) are the smallest publicly available geography provided by Statistics
+#' Canada. Travel times are for a representative day in two different years:
+#' 2019 (pre-covid measures) and 2023 (post-covid measures). Travel times are
+#' calculated as the median of all times within a 15 minutes window of the departure.
+#'
+#' @format A data frame with XXX rows and 9 variables:
+#' \describe{
+#'   \item{from_id}{Unique identifier of the Dissemination Area of the zone of origin}
+#'   \item{to_id}{Unique identifier of the Dissemination Area of the zone of destination}
+#'   \item{travel_time_p50}{Median travel time in minutes by transit within a 15 minute window of the time of departure of 8:00 am}
+#'   \item{region}{Identifier of the region: "cal" for Calgary}
+#'   \item{start_year}{Year for which travel time was calculated}
+#'   \item{start_datetime}{Date and time of departure for a trip that originates at the Dissemination Area of origin}
+#'   \item{study_region_name}{Name of the Metropolita Area: Calgary}
+#'   \item{Population}{Total population in the Dissemination Area of origin (2016 Census)}
+#'   \item{Total_Emp}{Total employment in the Dissemination Area of destination (20XX Census)}
+#' }
+#'
+#' @docType data
+#' @keywords datasets
+#' @name travel_matrix_emp_cal
+#' @usage data(travel_matrix_emp_cal)
+#' @source Population: Canada Census 2016. See: https://mountainmath.github.io/cancensus/index.html
+#' @source Employment: Statistics Canada
+#' @source Travel times: calculated based on General Transit Feed Standard and OpenStreetMaps data
+#' @examples
+#'  data(travel_matrix_emp_cal)
+#'  summary(travel_matrix_emp_cal$travel_time_p50)
+"travel_matrix_emp_cal"
+
+#' Travel time matrix and employment statistics for Edmonton Metropolitan Area.
+#'
+#' A data frame with XX origin-destination pairs, travel time by transit, and
+#' population and employment by Census Dissemination Area. Census Dissemination
[…]

### 2024-11-04 data-raw/04_accessibility_measure_calcs.qmd
Added grocery travel time tables

@@
 
-<!--
-
-**WARNING**: The travel time matrices for 12 pm are missing.
-
-```{r}
-#| eval: false
-ttm_small_path <- fs::dir_create("./ttms_12pm")
-
-ttm_12pm_folders <- fs::dir_ls("./ttms", 
-                               type = "directory", 
-                               regexp = "12_00_00", 
-                               recurse = TRUE)
-
-walk(ttm_12pm_folders, ~dir_copy(.x, fs::path(ttm_small_path, (.x)), overwrite = TRUE))
-```
-
 read in travel time matrices as a data frame:
@@
 
+Save matrices with travel time to groceries:
+```{r}
+regions <- levels(as.factor(travel_matrix_grocer$region))
+
+for(i in regions){
+  travel_matrix_grocer |>
+    filter(region == i) |>
+    assign(x = glue::glue("travel_matrix_grc_{i}"),
+         value = _)
+
+  save(list = glue::glue("travel_matrix_grc_{i}"),
+       file = glue::glue(here::here(),
+                       "/data/travel_matrix_grc_{i}.rda"),
+     compress = "xz")
+}
+
+```
+
+<!--
 Assigning grocery store related travel impedance weights, for now, from Kwan 1998
@@
 ```
-
-## grocer spatial accessibility
-
-Let's also add non-competitive grocery store spatial accessibility to the frame:
-```{r}
-access_grocers_2019 <- travel_matrix_grocer_2019 |> group_by(from_id) |> summarise(S_i = sum(f*o_j))
-access_grocers_2023 <- travel_matrix_grocer_2023 |> group_by(from_id) |> summarise(S_i = sum(f*o_j))
-
-#also adding per capita spatial accessibility 
-spatial_mea_grocers_2019_DA <- spatial_mea_grocers_2019_DA |> left_join(access_grocers_2019, by="from_id") |>
-  mutate(s_i = S_i/Population)
-spatial_mea_grocers_2023_DA <- spatial_mea_grocers_2023_DA |> left_join(access_grocers_2023, by="from_id") |>
-  mutate(s_i = S_i/Population)
-```
-```{r}
-#to save space, remove ttms grocers
-rm(travel_matrix_grocer, travel_matrix_grocer_2019, travel_matrix_grocer_2023)
-```
 -->
 
-# Accessibility to employment
+# Travel time to employment
 

### 2024-11-04 R/canaccessR-package.R

@@ -7,14 +7,15 @@ NULL
 
 #' Travel time matrix and employment statistics for Calgary Metropolitan Area.
 #'
-#' A data frame with XX origin-destination pairs, travel time by transit, and
+#' A data frame with origin-destination pairs, travel time by transit, and
 #' population and employment by Census Dissemination Area. Census Dissemination
 #' Areas (DAs) are the smallest publicly available geography provided by Statistics
 #' Canada. Travel times are for a representative day in two different years:
-#' 2019 (pre-covid measures) and 2023 (post-covid measures). Travel times are
-#' calculated as the median of all times within a 15 minutes window of the departure.
+#' 2019 (pre-covid measures) and 2023 (post-covid measures) from the centroid of 
+#' the origin DA to the centroid of the destination DA. Times are calculated as 
+#' the median of all times within a 15 minutes window of the departure.
 #'
-#' @format A data frame with XXX rows and 9 variables:
+#' @format A data frame with 3,022,244 rows and 9 variables:
 #' \describe{
 #'   \item{from_id}{Unique identifier of the Dissemination Area of the zone of origin}
 #'   \item{to_id}{Unique identifier of the Dissemination Area of the zone of destination}
@@ -24,7 +25,7 @@ NULL
 #'   \item{start_datetime}{Date and time of departure for a trip that originates at the Dissemination Area of origin}
 #'   \item{study_region_name}{Name of the Metropolita Area: Calgary}
 #'   \item{Population}{Total population in the Dissemination Area of origin (2016 Census)}
-#'   \item{Total_Emp}{Total employment in the Dissemination Area of destination (20XX Census)}
+#'   \item{Total_Emp}{Total employment in the Dissemination Area of destination (2016 Census)}
 #' }
 #'
 #' @docType data
@@ -41,14 +42,15 @@ NULL
 
 #' Travel time matrix and employment statistics for Edmonton Metropolitan Area.
 #'
-#' A data frame with XX origin-destination pairs, travel time by transit, and
+#' A data frame with origin-destination pairs, travel time by transit, and
 #' population and employment by Census Dissemination Area. Census Dissemination
 #' Areas (DAs) are the smallest publicly available geography provided by Statistics
 #' Canada. Travel times are for a representative day in two different years:
-#' 2019 (pre-covid measures) and 2023 (post-covid measures). Travel times are
-#' calculated as the median of all times within a 15 minutes window of the departure.
[…]

### 2024-11-04 R/canaccessR-package.R
Added transit statistics

@@ -1870,3 +1870,26 @@ NULL
 #' @examples
 #'  data(region_background_win)
 "region_background_win"
+
+#' National and regional transit revenue and ridership in Canada.
+#'
+#' A data frame with monthly ridership and revenue statistics for urban transit 
+#' agencies, from April, 2017 to August, 2024.
+#'
+#' @format A data frame with 368 rows and 4 variables:
+#' \describe{
+#'   \item{GEO}{Names of four geographies as a factor: Canada, Atlantic, Quebec and Ontario, Prairies, British Columbia and Territories}
+#'   \item{REF_DATE}{The date of the data}
+#'   \item{Revenue}{Revenue of transit agencies, before subsidies, in dollars}
+#'   \item{Ridership}{Total number of passengers}
+#' }
+#'
+#' @docType data
+#' @keywords datasets
+#' @name transit_statistics
+#' @usage data(transit_statistics)
+#' @source Statistics Canada. Table 23-10-0251-01  Passenger bus and urban transit statistics, by the North American Industry Classification System (NAICS) (x 1,000,000) (https://doi.org/10.25318/2310025101-eng)
+#' @examples
+#'  data(transit_statistics)
+#'  summary(transit_statistics)
+"transit_statistics"

### 2024-11-04 data-raw/05_transit_statistics.qmd
Added table with population statistics

---
title: "transit statistics"
format: html
---

Libraries:
```{r}
library(dplyr) # A Grammar of Data Manipulation
library(glue) # Interpreted String Literals
library(here) # A Simpler Way to Find Your Files
library(lubridate) # Make Dealing with Dates a Little Easier
library(readr) # Read Rectangular Text Data
library(tidyr) # Tidy Messy Data
#library(sf) # Simple Features for R
#library(tibble) # Simple Data Frames
#library(tmap) # Thematic Maps
```

- Downloaded "Urban Public Transit, February 2024" table: https://www150.statcan.gc.ca/n1/daily-quotidien/240419/dq240419b-eng.htm and https://doi.org/10.25318/2310025101-eng

#Reading in and filtering

Reading in files
```{r}
transit_statistics <- read_csv(glue::glue(here::here(),
                                          "/data-raw/2310025101_databaseLoadingData.csv"))
```

```{r}
summary(transit_statistics)
```

Select the relevant information:
```{r}
transit_statistics <- transit_statistics |>
  transmute(GEO = factor(GEO),
            REF_DATE = lubridate::ym(REF_DATE) |>
              as.Date(),
         # Values are in millions
         VALUE = VALUE * 1000000,
         VAR = case_when(`Total revenue and total passenger trips` == "Total revenue, excluding subsidies" ~ "Revenue",
                         `Total revenue and total passenger trips` == "Total passenger trips" ~ "Ridership"))
```

Pivot to a wide table to have a column for revenue and other for ridership:
```{r}
transit_statistics <- transit_statistics |>
  pivot_wider(names_from = "VAR", values_from = "VALUE")
```

Save:
```{r}
save(transit_statistics,
     file = glue::glue(here::here(),
                       "/data/transit_statistics.rda"),
     compress = "xz")
```

### 2024-11-04 data-raw/06_population_statistics.qmd
Added table with population statistics

---
title: "transit statistics"
format: html
---

Libraries:
```{r}
library(dplyr) # A Grammar of Data Manipulation
library(glue) # Interpreted String Literals
library(here) # A Simpler Way to Find Your Files
#library(lubridate) # Make Dealing with Dates a Little Easier
library(readr) # Read Rectangular Text Data
library(tidyr) # Tidy Messy Data
```

- Downloaded "Urban Public Transit, February 2024" table: https://www150.statcan.gc.ca/n1/daily-quotidien/240419/dq240419b-eng.htm and https://doi.org/10.25318/2310025101-eng

#Reading in and filtering

Reading in files
```{r}
population_statistics <- read_csv(glue::glue(here::here(),
                                          "/data-raw/population_statistics.csv"))
```

```{r}
summary(population_statistics)
```

Select the relevant information:
```{r}
population_statistics <- population_statistics |>
  transmute(size_rank = factor(`Size Rank`),
            year = 2021,
            cma_name = factor(`CMA Geographic Name`),
            province = factor(Province),
            population = `Population (2021)`)
```

Save:
```{r}
save(population_statistics,
     file = glue::glue(here::here(),
                       "/data/population_statistics.rda"),
     compress = "xz")
```

### 2024-11-04 R/canaccessR-package.R

@@ -1893,3 +1893,27 @@ NULL
 #'  data(transit_statistics)
 #'  summary(transit_statistics)
 "transit_statistics"
+
+#' Population statistics for selected metropolitan areas in Canada.
+#'
+#' A data frame with population for selected Census Metropolitan Areas in Canada
+#' for 2021.
+#'
+#' @format A data frame with 14 rows and 5 variables:
+#' \describe{
+#'   \item{size_rank}{Rank of the metropolitan area according to population}
+#'   \item{year}{The date of the data}
+#'   \item{cma_name}{Name of the Census Metropolitan Area}
+#'   \item{province}{Name of the province that is home to the Census Metropolitan Area}
+#'   \item{population}{Population of the Census Metropolitan Area}
+#' }
+#'
+#' @docType data
+#' @keywords datasets
+#' @name population_statistics
+#' @usage data(population_statistics)
+#' @source Statistics Canada. 
+#' @examples
+#'  data(population_statistics)
+#'  summary(population_statistics)
+"population_statistics"

### 2024-11-05 data-raw/00_census_data.qmd
Added a relevant variable that was missing from the census data tables, and updated the documentation

@@
 
-```{r}
-# save(
-#   census_data_da, 
-#   , 
-#   region_backgrounds, 
-#   file = "./data/census_data.Rdata")
-```
-
 Regions used for routing in {r5r}:
@@
     filter(region == i) |>
-    select(Type, GeoUID, name, region) |>
+    select(Type, GeoUID, name, region, study_region_name) |>
     assign(x = glue::glue("census_data_cma_{i}"),
@@
     filter(region == i) |>
-    select(-c(`Shape Area`, `Region Name`, `Area (sq km)`, popn_dens, `study_region_name`)) |>
+    select(-c(`Shape Area`, `Region Name`, `Area (sq km)`, popn_dens)) |>
     select(Type, GeoUID, region, everything()) |>

### 2024-11-05 R/canaccessR-package.R

@@ -23,7 +23,7 @@ NULL
 #'   \item{region}{Identifier of the region: "cal" for Calgary}
 #'   \item{start_year}{Year for which travel time was calculated}
 #'   \item{start_datetime}{Date and time of departure for a trip that originates at the Dissemination Area of origin}
-#'   \item{study_region_name}{Name of the Metropolita Area: Calgary}
+#'   \item{study_region_name}{Name of the Census Metropolitan Area (CMA)}
 #'   \item{Population}{Total population in the Dissemination Area of origin (2016 Census)}
 #'   \item{Total_Emp}{Total employment in the Dissemination Area of destination (2016 Census)}
 #' }
@@ -58,7 +58,7 @@ NULL
 #'   \item{region}{Identifier of the region: "edm" for Edmonton}
 #'   \item{start_year}{Year for which travel time was calculated}
 #'   \item{start_datetime}{Date and time of departure for a trip that originates at the Dissemination Area of origin}
-#'   \item{study_region_name}{Name of the Metropolita Area: Edmonton}
+#'   \item{study_region_name}{Name of the Census Metropolitan Area (CMA)}
 #'   \item{Population}{Total population in the Dissemination Area of origin (2016 Census)}
 #'   \item{Total_Emp}{Total employment in the Dissemination Area of destination (2016 Census)}
 #' }
@@ -93,7 +93,7 @@ NULL
 #'   \item{region}{Identifier of the region: "ggh" for Greater Golden Horseshoe}
 #'   \item{start_year}{Year for which travel time was calculated}
 #'   \item{start_datetime}{Date and time of departure for a trip that originates at the Dissemination Area of origin}
-#'   \item{study_region_name}{Name of the Metropolita Area: Greater Golden Horseshoe}
+#'   \item{study_region_name}{Name of the Census Metropolitan Area (CMA)}
 #'   \item{Population}{Total population in the Dissemination Area of origin (2016 Census)}
 #'   \item{Total_Emp}{Total employment in the Dissemination Area of destination (2016 Census)}
 #' }
@@ -128,7 +128,7 @@ NULL
 #'   \item{region}{Identifier of the region: "hal" for Halifax}
 #'   \item{start_year}{Year for which travel time was calculated}
 #'   \item{start_datetime}{Date and time of departure for a trip that originates at the Dissemination Area of origin}
-#'   \item{study_region_name}{Name of the Metropolita Area: Halifax}
[…]

### 2024-11-05 R/canaccessR-package.R
Finalized the documentation of the census data

@@ -1060,46 +1060,46 @@ NULL
 #'   \item{CSD_UID}{Unique identifier of the census CSD}
 #'   \item{CT_UID}{Unique identifier of the Census Tract}
 #'   \item{CMA_UID}{Unique identifier of the census Census Metropolitan Area}
-#'   \item{tot_age_popn}{Total age population}
-#'   \item{age_0_to_14}{Total population aged 0-14 years old in dissemination area in 2016}
-#'   \item{age_15_to_64}{Total population aged 15-64 years old in dissemination area in 2016}
-#'   \item{age_65_plus}{Total population aged 65 years and older in dissemination area in 2016}
-#'   \item{tot_cens_fam_priv_hh}{Total number of census families in private households in dissemination area in 2016}
-#'   \item{fem_parent}{define_fem_parent}
-#'   \item{LICO_AT}{define_lico_at}
-#'   \item{tot_LICO_status_2015}{define_tot_LICO}
-#'   \item{knwl_off_lang}{define_knwl_off_lang}
-#'   \item{neither_EN_FR}{define_neither_EN_FR}
-#'   \item{Aborin_ID_priv_HH}{define_Aborin_ID_priv_HH}
-#'   \item{Aborin_ID}{define_aborivin_ID}
-#'   \item{tot_priv_hh_hsing_suit}{define_tot_priv_hh_hsing_suit}
-#'   \item{not_suitable}{define_not_suitable}
-#'   \item{dwell_condtns}{define_dwell_condtns}
-#'   \item{maj_rep_needed}{define_maj_rep_needed}
-#'   \item{tot_shelt_cost}{define_tot_shelt_cost}
-#'   \item{shelt_gt30pc}{define_shelt_gt30pc}
-#'   \item{tot_tenant_HH}{define_tot_tenant_HH}
-#'   \item{tenant_HH_subsid}{define_tenant_HH_subsid}
-#'   \item{tot_immig_stat}{define_tot_immig_stat}
-#'   \item{immig_stat_last5}{define_immig_stat_last5}
-#'   \item{tot_adm_immig_popn_1980_to_2016}{define_tot_adm_immig_popn_1980_to_2016}
-#'   \item{refugees}{define_refugees}
-#'   \item{visible_min}{define_visible_min}
-#'   \item{tot_visible_min}{define_tot_visible_min}
-#'   \item{Black}{define_Black}
[…]

### 2025-01-22 paper/Manuscript-Data-Package.Rmd
Created a folder for the data paper and added it to .Rbuildignore

---
title: "The Title of the Paper"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "Abstract"

keywords: Accessibility; transit; open data products; travel time; employment; grocery stores 
classoption:
  - Royal
  - times
bibliography: bibfile.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  warning = FALSE,
  comment = '', 
  out.width = "1\\linewidth")

knitr::knit_hooks$set(inline = function(x) {
  if (is.numeric(x)) {
    format(x, big.mark = ",", scientific = FALSE)
  } else {
    x
  }
})

opts <- options(knitr.kable.NA = "")
```

```{r install-data-package, include=FALSE}
if (!require("canaccessR", character.only = TRUE)) {
      remotes::install_github("paezha/canaccessR",
                        build_vignettes = TRUE)
}

rm(list=ls())
```

```{r load-packages, include=FALSE, cache=FALSE}
# Load packages

```

```{r load-data, include = FALSE}

```

# Introduction

The objective of this paper is...

# Data Sources

Quickly describe the data sources before going into more detail.

## OpenStreetMaps

Files used for routing.

## Census Data

Employment etc.

# Data processing

Routing, cleaning, etc.

# How to use {canaccessR}

This section presents some potential applications of the data package.

## Descriptive statistics

Maybe?

<!--
# Python integration

Possibly provide a Jupyter Notebook containing a Python script to demonstrate how to read R data files (.rda) and convert them into Pandas DataFrames.

-->

# Concluding remarks

What is the value added to the raw data.

# Declaration of Conflicting Interests

The author(s) declared no potential conflicts of interest with respect
to the research, authorship, and/or publication of this article.

# Funding

The author(s) disclosed receipt of the following financial support for
the research, authorship, and/or publication of this article: This work
was supported by the Social Sciences and Humanities Research Council of
Canada (*More description about the funding source after the review
process*).

# ORCID

Author 1

Author 2

Author 3

# Data availability statement

The {canaccessR} data package can be found and installed on Github
(*link*).

# References {#references .unnumbered}

### 2025-01-22 paper/Manuscript-Data-Package.tex
Created a folder for the data paper and added it to .Rbuildignore

\documentclass[Royal,times,sageh]{sagej}

\usepackage{moreverb,url,natbib, multirow, tabularx}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

\setcitestyle{aysep={,}}

\title{The Title of the Paper}

\runninghead{}

\author{Anon1\affilnum{}, Anon2\affilnum{}, Anon3\affilnum{}}

\affiliation{\affilnum{}{}}

\begin{abstract}
Abstract
\end{abstract}

\keywords{Accessibility; transit; open data products; travel time;
employment; grocery stores}

\maketitle

\section{Introduction}\label{introduction}

The objective of this paper is\ldots{}

\section{Data Sources}\label{data-sources}

Quickly describe the data sources before going into more detail.

\subsection{OpenStreetMaps}\label{openstreetmaps}

Files used for routing.

\subsection{Census Data}\label{census-data}

Employment etc.

\section{Data processing}\label{data-processing}

Routing, cleaning, etc.

\section{How to use \{canaccessR\}}\label{how-to-use-canaccessr}

This section presents some potential applications of the data package.

\subsection{Descriptive statistics}\label{descriptive-statistics}

Maybe?

\section{Concluding remarks}\label{concluding-remarks}

What is the value added to the raw data.

\section{Declaration of Conflicting
Interests}\label{declaration-of-conflicting-interests}

The author(s) declared no potential conflicts of interest with respect
to the research, authorship, and/or publication of this article.

\section{Funding}\label{funding}

The author(s) disclosed receipt of the following financial support for
the research, authorship, and/or publication of this article: This work
was supported by the Social Sciences and Humanities Research Council of
Canada (\emph{More description about the funding source after the review
process}).

\section{ORCID}\label{orcid}

Author 1

Author 2

Author 3

\section{Data availability statement}\label{data-availability-statement}

The \{canaccessR\} data package can be found and installed on Github
(\emph{link}).

\bibliographystyle{sageh}
\bibliography{bibfile.bib}

\end{document}

### 2025-02-27 paper/Manuscript-Data-Package.Rmd
First draft of the paper
Initial draft of the data paper; introduction and conclusion done; still incomplete

@@
 ---
-title: "The Title of the Paper"
+title: "canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada’s largest metropolitan areas."
 author:
-  - name: "Anon1"
-    email: "Anon1@anon.org"
+  - name: "João Pedro Figueira Amorim Parga"
+    email: "joao.parga@mail.utoronto.ca"
     affiliation: CODE
     footnote: 1
-  - name: "Anon2"
-    email: "Anon2@anon.org"
+  - name: "Anastasia Soukhov"
+    email: "soukhoa@mcmaster.ca"
     affiliation: CODE
-  - name: "Anon3"
-    email: "Anon3@anon.org"
+  - name: "Robert Nutifafa Arku"
+    email: "rn.arku@mail.utoronto.ca"
+    affiliation: CODE
+  - name: "Christopher Higgins"
+    email: "cd.higgins@utoronto.ca"
+    affiliation: CODE
+  - name: "Antonio Páez"
+    email: "paezha@mcmaster.ca"
     affiliation: CODE        
@@
 abstract: 
-  "Abstract"
+  "In this paper, we describe the {canaccesR} package, an open data product (OPS) created using the R statistical language.
+{canaccess} is a data package that provides public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas.
+We calculate these estimates for each Dissemination Area (DA) within these regions for the years 2019 and 2023.
+To do so, we use the {r5r} R package, General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
+This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility to essential services across these regions.
+These estimates can be used to compare different regions across Canada in terms of their accessibility and to conduct within-region equity assessments regarding access to services, which can inform improvements in transportation policies related to accessibility.
+The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities).
+Finally, as an OPS, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."
 
-keywords: Accessibility; transit; open data products; travel time; employment; grocery stores 
+keywords: Accessibility; public transit; open data products; OPS; travel time; employment; grocery stores.
 classoption:
@@
 
-The objective of this paper is...
+The objective of this paper is to describe the {canaccesR} Open Data Product (OPS), an open data package.
+The package's main contents are public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in the years 2019 and 2023.
+These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility, using the {r5r} R package [PEREIRA CITATION].
+We used transport and street network, population, and employment data from different sources to arrive at our estimations.
+
+The main contribution of this data package is to provide analysis-ready data for Canada's largest cities, thus making the field of accessibility research and urban analytics more accessible.
+Estimating accessibility, i.e., the potential offered by the transportation system to reach destinations [Páez, Scott, and Morency, “Measuring Accessibility.”] requires specialized datasets and technical expertise.
+By integrating and processing raw data from diverse sources, estimating TTM's for two destinations types (e.g., jobs and groceries) across Canada's largest cities of the country, and making these findings publicly available, we hope this OPS can advance the field of urban analytics in Canada.
+
+We anticipate that these datasets will enable researchers to analyze changes in public transit accessibility to essential destinations, such as employment centers and grocery stores, in the largest urban areas of the country.
+Consequently, the package addresses the need for ready-to-use public transit accessibility data, helping researchers, transit agencies, and the general public in evaluating equity in transportation accessibility within and across Canada.
+
+Besides this introduction, we organize this paper as follows. 
+The next section contains a description of the data sources we used to construct the data package.
+Then, we recount the data processing necessary to create the package.
+Next, we go through the main contents of the data package, i.e., the travel time matrices estimated through our analysis.
+We present some basic descriptive statistics of these datasets, and elucidate how one can use them in accessibility analysis. 
+Finally, we conclude by explaining how we expect {canaccesR} to contribute to the urban analytics and science community.
 
@@
 
-Quickly describe the data sources before going into more detail.
+We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
+
+## General Transit Feed Specification
 
@@
 
+## DMTI’s Enhanced Points of Interest
+
 # Data processing
@@
 
-# How to use {canaccessR}
+# {canaccessR}'s contents
 
-This section presents some potential applications of the data package.
+The main contents of the {canaccessR} package are the travel time matrices estimates for all the 12 largest Canadian cities.
 
@@
 
-Maybe?
+Below, we present some of the basic statistics of the
 
@@
 
+# How to use {canaccessR}
+
+This section presents some potential applications of the data package.
+
 # Concluding remarks
 
-What is the value added to the raw data.
+In this paper, we describe the {canaccesR} data package, created using the {r5r} R package and transit schedule, street network, employment, and population data.
+The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.
+We expect the contents of the package to be used in transportation accessibility evaluations within and across those regions.
+Moreover, these datasets can be used in further equity assessments that evaluate the distribution of accessibility across space and between social groups.
+In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [MARTENS JUSTICE], and the inherent connection between public transit and the "right to the city" [COGGIN].
+
+<!-- In conclusion, we argue that the {canaccesR} package adds  -->
+
+<!-- In this sense, the package fulfill one of the main tenets of OPS products by combining expertise in data science, transportation accessibility, and demographics and generating a set of useful and ready-to-use data vital to estimating transportation accessibility, rather than simply making raw data available.
+-->
+<!-- In other words, the travel time matrics presented by this  -->
+
+<!-- value to the urban analytics and science community by  -->
+
+<!-- We expect the package to have two main contribu -->
+<!-- The main contribution from the {canaccesR} package is the  -->
+
+<!-- Finally, we expect that this data package can substantiate accessibility analysis across Canada. -->
 
@@
 
-The {canaccessR} data package can be found and installed on Github
-(*link*).
+The {canaccessR} data package can be found and installed on its Github [respository](https://github.com/paezha/canaccessR).
 

### 2025-02-28 paper/Manuscript-Data-Package.Rmd
correct odp

@@
 abstract: 
-  "In this paper, we describe the {canaccesR} package, an open data product (OPS) created using the R statistical language.
+  "In this paper, we describe the {canaccesR} package, an open data product (ODPs) created using the R statistical language.
 {canaccess} is a data package that provides public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas.
@@
 The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities).
-Finally, as an OPS, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."
+Finally, as an ODPs, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."
 
-keywords: Accessibility; public transit; open data products; OPS; travel time; employment; grocery stores.
+keywords: Accessibility; public transit; open data products (ODPs); travel time; employment; grocery stores.
 classoption:
@@
 
-The objective of this paper is to describe the {canaccesR} Open Data Product (OPS), an open data package.
+The objective of this paper is to describe the {canaccesR} Open Data Product (ODP), an open data package.
 The package's main contents are public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in the years 2019 and 2023.
@@
 Estimating accessibility, i.e., the potential offered by the transportation system to reach destinations [Páez, Scott, and Morency, “Measuring Accessibility.”] requires specialized datasets and technical expertise.
-By integrating and processing raw data from diverse sources, estimating TTM's for two destinations types (e.g., jobs and groceries) across Canada's largest cities of the country, and making these findings publicly available, we hope this OPS can advance the field of urban analytics in Canada.
+By integrating and processing raw data from diverse sources, estimating TTM's for two destinations types (e.g., jobs and groceries) across Canada's largest cities of the country, and making these findings publicly available, we hope this ODP can advance the field of urban analytics in Canada.
 
@@
 
-<!-- In this sense, the package fulfill one of the main tenets of OPS products by combining expertise in data science, transportation accessibility, and demographics and generating a set of useful and ready-to-use data vital to estimating transportation accessibility, rather than simply making raw data available.
+<!-- In this sense, the package fulfill one of the main tenets of ODPs products by combining expertise in data science, transportation accessibility, and demographics and generating a set of useful and ready-to-use data vital to estimating transportation accessibility, rather than simply making raw data available.
 -->

### 2025-02-28 paper/Manuscript-Data-Package.tex
correct odp

@@
 
-\title{The Title of the Paper}
+\title{canaccessR: An open data product for analyzing transportation
+accessibility to employment and grocery stores in Canada's largest
+metropolitan areas.}
 
@@
 
-\author{Anon1\affilnum{}, Anon2\affilnum{}, Anon3\affilnum{}}
+\author{João Pedro Figueira Amorim Parga\affilnum{}, Anastasia
+Soukhov\affilnum{}, Robert Nutifafa Arku\affilnum{}, Christopher
+Higgins\affilnum{}, Antonio Páez\affilnum{}}
 
@@
 \begin{abstract}
-Abstract
+In this paper, we describe the \{canaccesR\} package, an open data
+product (OPS) created using the R statistical language. \{canaccess\} is
+a data package that provides public transit travel time estimates
+(travel time matrices - TTM) to employment locations and grocery stores
+across the 12 largest Canadian metropolitan areas. We calculate these
+estimates for each Dissemination Area (DA) within these regions for the
+years 2019 and 2023. To do so, we use the \{r5r\} R package, General
+Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
+Points of Interest, and Statistics Canada Census data. This data package
+can be used by researchers, practitioners, and transit agencies to
+estimate accessibility to essential services across these regions. These
+estimates can be used to compare different regions across Canada in
+terms of their accessibility and to conduct within-region equity
+assessments regarding access to services, which can inform improvements
+in transportation policies related to accessibility. The package is
+still in its initial phase and may undergo expansions in the future by
+adding TTM's for other destinations (e.g., schools, healthcare
+facilities). Finally, as an OPS, the \{canaccess\} package allows for
+open exploration, use, and contribution by users through its GitHub
+repository.
 \end{abstract}
 
-\keywords{Accessibility; transit; open data products; travel time;
-employment; grocery stores}
+\keywords{Accessibility; public transit; open data products; OPS; travel
+time; employment; grocery stores.}
 
@@
 
-The objective of this paper is\ldots{}
+The objective of this paper is to describe the \{canaccesR\} Open Data
+Product (OPS), an open data package. The package's main contents are
+public transit travel time matrices (TTM) estimates to employment and
+groceries stores from the 12 largest Canadian metropolitan areas in the
+years 2019 and 2023. These estimates were created by leveraging
+expertise in data science, computer programming, and transportation
+accessibility, using the \{r5r\} R package {[}PEREIRA CITATION{]}. We
+used transport and street network, population, and employment data from
+different sources to arrive at our estimations.
+
+The main contribution of this data package is to provide analysis-ready
+data for Canada's largest cities, thus making the field of accessibility
+research and urban analytics more accessible. Estimating accessibility,
+i.e., the potential offered by the transportation system to reach
+destinations {[}Páez, Scott, and Morency, ``Measuring
+Accessibility.''{]} requires specialized datasets and technical
+expertise. By integrating and processing raw data from diverse sources,
+estimating TTM's for two destinations types (e.g., jobs and groceries)
+across Canada's largest cities of the country, and making these findings
+publicly available, we hope this OPS can advance the field of urban
+analytics in Canada.
+
+We anticipate that these datasets will enable researchers to analyze
+changes in public transit accessibility to essential destinations, such
+as employment centers and grocery stores, in the largest urban areas of
+the country. Consequently, the package addresses the need for
+ready-to-use public transit accessibility data, helping researchers,
+transit agencies, and the general public in evaluating equity in
+transportation accessibility within and across Canada.
+
+Besides this introduction, we organize this paper as follows. The next
+section contains a description of the data sources we used to construct
+the data package. Then, we recount the data processing necessary to
+create the package. Next, we go through the main contents of the data
+package, i.e., the travel time matrices estimated through our analysis.
+We present some basic descriptive statistics of these datasets, and
+elucidate how one can use them in accessibility analysis. Finally, we
+conclude by explaining how we expect \{canaccesR\} to contribute to the
+urban analytics and science community.
 
@@
 
-Quickly describe the data sources before going into more detail.
+We used four main data sources to construct the \{canaccesR\} data
+package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM),
+DMTI's Enhanced Points of Interest, and Statistics Canada Census data.
+
+\subsection{General Transit Feed
+Specification}\label{general-transit-feed-specification}
 
@@
 
+\subsection{DMTI's Enhanced Points of
+Interest}\label{dmtis-enhanced-points-of-interest}
+
 \section{Data processing}\label{data-processing}
@@
 Routing, cleaning, etc.
+
+\section{\{canaccessR\}'s contents}\label{canaccessrs-contents}
+
+The main contents of the \{canaccessR\} package are the travel time
+matrices estimates for all the 12 largest Canadian cities.
+
+\subsection{Descriptive statistics}\label{descriptive-statistics}
+
+Below, we present some of the basic statistics of the
 
@@
 
-\subsection{Descriptive statistics}\label{descriptive-statistics}
-
-Maybe?
-
 \section{Concluding remarks}\label{concluding-remarks}
 
-What is the value added to the raw data.
+In this paper, we describe the \{canaccesR\} data package, created using
+the \{r5r\} R package and transit schedule, street network, employment,
+and population data. The package's main contents refers to the
+ready-to-use travel time matrices for public transit to reach employment
+and groceries stores in Canada's 12 largest urban areas. We expect the
+contents of the package to be used in transportation accessibility
+evaluations within and across those regions. Moreover, these datasets
+can be used in further equity assessments that evaluate the distribution
+of accessibility across space and between social groups. In other words,
+we hope that by making these datasets publicly available, future
+analysis can contribute to making Canada's transportation system more
+just and fair, considering accessibility's as the main social good of
+transportation {[}MARTENS JUSTICE{]}, and the inherent connection
+between public transit and the ``right to the city'' {[}COGGIN{]}.
 
@@
 
-The \{canaccessR\} data package can be found and installed on Github
-(\emph{link}).
+The \{canaccessR\} data package can be found and installed on its Github
+\href{https://github.com/paezha/canaccessR}{respository}.
 

### 2025-03-06 paper/Manuscript-Data-Package.Rmd
Test commit
writing data paper
Complete introduction; continue data sources description; include orcid and authors' information.

@@
 author:
-  - name: "João Pedro Figueira Amorim Parga"
-    email: "joao.parga@mail.utoronto.ca"
-    affiliation: CODE
+  - name: João Pedro Figueira Amorim Parga
+    email: joao.parga@mail.utoronto.ca
+    affil-id: 1
+    orcid: 0000-0002-4105-5927
     footnote: 1
-  - name: "Anastasia Soukhov"
-    email: "soukhoa@mcmaster.ca"
-    affiliation: CODE
-  - name: "Robert Nutifafa Arku"
-    email: "rn.arku@mail.utoronto.ca"
-    affiliation: CODE
-  - name: "Christopher Higgins"
-    email: "cd.higgins@utoronto.ca"
-    affiliation: CODE
-  - name: "Antonio Páez"
-    email: "paezha@mcmaster.ca"
-    affiliation: CODE        
-address:
-  - code: CODE
-    address: "Some school"
+  - name: Anastasia Soukhov
+    email: soukhoa@mcmaster.ca
+    affil-id: 2
+    orcid: 0000-0003-4371-4831
+  - name: Robert Nutifafa Arku
+    email: rn.arku@mail.utoronto.ca
+    affil-id: 1
+    orcid: 0000-0002-2018-886X
+  - name: Christopher Higgins
+    email: cd.higgins@utoronto.ca
+    affil-id: 3
+    orcid: 0000-0002-3551-7750
+  - name: Antonio Páez
+    email: paezha@mcmaster.ca
+    affil-id: 2
+    orcid: 0000-0001-6912-9919
+affiliations:
+  - id: 1
+    name: University of Toronto
+  - id: 2
+    name: McMaster University
+  - id: 3
+    name: University of Toronto Scarborough
+# address:
+#   - code: CODE
+#     address: "Some school"
 footnote:
@@
 
-The objective of this paper is to describe the {canaccesR} Open Data Product (ODP), an open data package.
-The package's main contents are public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in the years 2019 and 2023.
-These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility, using the {r5r} R package [PEREIRA CITATION].
-We used transport and street network, population, and employment data from different sources to arrive at our estimations.
+The objective of this paper is to describe the {canaccesR} open data package.
+Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in the years 2019 and 2023.
+These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility, using the {r5r} R package [@pereiraR5rRapidRealistic2021].
+We used transport and street network, population, and employment data from different sources to estimate these TTM's datasets.
 
-The main contribution of this data package is to provide analysis-ready data for Canada's largest cities, thus making the field of accessibility research and urban analytics more accessible.
-Estimating accessibility, i.e., the potential offered by the transportation system to reach destinations [Páez, Scott, and Morency, “Measuring Accessibility.”] requires specialized datasets and technical expertise.
-By integrating and processing raw data from diverse sources, estimating TTM's for two destinations types (e.g., jobs and groceries) across Canada's largest cities of the country, and making these findings publicly available, we hope this ODP can advance the field of urban analytics in Canada.
+Estimating accessibility - the potential offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012] - requires specialized datasets and technical expertise.
+Despite recent advancements in the field, ready-to-use available data on transportation accessibility is still sparse.
+Thus, we leverage technical knowledge to make publicly available ready-to-use data on transportation accessibility.
+To create the package, we integrated and processed raw data from diverse sources, estimated TTM's for two destinations types (e.g., jobs and groceries) across the largest cities in the country, and distributed these findings through this transparent and open source data product.
+Our main contribution is to provide analysis-ready data for Canada's largest cities, thus making the fields of accessibility research and urban analytics more accessible.
 
-We anticipate that these datasets will enable researchers to analyze changes in public transit accessibility to essential destinations, such as employment centers and grocery stores, in the largest urban areas of the country.
-Consequently, the package addresses the need for ready-to-use public transit accessibility data, helping researchers, transit agencies, and the general public in evaluating equity in transportation accessibility within and across Canada.
+The package's main audiences are Canadian researchers in urban planning and transportation and transportation system agencies.
+We anticipate three main uses for the open data product (ODP) described in this paper.
+First, the datasets allow for static assessment of the level of public transit accessibility across the country's largest cities before and after the COVID-19 pandemic.
+In other words, we make it easier for those interested to compare cities and their level of public transit accessibility to essential destinations (such as employment centers and groceries stores).
+Second, researchers can use the package's information as inputs to evaluate changes in accessibility throughout this period. 
+Researchers may also investigate disparities in these changes across space within those cities, given the spacial character of the datasets made available here.
+Third, as is common practice in transportation accessibility research, these evaluations can substantiate broader investigations on matters of justice and equity in transportation.
+For example, the TTM estimates allow for evaluating the evolution of public transit's accessibility by income or spatial distribution across all Dissemination Areas (DA's) of each of the 12 cities in the sample [@pargaDemocraticAccessOur2024].
 
@@
 
+The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1], which are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
+
+[^1]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter.
+We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.
+
 We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
 
-## General Transit Feed Specification
+<!-- ## General Transit Feed Specification -->
+The GTFS files from the transit agencies from all the related areas provided the information on public transit schedule and their changes from 2019 to 2023.
 
-## OpenStreetMaps
+<!-- ## OpenStreetMaps -->
+The OpenStreetMap data provided information on the transit network and was necessary for the routing used to estimate the travel time matrices.
 
-Files used for routing.
+<!-- ## Census Data -->
+The 2016 and 2021 Statistics' Canada Census data provided the information on the spatial distribution of the population and the number of workplace locations, which represent the quantity of employment across the CMA's [@governmentofcanada2016CensusPopulation2016, governmentofcanada2021CensusPopulation2021].
+All of the resulting travel time matrices were aggregated at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
 
-## Census Data
-
-Employment etc.
-
-## DMTI’s Enhanced Points of Interest
+<!-- ## DMTI’s Enhanced Points of Interest -->
+The DMTI’s Enhanced Points of Interest
 
@@
 
+Other sets of data are also available at the {canaccessR} package besides the travel time matrices.
+These are the boundaries, socio-economic and demographic data (e.g., population, number of dwellings, number of individuals below the Low Income Measure, etc.) of the selected CMA's, disaggregated by DA.
+In addition, the package also contains aggregated population statistics (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale.
+
 ## Descriptive statistics
 
-Below, we present some of the basic statistics of the
+Below, we present some of the basic statistics of
 
@@
 
-Author 1
-
-Author 2
-
-Author 3
+  - name: João Pedro Figueira Amorim Parga
+    orcid: 0000-0002-4105-5927
+  - name: Anastasia Soukhov
+    orcid: 0000-0003-4371-4831
+  - name: Robert Nutifafa Arku
+    orcid: 0000-0002-2018-886X
+  - name: Christopher Higgins
+    orcid: 0000-0002-3551-7750
+  - name: Antonio Páez
+    orcid: 0000-0001-6912-9919
 

### 2025-03-06 paper/Manuscript-Data-Package.tex
writing data paper
Complete introduction; continue data sources description; include orcid and authors' information.

@@
 
-\affiliation{\affilnum{}{}}
+\affiliation{}
 
@@
 In this paper, we describe the \{canaccesR\} package, an open data
-product (OPS) created using the R statistical language. \{canaccess\} is
-a data package that provides public transit travel time estimates
+product (ODPs) created using the R statistical language. \{canaccess\}
+is a data package that provides public transit travel time estimates
 (travel time matrices - TTM) to employment locations and grocery stores
@@
 adding TTM's for other destinations (e.g., schools, healthcare
-facilities). Finally, as an OPS, the \{canaccess\} package allows for
+facilities). Finally, as an ODPs, the \{canaccess\} package allows for
 open exploration, use, and contribution by users through its GitHub
@@
 
-\keywords{Accessibility; public transit; open data products; OPS; travel
-time; employment; grocery stores.}
+\keywords{Accessibility; public transit; open data products (ODPs);
+travel time; employment; grocery stores.}
 
@@
 
-The objective of this paper is to describe the \{canaccesR\} Open Data
-Product (OPS), an open data package. The package's main contents are
-public transit travel time matrices (TTM) estimates to employment and
-groceries stores from the 12 largest Canadian metropolitan areas in the
-years 2019 and 2023. These estimates were created by leveraging
-expertise in data science, computer programming, and transportation
-accessibility, using the \{r5r\} R package {[}PEREIRA CITATION{]}. We
-used transport and street network, population, and employment data from
-different sources to arrive at our estimations.
+The objective of this paper is to describe the \{canaccesR\} open data
+package. Its main contents are a set of public transit travel time
+matrices (TTM) estimates to employment and groceries stores from the 12
+largest Canadian metropolitan areas in the years 2019 and 2023. These
+estimates were created by leveraging expertise in data science, computer
+programming, and transportation accessibility, using the \{r5r\} R
+package \citep{pereiraR5rRapidRealistic2021}. We used transport and
+street network, population, and employment data from different sources
+to estimate these TTM's datasets.
 
-The main contribution of this data package is to provide analysis-ready
-data for Canada's largest cities, thus making the field of accessibility
-research and urban analytics more accessible. Estimating accessibility,
-i.e., the potential offered by the transportation system to reach
-destinations {[}Páez, Scott, and Morency, ``Measuring
-Accessibility.''{]} requires specialized datasets and technical
-expertise. By integrating and processing raw data from diverse sources,
-estimating TTM's for two destinations types (e.g., jobs and groceries)
-across Canada's largest cities of the country, and making these findings
-publicly available, we hope this OPS can advance the field of urban
-analytics in Canada.
+Estimating accessibility - the potential offered by the transportation
+system to reach destinations
+\citep{paezMeasuringAccessibilityPositive2012} - requires specialized
+datasets and technical expertise. Despite recent advancements in the
+field, ready-to-use available data on transportation accessibility is
+still sparse. Thus, we leverage technical knowledge to make publicly
+available ready-to-use data on transportation accessibility. To create
+the package, we integrated and processed raw data from diverse sources,
+estimated TTM's for two destinations types (e.g., jobs and groceries)
+across the largest cities in the country, and distributed these findings
+through this transparent and open source data product. Our main
+contribution is to provide analysis-ready data for Canada's largest
+cities, thus making the fields of accessibility research and urban
+analytics more accessible.
 
-We anticipate that these datasets will enable researchers to analyze
-changes in public transit accessibility to essential destinations, such
-as employment centers and grocery stores, in the largest urban areas of
-the country. Consequently, the package addresses the need for
-ready-to-use public transit accessibility data, helping researchers,
-transit agencies, and the general public in evaluating equity in
-transportation accessibility within and across Canada.
+The package's main audiences are Canadian researchers in urban planning
+and transportation and transportation system agencies. We anticipate
+three main uses for the open data product (ODP) described in this paper.
+First, the datasets allow for static assessment of the level of public
+transit accessibility across the country's largest cities before and
+after the COVID-19 pandemic. In other words, we make it easier for those
+interested to compare cities and their level of public transit
+accessibility to essential destinations (such as employment centers and
+groceries stores). Second, researchers can use the package's information
+as inputs to evaluate changes in accessibility throughout this period.
+Researchers may also investigate disparities in these changes across
+space within those cities, given the spacial character of the datasets
+made available here. Third, as is common practice in transportation
+accessibility research, these evaluations can substantiate broader
+investigations on matters of justice and equity in transportation. For
+example, the TTM estimates allow for evaluating the evolution of public
+transit's accessibility by income or spatial distribution across all
+Dissemination Areas (DA's) of each of the 12 cities in the sample
+\citep{pargaDemocraticAccessOur2024}.
 
@@
 
+The locations included in the data package comprise the 12 largest
+(population-wise) Census metropolitan areas (CMA's) based on the 2021
+Canadian Census \citep{governmentofcanada2021CensusPopulation2021}
+\footnote{We included Oshawa, Ontario, as part of the Greater Toronto
+  Area (GTA) because of the former's proximity to the latter. We also
+  included Abbotsford-Mission, British Columbia, as part of the
+  Vancouver metropolitan area because of the former's proximity to a
+  transit station on the region's West Coast Express commuter rail line.},
+which are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary,
+Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo,
+London, and Halifax.
+
 We used four main data sources to construct the \{canaccesR\} data
@@
 
-\subsection{General Transit Feed
-Specification}\label{general-transit-feed-specification}
+The GTFS files from the transit agencies from all the related areas
+provided the information on public transit schedule and their changes
+from 2019 to 2023.
 
-\subsection{OpenStreetMaps}\label{openstreetmaps}
+The OpenStreetMap data provided information on the transit network and
+was necessary for the routing used to estimate the travel time matrices.
 
-Files used for routing.
+The 2016 and 2021 Statistics' Canada Census data provided the
+information on the spatial distribution of the population and the number
+of workplace locations, which represent the quantity of employment
+across the CMA's
+\citep[governmentofcanada2021CensusPopulation2021]{governmentofcanada2016CensusPopulation2016}.
+All of the resulting travel time matrices were aggregated at the
+Dissemination Area (DA) level, which comprise the fundamental unit of
+analysis in data package.
 
-\subsection{Census Data}\label{census-data}
-
-Employment etc.
-
-\subsection{DMTI's Enhanced Points of
-Interest}\label{dmtis-enhanced-points-of-interest}
+The DMTI's Enhanced Points of Interest
 
@@
 
+Other sets of data are also available at the \{canaccessR\} package
+besides the travel time matrices. These are the boundaries,
+socio-economic and demographic data (e.g., population, number of
+dwellings, number of individuals below the Low Income Measure, etc.) of
+the selected CMA's, disaggregated by DA. In addition, the package also
+contains aggregated population statistics (for the selected CMA's) and
+transit revenue and ridership data aggregated by regional and national
+scale.
+
 \subsection{Descriptive statistics}\label{descriptive-statistics}
 
-Below, we present some of the basic statistics of the
+Below, we present some of the basic statistics of
 
@@
 
-Author 1
-
-Author 2
-
-Author 3
+\begin{itemize}
+\tightlist
+\item
+  name: João Pedro Figueira Amorim Parga orcid: 0000-0002-4105-5927
+\item
+  name: Anastasia Soukhov orcid: 0000-0003-4371-4831
+\item
+  name: Robert Nutifafa Arku orcid: 0000-0002-2018-886X
+\item
+  name: Christopher Higgins orcid: 0000-0002-3551-7750
+\item
+  name: Antonio Páez orcid: 0000-0001-6912-9919
+\end{itemize}
 

### 2025-03-07 paper/Manuscript-Data-Package.Rmd
minor updates
minor updates to draft; descriptive statistics and how to use still undone.

@@
 
-The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1], which are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
+The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1].
+These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
 
@@
 We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.
-
 We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
@@
 <!-- ## General Transit Feed Specification -->
-The GTFS files from the transit agencies from all the related areas provided the information on public transit schedule and their changes from 2019 to 2023.
-
+We used the GTFS files from all transit agencies within the selected CMA's to obtain the information on their public transit schedule from 2019 to 2023.
 <!-- ## OpenStreetMaps -->
 The OpenStreetMap data provided information on the transit network and was necessary for the routing used to estimate the travel time matrices.
-
 <!-- ## Census Data -->
-The 2016 and 2021 Statistics' Canada Census data provided the information on the spatial distribution of the population and the number of workplace locations, which represent the quantity of employment across the CMA's [@governmentofcanada2016CensusPopulation2016, governmentofcanada2021CensusPopulation2021].
-All of the resulting travel time matrices were aggregated at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
-
+We used the 2016 and 2021 Canadian Census' information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016, governmentofcanada2021CensusPopulation2021].
+We aggregayed all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
 <!-- ## DMTI’s Enhanced Points of Interest -->
-The DMTI’s Enhanced Points of Interest
+Finally, we use the DMTI’s Enhanced Points of Interest dataset to obtain the location of the groceries stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015]. 
+We filtered the locations within the DMTI dataset using the grocery stores code from the North American Industry Classification System (NAICS) and the Standard Industrial Classification (SIC).
 
@@
 
-Below, we present some of the basic statistics of
+Below, we present some of the basic statistics from the travel time matrices contained in the {canaccessR} package.
 
@@
 Moreover, these datasets can be used in further equity assessments that evaluate the distribution of accessibility across space and between social groups.
-In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [MARTENS JUSTICE], and the inherent connection between public transit and the "right to the city" [COGGIN].
+In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [@martensTransportJusticeDesigning2016], and the inherent connection between public transit and the "right to the city" [@cogginRightTransportMoving2015].
 

### 2025-03-07 paper/Manuscript-Data-Package.tex
minor updates
minor updates to draft; descriptive statistics and how to use still undone.

@@
   Vancouver metropolitan area because of the former's proximity to a
-  transit station on the region's West Coast Express commuter rail line.},
-which are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary,
-Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo,
-London, and Halifax.
+  transit station on the region's West Coast Express commuter rail line.
+  We used four main data sources to construct the \{canaccesR\} data
+  package: General Transit Feed Specification (GTFS), OpenStreetMap
+  (OSM), DMTI's Enhanced Points of Interest, and Statistics Canada
+  Census data.}. These locations are Toronto, Montreal, Vancouver,
+Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
+Kitchener-Cambridge-Waterloo, London, and Halifax.
 
-We used four main data sources to construct the \{canaccesR\} data
-package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM),
-DMTI's Enhanced Points of Interest, and Statistics Canada Census data.
-
-The GTFS files from the transit agencies from all the related areas
-provided the information on public transit schedule and their changes
-from 2019 to 2023.
-
-The OpenStreetMap data provided information on the transit network and
-was necessary for the routing used to estimate the travel time matrices.
-
-The 2016 and 2021 Statistics' Canada Census data provided the
-information on the spatial distribution of the population and the number
-of workplace locations, which represent the quantity of employment
-across the CMA's
+We used the GTFS files from all transit agencies within the selected
+CMA's to obtain the information on their public transit schedule from
+2019 to 2023. The OpenStreetMap data provided information on the transit
+network and was necessary for the routing used to estimate the travel
+time matrices. We used the 2016 and 2021 Canadian Census' information on
+the spatial distribution of the population and the number of workplace
+locations (employment) across the CMA's
 \citep[governmentofcanada2021CensusPopulation2021]{governmentofcanada2016CensusPopulation2016}.
-All of the resulting travel time matrices were aggregated at the
+We aggregayed all the resulting travel time matrices at the
 Dissemination Area (DA) level, which comprise the fundamental unit of
-analysis in data package.
-
-The DMTI's Enhanced Points of Interest
+analysis in data package. Finally, we use the DMTI's Enhanced Points of
+Interest dataset to obtain the location of the groceries stores within
+every urban area selected
+\citep{dmtispatialincEnhancedPointsInterest2015}. We filtered the
+locations within the DMTI dataset using the grocery stores code from the
+North American Industry Classification System (NAICS) and the Standard
+Industrial Classification (SIC).
 
@@
 
-Below, we present some of the basic statistics of
+Below, we present some of the basic statistics from the travel time
+matrices contained in the \{canaccessR\} package.
 
@@
 just and fair, considering accessibility's as the main social good of
-transportation {[}MARTENS JUSTICE{]}, and the inherent connection
-between public transit and the ``right to the city'' {[}COGGIN{]}.
+transportation \citep{martensTransportJusticeDesigning2016}, and the
+inherent connection between public transit and the ``right to the city''
+\citep{cogginRightTransportMoving2015}.
 

### 2025-03-07 paper/Manuscript-Data-Package.Rmd
Data and methods
Finished data sources and methods; added other ODP references; descriptive statistics and how to use still incomplete

@@
   "In this paper, we describe the {canaccesR} package, an open data product (ODPs) created using the R statistical language.
-{canaccess} is a data package that provides public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas.
+{canaccess} is a data package that disseminates public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas.
 We calculate these estimates for each Dissemination Area (DA) within these regions for the years 2019 and 2023.
@@
 The objective of this paper is to describe the {canaccesR} open data package.
-Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in the years 2019 and 2023.
-These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility, using the {r5r} R package [@pereiraR5rRapidRealistic2021].
-We used transport and street network, population, and employment data from different sources to estimate these TTM's datasets.
+Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023.
+These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} R package [@pereiraR5rRapidRealistic2021].
+We used public transit schedule, transport network, population, and business location data from different sources to estimate these TTM's datasets.
 
 Estimating accessibility - the potential offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012] - requires specialized datasets and technical expertise.
-Despite recent advancements in the field, ready-to-use available data on transportation accessibility is still sparse.
-Thus, we leverage technical knowledge to make publicly available ready-to-use data on transportation accessibility.
-To create the package, we integrated and processed raw data from diverse sources, estimated TTM's for two destinations types (e.g., jobs and groceries) across the largest cities in the country, and distributed these findings through this transparent and open source data product.
-Our main contribution is to provide analysis-ready data for Canada's largest cities, thus making the fields of accessibility research and urban analytics more accessible.
+Recent efforts following an open-source and transparent philosophy have been made to disseminate useful data and information on transportation in the Canadian context [@soukhovTTS2016RDataSet2023]. 
+However, despite these initiatives, pre-processed and available data on transportation accessibility is scarse.
+Within this context, we expect to help filling this gap by the processing raw data into user-ready data and making them publicly available to advance knowledge on the field.
+
+To create the package, we integrated and processed raw data from diverse sources, estimated TTM's for two destinations types (*e.g.*, jobs and groceries) across the largest cities in the country, and distributed these findings through this transparent and open source data product.
+Our main contribution is to provide analysis-ready data for Canada's largest cities on the topic of transportation accessibility, thus making urban analytics in the country more accessible and contributing to future research.
 
 The package's main audiences are Canadian researchers in urban planning and transportation and transportation system agencies.
-We anticipate three main uses for the open data product (ODP) described in this paper.
+We anticipate three primary uses for the open data product (ODP) described in this paper.
 First, the datasets allow for static assessment of the level of public transit accessibility across the country's largest cities before and after the COVID-19 pandemic.
-In other words, we make it easier for those interested to compare cities and their level of public transit accessibility to essential destinations (such as employment centers and groceries stores).
-Second, researchers can use the package's information as inputs to evaluate changes in accessibility throughout this period. 
-Researchers may also investigate disparities in these changes across space within those cities, given the spacial character of the datasets made available here.
-Third, as is common practice in transportation accessibility research, these evaluations can substantiate broader investigations on matters of justice and equity in transportation.
+In other words, {canaccesR} makes it easier for those interested in comparing cities regarding their level of public transit accessibility to essential destinations (such as employment centers and groceries stores) to do so.
+Second, the temporal and spatial characters of the datasets made available here allow researchers to evaluate accessibility changes through time and across space within the largest Canadian urban areas.
+Third, as is now common practice in transportation accessibility research, used as inputs, these estimates can substantiate broader investigations on transportation justice and equity [@higginsChangesAccessibilityEmergency2021; @humbertoHowTranslateJustice2023; @pereiraGeographicAccessCOVID192021].
 For example, the TTM estimates allow for evaluating the evolution of public transit's accessibility by income or spatial distribution across all Dissemination Areas (DA's) of each of the 12 cities in the sample [@pargaDemocraticAccessOur2024].
+In other words, the package's contents can be used from straightforward assessments of accessibility in Canadian urban areas to more theoretically and morally complex evaluations of justice in the country's urban transportation system.
 
@@
 
-# Data Sources
+# Data and methods
+
+## Raw data sources
 
@@
 These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
+We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
 
@@
 We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.
-We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
 
 <!-- ## General Transit Feed Specification -->
-We used the GTFS files from all transit agencies within the selected CMA's to obtain the information on their public transit schedule from 2019 to 2023.
+We manually collected and processed the GTFS files from all transit agencies within the selected CMA's to use their information on the public transit schedule in 2019 and 2023.
 <!-- ## OpenStreetMaps -->
-The OpenStreetMap data provided information on the transit network and was necessary for the routing used to estimate the travel time matrices.
+The OpenStreetMap data for the selected areas were collected through the {osmextract} package [@gilardiOsmextractDownloadImport2025].
+We used OSM data from 2019 and 2023, which provided information on the areas' transit network in two points in time.
 <!-- ## Census Data -->
-We used the 2016 and 2021 Canadian Census' information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016, governmentofcanada2021CensusPopulation2021].
-We aggregayed all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
+We collected data from the 2016 Canadian Census using the {cancensus} R package [@vonbergmannCancensusPackageAccess2022] and used its information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016].
 <!-- ## DMTI’s Enhanced Points of Interest -->
-Finally, we use the DMTI’s Enhanced Points of Interest dataset to obtain the location of the groceries stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015]. 
+Finally, we gathered and cleaned the 2023 DMTI’s Enhanced Points of Interest dataset to obtain the location of the groceries stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015].
 We filtered the locations within the DMTI dataset using the grocery stores code from the North American Industry Classification System (NAICS) and the Standard Industrial Classification (SIC).
+{**THE CODE FOR THIS ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE transit_death_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A PROBLEM?**}
 
-# Data processing
+## Methods: travel time matrices processing
 
-Routing, cleaning, etc.
+Using the {r5r} R package, we estimated public transit travel times for two destination types, grocery stores and jobs. 
+For each amenity type, we chose a likely travel time and day of the week. 
+We set a 15 minutes time window and the maximum trip duration to 120 minutes. 
+For groceries stores, we set the departure date to a weekend afternoon and the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and April 22, 2023.
+For employment, we ran the analysis on a typical weekday morning rush-hour commute, more specifically 8:00 to 8:15 AM departure on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 [^3].
+In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
+We aggregated all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
+[^3]: The one exception is Quebec City, where the routing for 2019 occurs on a Saturday and Tuesday in June (instead of April) due to the GTFS data unavailability.
 
@@
 Moreover, these datasets can be used in further equity assessments that evaluate the distribution of accessibility across space and between social groups.
+Furthermore, in the spirit of open data products [@arribas-belOpenDataProductsA2021], the package can be expanded through collaboration with other researchers by, for example, including travel time matrices to other essential destinations within the DMTI's dataset (*e.g.*, schools, healthcare, etc.).
 In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [@martensTransportJusticeDesigning2016], and the inherent connection between public transit and the "right to the city" [@cogginRightTransportMoving2015].

### 2025-03-07 paper/Manuscript-Data-Package.tex
Data and methods
Finished data sources and methods; added other ODP references; descriptive statistics and how to use still incomplete

@@
 product (ODPs) created using the R statistical language. \{canaccess\}
-is a data package that provides public transit travel time estimates
+is a data package that disseminates public transit travel time estimates
 (travel time matrices - TTM) to employment locations and grocery stores
@@
 matrices (TTM) estimates to employment and groceries stores from the 12
-largest Canadian metropolitan areas in the years 2019 and 2023. These
-estimates were created by leveraging expertise in data science, computer
-programming, and transportation accessibility, using the \{r5r\} R
-package \citep{pereiraR5rRapidRealistic2021}. We used transport and
-street network, population, and employment data from different sources
-to estimate these TTM's datasets.
+largest Canadian metropolitan areas in 2019 and 2023. These estimates
+were created by leveraging expertise in data science, computer
+programming, and transportation accessibility using the \{r5r\} R
+package \citep{pereiraR5rRapidRealistic2021}. We used public transit
+schedule, transport network, population, and business location data from
+different sources to estimate these TTM's datasets.
 
@@
 \citep{paezMeasuringAccessibilityPositive2012} - requires specialized
-datasets and technical expertise. Despite recent advancements in the
-field, ready-to-use available data on transportation accessibility is
-still sparse. Thus, we leverage technical knowledge to make publicly
-available ready-to-use data on transportation accessibility. To create
-the package, we integrated and processed raw data from diverse sources,
-estimated TTM's for two destinations types (e.g., jobs and groceries)
-across the largest cities in the country, and distributed these findings
-through this transparent and open source data product. Our main
-contribution is to provide analysis-ready data for Canada's largest
-cities, thus making the fields of accessibility research and urban
-analytics more accessible.
+datasets and technical expertise. Recent efforts following an
+open-source and transparent philosophy have been made to disseminate
+useful data and information on transportation in the Canadian context
+\citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
+pre-processed and available data on transportation accessibility is
+scarse. Within this context, we expect to help filling this gap by the
+processing raw data into user-ready data and making them publicly
+available to advance knowledge on the field.
+
+To create the package, we integrated and processed raw data from diverse
+sources, estimated TTM's for two destinations types (\emph{e.g.}, jobs
+and groceries) across the largest cities in the country, and distributed
+these findings through this transparent and open source data product.
+Our main contribution is to provide analysis-ready data for Canada's
+largest cities on the topic of transportation accessibility, thus making
+urban analytics in the country more accessible and contributing to
+future research.
 
@@
 and transportation and transportation system agencies. We anticipate
-three main uses for the open data product (ODP) described in this paper.
-First, the datasets allow for static assessment of the level of public
-transit accessibility across the country's largest cities before and
-after the COVID-19 pandemic. In other words, we make it easier for those
-interested to compare cities and their level of public transit
-accessibility to essential destinations (such as employment centers and
-groceries stores). Second, researchers can use the package's information
-as inputs to evaluate changes in accessibility throughout this period.
-Researchers may also investigate disparities in these changes across
-space within those cities, given the spacial character of the datasets
-made available here. Third, as is common practice in transportation
-accessibility research, these evaluations can substantiate broader
-investigations on matters of justice and equity in transportation. For
-example, the TTM estimates allow for evaluating the evolution of public
-transit's accessibility by income or spatial distribution across all
-Dissemination Areas (DA's) of each of the 12 cities in the sample
-\citep{pargaDemocraticAccessOur2024}.
+three primary uses for the open data product (ODP) described in this
+paper. First, the datasets allow for static assessment of the level of
+public transit accessibility across the country's largest cities before
+and after the COVID-19 pandemic. In other words, \{canaccesR\} makes it
+easier for those interested in comparing cities regarding their level of
+public transit accessibility to essential destinations (such as
+employment centers and groceries stores) to do so. Second, the temporal
+and spatial characters of the datasets made available here allow
+researchers to evaluate accessibility changes through time and across
+space within the largest Canadian urban areas. Third, as is now common
+practice in transportation accessibility research, used as inputs, these
+estimates can substantiate broader investigations on transportation
+justice and equity
+\citep{higginsChangesAccessibilityEmergency2021, humbertoHowTranslateJustice2023, pereiraGeographicAccessCOVID192021}.
+For example, the TTM estimates allow for evaluating the evolution of
+public transit's accessibility by income or spatial distribution across
+all Dissemination Areas (DA's) of each of the 12 cities in the sample
+\citep{pargaDemocraticAccessOur2024}. In other words, the package's
+contents can be used from straightforward assessments of accessibility
+in Canadian urban areas to more theoretically and morally complex
+evaluations of justice in the country's urban transportation system.
 
@@
 
-\section{Data Sources}\label{data-sources}
+\section{Data and methods}\label{data-and-methods}
+
+\subsection{Raw data sources}\label{raw-data-sources}
 
@@
   Vancouver metropolitan area because of the former's proximity to a
-  transit station on the region's West Coast Express commuter rail line.
-  We used four main data sources to construct the \{canaccesR\} data
-  package: General Transit Feed Specification (GTFS), OpenStreetMap
-  (OSM), DMTI's Enhanced Points of Interest, and Statistics Canada
-  Census data.}. These locations are Toronto, Montreal, Vancouver,
-Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
-Kitchener-Cambridge-Waterloo, London, and Halifax.
+  transit station on the region's West Coast Express commuter rail line.}.
+These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau,
+Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
+Kitchener-Cambridge-Waterloo, London, and Halifax. We used four main
+data sources to construct the \{canaccesR\} data package: General
+Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
+Points of Interest, and Statistics Canada Census data.
 
-We used the GTFS files from all transit agencies within the selected
-CMA's to obtain the information on their public transit schedule from
-2019 to 2023. The OpenStreetMap data provided information on the transit
-network and was necessary for the routing used to estimate the travel
-time matrices. We used the 2016 and 2021 Canadian Census' information on
-the spatial distribution of the population and the number of workplace
-locations (employment) across the CMA's
-\citep[governmentofcanada2021CensusPopulation2021]{governmentofcanada2016CensusPopulation2016}.
-We aggregayed all the resulting travel time matrices at the
-Dissemination Area (DA) level, which comprise the fundamental unit of
-analysis in data package. Finally, we use the DMTI's Enhanced Points of
-Interest dataset to obtain the location of the groceries stores within
-every urban area selected
-\citep{dmtispatialincEnhancedPointsInterest2015}. We filtered the
-locations within the DMTI dataset using the grocery stores code from the
-North American Industry Classification System (NAICS) and the Standard
-Industrial Classification (SIC).
+We manually collected and processed the GTFS files from all transit
+agencies within the selected CMA's to use their information on the
+public transit schedule in 2019 and 2023. The OpenStreetMap data for the
+selected areas were collected through the \{osmextract\} package
+\citep{gilardiOsmextractDownloadImport2025}. We used OSM data from 2019
+and 2023, which provided information on the areas' transit network in
+two points in time. We collected data from the 2016 Canadian Census
+using the \{cancensus\} R package
+\citep{vonbergmannCancensusPackageAccess2022} and used its information
+on the spatial distribution of the population and the number of
+workplace locations (employment) across the CMA's
+\citep{governmentofcanada2016CensusPopulation2016}. Finally, we gathered
+and cleaned the 2023 DMTI's Enhanced Points of Interest dataset to
+obtain the location of the groceries stores within every urban area
+selected \citep{dmtispatialincEnhancedPointsInterest2015}. We filtered
+the locations within the DMTI dataset using the grocery stores code from
+the North American Industry Classification System (NAICS) and the
+Standard Industrial Classification (SIC). \{\textbf{THE CODE FOR THIS
+ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE
+transit\_death\_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A
+PROBLEM?}\}
 
-\section{Data processing}\label{data-processing}
+\subsection{Methods: travel time matrices
+processing}\label{methods-travel-time-matrices-processing}
 
-Routing, cleaning, etc.
+Using the \{r5r\} R package, we estimated public transit travel times
+for two destination types, grocery stores and jobs. For each amenity
+type, we chose a likely travel time and day of the week. We set a 15
+minutes time window and the maximum trip duration to 120 minutes. For
+groceries stores, we set the departure date to a weekend afternoon and
+the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and
+April 22, 2023. For employment, we ran the analysis on a typical weekday
+morning rush-hour commute, more specifically 8:00 to 8:15 AM departure
+on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 {[}\^{}3{]}. In
+both cases, we assumed that walking was the mode of travel from origin
+to transit stop and from transit stop to destination. We aggregated all
+the resulting travel time matrices at the Dissemination Area (DA) level,
+which comprise the fundamental unit of analysis in data package.
+{[}\^{}3{]}: The one exception is Quebec City, where the routing for
+2019 occurs on a Saturday and Tuesday in June (instead of April) due to
+the GTFS data unavailability.
 
@@
 can be used in further equity assessments that evaluate the distribution
-of accessibility across space and between social groups. In other words,
-we hope that by making these datasets publicly available, future
-analysis can contribute to making Canada's transportation system more
-just and fair, considering accessibility's as the main social good of
+of accessibility across space and between social groups. Furthermore, in
+the spirit of open data products
+\citep{arribas-belOpenDataProductsA2021}, the package can be expanded
+through collaboration with other researchers by, for example, including
+travel time matrices to other essential destinations within the DMTI's
+dataset (\emph{e.g.}, schools, healthcare, etc.). In other words, we
+hope that by making these datasets publicly available, future analysis
+can contribute to making Canada's transportation system more just and
+fair, considering accessibility's as the main social good of
 transportation \citep{martensTransportJusticeDesigning2016}, and the

### 2025-03-09 paper/Manuscript-Data-Package.Rmd
describe travel time matrices
describe travel time matrices; next step is estimate descriptive statistics.

@@
 abstract: 
-  "In this paper, we describe the {canaccesR} package, an open data product (ODPs) created using the R statistical language.
-{canaccess} is a data package that disseminates public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas.
-We calculate these estimates for each Dissemination Area (DA) within these regions for the years 2019 and 2023.
+  "In this paper, we describe the {canaccesR} package, an open data product (ODP) created in R that contains public transit travel time estimates to employment locations and grocery stores across Canada's 12 largest metropolitan areas.
+We calculate travel time matrices (TTM) from and to each Dissemination Area (DA) within these regions for the years 2019 and 2023.
+We add value to the urban analytics community by processing and integrating raw data, and disseminating user-ready data in the domain of transportation accessibility in Canada.
 To do so, we use the {r5r} R package, General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
-This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility to essential services across these regions.
-These estimates can be used to compare different regions across Canada in terms of their accessibility and to conduct within-region equity assessments regarding access to services, which can inform improvements in transportation policies related to accessibility.
+This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility levels to these two essential destinations within these urban areas.
+Moreover, these estimations can be used as inputs in equity and inequalities assessments, through the comparisons of within and across accessibility levels found throughout Canada's largest metropolitan areas.
+Consequently, we expect to contribute to informed and data-based decision making in transportation by disseminating these data.
+We hope that these datasets can substantiate future improvements in policy-making that may lead to greater justice in the country's urban transportation systems.
 The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities).
@@
 Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023.
-These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} R package [@pereiraR5rRapidRealistic2021].
+The results refer to all the Dissemination Areas (DA) (the smallest publicly available geography provided by Statistics Canada [@governmentofcanadaDictionaryCensusPopulation2021a]) within each metropolitan area, yielding origin-destination pairs containing information on travel time by public transit, population, and total employment within those spatial units.
+These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} package [@pereiraR5rRapidRealistic2021].
 We used public transit schedule, transport network, population, and business location data from different sources to estimate these TTM's datasets.
@@
 Recent efforts following an open-source and transparent philosophy have been made to disseminate useful data and information on transportation in the Canadian context [@soukhovTTS2016RDataSet2023]. 
-However, despite these initiatives, pre-processed and available data on transportation accessibility is scarse.
-Within this context, we expect to help filling this gap by the processing raw data into user-ready data and making them publicly available to advance knowledge on the field.
-
-To create the package, we integrated and processed raw data from diverse sources, estimated TTM's for two destinations types (*e.g.*, jobs and groceries) across the largest cities in the country, and distributed these findings through this transparent and open source data product.
-Our main contribution is to provide analysis-ready data for Canada's largest cities on the topic of transportation accessibility, thus making urban analytics in the country more accessible and contributing to future research.
+However, despite these initiatives, pre-processed and available data that allow for the ease estimation of accessibility indicators are still scarce.
+Within this context, we expect to help filling this gap by the processing raw into user-ready data and making them publicly available to advance knowledge on the field.
+Our main contribution is to provide analysis-ready data for Canada's largest cities on the topic of transportation accessibility, thus making urban analytics in the country more accessible and contributing to future research and data-based decision making.
 
@@
 
-The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1].
+The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^2].
 These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
@@
 
-[^1]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter.
+[^2]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter.
 We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.
@@
 <!-- ## Census Data -->
-We collected data from the 2016 Canadian Census using the {cancensus} R package [@vonbergmannCancensusPackageAccess2022] and used its information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016].
+We collected data from the 2016 Canadian Census using the {cancensus} package [@vonbergmannCancensusPackageAccess2022] and used its information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016].
 <!-- ## DMTI’s Enhanced Points of Interest -->
@@
 
-Using the {r5r} R package, we estimated public transit travel times for two destination types, grocery stores and jobs. 
+Using the {r5r} package, we estimated public transit travel times for two destination types, grocery stores and jobs. 
 For each amenity type, we chose a likely travel time and day of the week. 
 We set a 15 minutes time window and the maximum trip duration to 120 minutes. 
+The estimated times are the median of the 15 minute time window.
 For groceries stores, we set the departure date to a weekend afternoon and the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and April 22, 2023.
@@
 We aggregated all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
+
 [^3]: The one exception is Quebec City, where the routing for 2019 occurs on a Saturday and Tuesday in June (instead of April) due to the GTFS data unavailability.
@@
 
-The main contents of the {canaccessR} package are the travel time matrices estimates for all the 12 largest Canadian cities.
+## Travel time matrices
+
+The main contents of the {canaccessR} package are the travel time matrices.
+<!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  -->
+These matrices (data frames) comprise the estimated travel time by transit from each origin and destination Dissemination Area pairs.
+The matrices also contain identifiers, such as the unique origin and destination DA codes, region identifier (*e.g.*, "cal" for the region Calgary) and name, the year in which the travel times refers to, and the date and time of departure for a trip that originates at the Dissemination Area of origin.
+Besides these information, the matrices feature the total population in the origin DA and the total employment in the destination DA (both according to the 2016 Census).
+There are two matrices for each metropolitan area, one containing the travel times to jobs and the other to groceries stores, making it 20 datasets (the information from Toronto, Hamilton, and Waterloo are aggregated at the Greater Golden Horseshoe travel time matrices).
+
+**HOW TO MAKE THE INFORMATION FROM THE TTM INTO SPATIAL DATA?**
+The information can be turned into spatial data using the
+
+Below, we present some descriptive statistics from the travel time matrices contained in the {canaccessR} package.
+
+## Other contents
 
@@
 In addition, the package also contains aggregated population statistics (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale.
-
-## Descriptive statistics
-
-Below, we present some of the basic statistics from the travel time matrices contained in the {canaccessR} package.
 
@@
 
-In this paper, we describe the {canaccesR} data package, created using the {r5r} R package and transit schedule, street network, employment, and population data.
+In this paper, we describe the {canaccesR} data package, created using the {r5r} package and transit schedule, street network, employment, and population data.
 The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.

### 2025-03-09 paper/Manuscript-Data-Package.tex
describe travel time matrices
describe travel time matrices; next step is estimate descriptive statistics.

@@
 In this paper, we describe the \{canaccesR\} package, an open data
-product (ODPs) created using the R statistical language. \{canaccess\}
-is a data package that disseminates public transit travel time estimates
-(travel time matrices - TTM) to employment locations and grocery stores
-across the 12 largest Canadian metropolitan areas. We calculate these
-estimates for each Dissemination Area (DA) within these regions for the
-years 2019 and 2023. To do so, we use the \{r5r\} R package, General
-Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
-Points of Interest, and Statistics Canada Census data. This data package
-can be used by researchers, practitioners, and transit agencies to
-estimate accessibility to essential services across these regions. These
-estimates can be used to compare different regions across Canada in
-terms of their accessibility and to conduct within-region equity
-assessments regarding access to services, which can inform improvements
-in transportation policies related to accessibility. The package is
-still in its initial phase and may undergo expansions in the future by
-adding TTM's for other destinations (e.g., schools, healthcare
-facilities). Finally, as an ODPs, the \{canaccess\} package allows for
-open exploration, use, and contribution by users through its GitHub
-repository.
+product (ODP) created in R that contains public transit travel time
+estimates to employment locations and grocery stores across Canada's 12
+largest metropolitan areas. We calculate travel time matrices (TTM) from
+and to each Dissemination Area (DA) within these regions for the years
+2019 and 2023. We add value to the urban analytics community by
+processing and integrating raw data, and disseminating user-ready data
+in the domain of transportation accessibility in Canada. To do so, we
+use the \{r5r\} R package, General Transit Feed Specification (GTFS),
+OpenStreetMap (OSM), DMTI's Enhanced Points of Interest, and Statistics
+Canada Census data. This data package can be used by researchers,
+practitioners, and transit agencies to estimate accessibility levels to
+these two essential destinations within these urban areas. Moreover,
+these estimations can be used as inputs in equity and inequalities
+assessments, through the comparisons of within and across accessibility
+levels found throughout Canada's largest metropolitan areas.
+Consequently, we expect to contribute to informed and data-based
+decision making in transportation by disseminating these data. We hope
+that these datasets can substantiate future improvements in
+policy-making that may lead to greater justice in the country's urban
+transportation systems. The package is still in its initial phase and
+may undergo expansions in the future by adding TTM's for other
+destinations (e.g., schools, healthcare facilities). Finally, as an
+ODPs, the \{canaccess\} package allows for open exploration, use, and
+contribution by users through its GitHub repository.
 \end{abstract}
@@
 matrices (TTM) estimates to employment and groceries stores from the 12
-largest Canadian metropolitan areas in 2019 and 2023. These estimates
-were created by leveraging expertise in data science, computer
-programming, and transportation accessibility using the \{r5r\} R
-package \citep{pereiraR5rRapidRealistic2021}. We used public transit
-schedule, transport network, population, and business location data from
-different sources to estimate these TTM's datasets.
+largest Canadian metropolitan areas in 2019 and 2023. The results refer
+to all the Dissemination Areas (DA) (the smallest publicly available
+geography provided by Statistics Canada
+\citep{governmentofcanadaDictionaryCensusPopulation2021a}) within each
+metropolitan area, yielding origin-destination pairs containing
+information on travel time by public transit, population, and total
+employment within those spatial units. These estimates were created by
+leveraging expertise in data science, computer programming, and
+transportation accessibility using the \{r5r\} package
+\citep{pereiraR5rRapidRealistic2021}. We used public transit schedule,
+transport network, population, and business location data from different
+sources to estimate these TTM's datasets.
 
@@
 \citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
-pre-processed and available data on transportation accessibility is
-scarse. Within this context, we expect to help filling this gap by the
-processing raw data into user-ready data and making them publicly
-available to advance knowledge on the field.
-
-To create the package, we integrated and processed raw data from diverse
-sources, estimated TTM's for two destinations types (\emph{e.g.}, jobs
-and groceries) across the largest cities in the country, and distributed
-these findings through this transparent and open source data product.
-Our main contribution is to provide analysis-ready data for Canada's
-largest cities on the topic of transportation accessibility, thus making
-urban analytics in the country more accessible and contributing to
-future research.
+pre-processed and available data that allow for the ease estimation of
+accessibility indicators are still scarce. Within this context, we
+expect to help filling this gap by the processing raw into user-ready
+data and making them publicly available to advance knowledge on the
+field. Our main contribution is to provide analysis-ready data for
+Canada's largest cities on the topic of transportation accessibility,
+thus making urban analytics in the country more accessible and
+contributing to future research and data-based decision making.
 
@@
 two points in time. We collected data from the 2016 Canadian Census
-using the \{cancensus\} R package
+using the \{cancensus\} package
 \citep{vonbergmannCancensusPackageAccess2022} and used its information
@@
 
-Using the \{r5r\} R package, we estimated public transit travel times
-for two destination types, grocery stores and jobs. For each amenity
-type, we chose a likely travel time and day of the week. We set a 15
-minutes time window and the maximum trip duration to 120 minutes. For
-groceries stores, we set the departure date to a weekend afternoon and
-the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and
-April 22, 2023. For employment, we ran the analysis on a typical weekday
-morning rush-hour commute, more specifically 8:00 to 8:15 AM departure
-on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 {[}\^{}3{]}. In
-both cases, we assumed that walking was the mode of travel from origin
-to transit stop and from transit stop to destination. We aggregated all
-the resulting travel time matrices at the Dissemination Area (DA) level,
-which comprise the fundamental unit of analysis in data package.
-{[}\^{}3{]}: The one exception is Quebec City, where the routing for
-2019 occurs on a Saturday and Tuesday in June (instead of April) due to
-the GTFS data unavailability.
+Using the \{r5r\} package, we estimated public transit travel times for
+two destination types, grocery stores and jobs. For each amenity type,
+we chose a likely travel time and day of the week. We set a 15 minutes
+time window and the maximum trip duration to 120 minutes. The estimated
+times are the median of the 15 minute time window. For groceries stores,
+we set the departure date to a weekend afternoon and the departure time
+to between 12:00 PM to 12:15 PM on April 20, 2019 and April 22, 2023.
+For employment, we ran the analysis on a typical weekday morning
+rush-hour commute, more specifically 8:00 to 8:15 AM departure on
+Tuesday, April 16, 2019 and Tuesday, April 18, 2023 \footnote{The one
+  exception is Quebec City, where the routing for 2019 occurs on a
+  Saturday and Tuesday in June (instead of April) due to the GTFS data
+  unavailability.}. In both cases, we assumed that walking was the mode
+of travel from origin to transit stop and from transit stop to
+destination. We aggregated all the resulting travel time matrices at the
+Dissemination Area (DA) level, which comprise the fundamental unit of
+analysis in data package.
 
@@
 
+\subsection{Travel time matrices}\label{travel-time-matrices}
+
 The main contents of the \{canaccessR\} package are the travel time
-matrices estimates for all the 12 largest Canadian cities.
+matrices. These matrices (data frames) comprise the estimated travel
+time by transit from each origin and destination Dissemination Area
+pairs. The matrices also contain identifiers, such as the unique origin
+and destination DA codes, region identifier (\emph{e.g.}, ``cal'' for
+the region Calgary) and name, the year in which the travel times refers
+to, and the date and time of departure for a trip that originates at the
+Dissemination Area of origin. Besides these information, the matrices
+feature the total population in the origin DA and the total employment
+in the destination DA (both according to the 2016 Census). There are two
+matrices for each metropolitan area, one containing the travel times to
+jobs and the other to groceries stores, making it 20 datasets (the
+information from Toronto, Hamilton, and Waterloo are aggregated at the
+Greater Golden Horseshoe travel time matrices).
+
+\textbf{HOW TO MAKE THE INFORMATION FROM THE TTM INTO SPATIAL DATA?} The
+information can be turned into spatial data using the
+
+Below, we present some descriptive statistics from the travel time
+matrices contained in the \{canaccessR\} package.
+
+\subsection{Other contents}\label{other-contents}
 
@@
 
-\subsection{Descriptive statistics}\label{descriptive-statistics}
-
-Below, we present some of the basic statistics from the travel time
-matrices contained in the \{canaccessR\} package.
-
 \section{How to use \{canaccessR\}}\label{how-to-use-canaccessr}
@@
 In this paper, we describe the \{canaccesR\} data package, created using
-the \{r5r\} R package and transit schedule, street network, employment,
+the \{r5r\} package and transit schedule, street network, employment,
 and population data. The package's main contents refers to the

### 2025-03-10 paper/Manuscript-Data-Package.Rmd
generate descriptive statistics
create a separate R file to generate the descriptive statistics used in the paper

@@
 
-keywords: Accessibility; public transit; open data products (ODPs); travel time; employment; grocery stores.
+keywords: Public transit accessibility; open data products (ODPs); R data package; travel time matrices.
 classoption:
@@
 Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023.
-The results refer to all the Dissemination Areas (DA) (the smallest publicly available geography provided by Statistics Canada [@governmentofcanadaDictionaryCensusPopulation2021a]) within each metropolitan area, yielding origin-destination pairs containing information on travel time by public transit, population, and total employment within those spatial units.
+The results are stored at the Dissemination Areas (DA) [^1] level, yielding origin-destination pairs containing information on travel time by public transit, population, and total employment within each metropolitan area.
 These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} package [@pereiraR5rRapidRealistic2021].
 We used public transit schedule, transport network, population, and business location data from different sources to estimate these TTM's datasets.
+
+[^1]: Dissemination Areas are the smallest publicly available spatial unit provided by Statistics Canada [@governmentofcanadaDictionaryCensusPopulation2021a].
 
@@
 
-The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^2].
+The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2016 Canadian Census [@governmentofcanada2016CensusPopulation2016] [^2].
 These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
@@
 
-## Travel time matrices
+<!-- ## Travel time matrices -->
+
+The package contains the following contents: travel time matrices, socio-economic and demographic data disaggregated at the DA level, the CMA areas' boundaries and backgrounds for plotting the data spatially, and aggregated statistics.
 
@@
 <!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  -->
-These matrices (data frames) comprise the estimated travel time by transit from each origin and destination Dissemination Area pairs.
-The matrices also contain identifiers, such as the unique origin and destination DA codes, region identifier (*e.g.*, "cal" for the region Calgary) and name, the year in which the travel times refers to, and the date and time of departure for a trip that originates at the Dissemination Area of origin.
-Besides these information, the matrices feature the total population in the origin DA and the total employment in the destination DA (both according to the 2016 Census).
-There are two matrices for each metropolitan area, one containing the travel times to jobs and the other to groceries stores, making it 20 datasets (the information from Toronto, Hamilton, and Waterloo are aggregated at the Greater Golden Horseshoe travel time matrices).
+These matrices (datasets) comprise the estimated travel time by transit from and to each origin and destination Dissemination Area pairs.
+The datasets also feature the total population in the origin DA and the total employment in the destination DA.
+Besides these information, the matrices contain the unique origin and destination DA codes, region and name identifiers, the year in which the travel times refers to, and the date and time of departure for a trip that originates at the Dissemination Area of origin.
+There are two matrices for each metropolitan area, one containing the travel times to jobs and the other to groceries stores [^4].
+These information can be turned into spatial data using the socio-economic and demographic datasets.
 
-**HOW TO MAKE THE INFORMATION FROM THE TTM INTO SPATIAL DATA?**
-The information can be turned into spatial data using the
+[^4]: The information from Toronto, Hamilton, and Waterloo are aggregated at the Greater Golden Horseshoe travel time matrices, making it 20 datasets in total.
+
+The socio-economic and demographic data contain other information disaggregated at the DA level for each CMA. 
+These refer to total population by age groups, number of dwellings, number of individuals below the Low Income Measure, etc.
+The aggregated statistics refer to population aggregates (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale.
 
@@
 
-## Other contents
+```{r}
 
-Other sets of data are also available at the {canaccessR} package besides the travel time matrices.
-These are the boundaries, socio-economic and demographic data (e.g., population, number of dwellings, number of individuals below the Low Income Measure, etc.) of the selected CMA's, disaggregated by DA.
-In addition, the package also contains aggregated population statistics (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale.
+```
 

### 2025-03-10 paper/Manuscript-Data-Package.tex
generate descriptive statistics
create a separate R file to generate the descriptive statistics used in the paper

@@
 
-\keywords{Accessibility; public transit; open data products (ODPs);
-travel time; employment; grocery stores.}
+\keywords{Public transit accessibility; open data products (ODPs); R
+data package; travel time matrices.}
 
@@
 matrices (TTM) estimates to employment and groceries stores from the 12
-largest Canadian metropolitan areas in 2019 and 2023. The results refer
-to all the Dissemination Areas (DA) (the smallest publicly available
-geography provided by Statistics Canada
-\citep{governmentofcanadaDictionaryCensusPopulation2021a}) within each
-metropolitan area, yielding origin-destination pairs containing
-information on travel time by public transit, population, and total
-employment within those spatial units. These estimates were created by
-leveraging expertise in data science, computer programming, and
-transportation accessibility using the \{r5r\} package
+largest Canadian metropolitan areas in 2019 and 2023. The results are
+stored at the Dissemination Areas (DA) \footnote{Dissemination Areas are
+  the smallest publicly available spatial unit provided by Statistics
+  Canada \citep{governmentofcanadaDictionaryCensusPopulation2021a}.}
+level, yielding origin-destination pairs containing information on
+travel time by public transit, population, and total employment within
+each metropolitan area. These estimates were created by leveraging
+expertise in data science, computer programming, and transportation
+accessibility using the \{r5r\} package
 \citep{pereiraR5rRapidRealistic2021}. We used public transit schedule,
@@
 The locations included in the data package comprise the 12 largest
-(population-wise) Census metropolitan areas (CMA's) based on the 2021
-Canadian Census \citep{governmentofcanada2021CensusPopulation2021}
+(population-wise) Census metropolitan areas (CMA's) based on the 2016
+Canadian Census \citep{governmentofcanada2016CensusPopulation2016}
 \footnote{We included Oshawa, Ontario, as part of the Greater Toronto
@@
 
-\subsection{Travel time matrices}\label{travel-time-matrices}
+The package contains the following contents: travel time matrices,
+socio-economic and demographic data disaggregated at the DA level, the
+CMA areas' boundaries and backgrounds for plotting the data spatially,
+and aggregated statistics.
 
 The main contents of the \{canaccessR\} package are the travel time
-matrices. These matrices (data frames) comprise the estimated travel
-time by transit from each origin and destination Dissemination Area
-pairs. The matrices also contain identifiers, such as the unique origin
-and destination DA codes, region identifier (\emph{e.g.}, ``cal'' for
-the region Calgary) and name, the year in which the travel times refers
-to, and the date and time of departure for a trip that originates at the
-Dissemination Area of origin. Besides these information, the matrices
-feature the total population in the origin DA and the total employment
-in the destination DA (both according to the 2016 Census). There are two
-matrices for each metropolitan area, one containing the travel times to
-jobs and the other to groceries stores, making it 20 datasets (the
-information from Toronto, Hamilton, and Waterloo are aggregated at the
-Greater Golden Horseshoe travel time matrices).
+matrices. These matrices (datasets) comprise the estimated travel time
+by transit from and to each origin and destination Dissemination Area
+pairs. The datasets also feature the total population in the origin DA
+and the total employment in the destination DA. Besides these
+information, the matrices contain the unique origin and destination DA
+codes, region and name identifiers, the year in which the travel times
+refers to, and the date and time of departure for a trip that originates
+at the Dissemination Area of origin. There are two matrices for each
+metropolitan area, one containing the travel times to jobs and the other
+to groceries stores \footnote{The information from Toronto, Hamilton,
+  and Waterloo are aggregated at the Greater Golden Horseshoe travel
+  time matrices, making it 20 datasets in total.}. These information can
+be turned into spatial data using the socio-economic and demographic
+datasets.
 
-\textbf{HOW TO MAKE THE INFORMATION FROM THE TTM INTO SPATIAL DATA?} The
-information can be turned into spatial data using the
+The socio-economic and demographic data contain other information
+disaggregated at the DA level for each CMA. These refer to total
+population by age groups, number of dwellings, number of individuals
+below the Low Income Measure, etc. The aggregated statistics refer to
+population aggregates (for the selected CMA's) and transit revenue and
+ridership data aggregated by regional and national scale.
 
@@
 matrices contained in the \{canaccessR\} package.
-
-\subsection{Other contents}\label{other-contents}
-
-Other sets of data are also available at the \{canaccessR\} package
-besides the travel time matrices. These are the boundaries,
-socio-economic and demographic data (e.g., population, number of
-dwellings, number of individuals below the Low Income Measure, etc.) of
-the selected CMA's, disaggregated by DA. In addition, the package also
-contains aggregated population statistics (for the selected CMA's) and
-transit revenue and ridership data aggregated by regional and national
-scale.
 

### 2025-03-10 paper/descriptive_stats_paper.R

@@ -1,0 +1,99 @@
+
+# library -----------------------------------------------------------------
+
+
+library(tidyverse)
+library(canaccessR)
+library(sf)
+library(data.table)
+
+
+# read data ---------------------------------------------------------------
+
+
+# c_regions <- c(
+#   "cal"
+#   , "edm"
+#   # , "ggh"
+#   # , "hal"
+#   # , "ldn"
+#   # , "mtl"
+#   # , "ott"
+#   # , "que"
+#   # , "van"
+#   # , "win"
+# )
+# 
+# c_dest_types <- c("emp", "grc")
+# 
+# 
+# 
+# vec_matrices <- map2_chr(
+#   .x = c_dest_types
+#   , .y = c_regions
+#   , ~paste0("travel_matrix_", .x, "_", .y)
+# )
+# 
+# l_teste <- purrr::map(
+#   .x = vec_matrices
+#   , ~data(.x)
+# )
+
+f_ttm <- function(){
+  # datasets from canaccessR package
+  d <- data(package = "canaccessR")
+  # get travel matrices names
+  vec_matrices <- d$results[, "Item"] %>% str_subset("travel_matrix")
+  
+  # call promised data
+  # https://stackoverflow.com/a/27710355
+  data(list = vec_matrices, package = "canaccessR")
+  
+  # separate ggh matrices (if necessary)
+  c_names <- travel_matrix_emp_ggh %>% 
+    pull(study_region_name) %>% 
+    unique() %>% 
+    droplevels.factor() %>% 
+    as.character()
+  
+  df_region <- c("tor", "ham", "wat")
+  
+  c_dest_types <- c("emp", "grc")
+  
+  # purrr into list
+  l_emp_ggh <- purrr::map(
+    .x = c_names
+    , ~dplyr::filter(.data = travel_matrix_emp_ggh, study_region_name == .x)
+  )
+  # rename list elements
+  names(l_emp_ggh) <- paste0("travel_matrix_emp_", df_region)
+  # return list elements into global environment
+  list2env(l_emp_ggh, globalenv())
+  
+  # purrr into list
+  l_grc_ggh <- purrr::map(
+    .x = c_names
+    , ~dplyr::filter(.data = travel_matrix_grc_ggh, study_region_name == .x)
+  )
+  # rename list elements
+  names(l_grc_ggh) <- paste0("travel_matrix_grc_", df_region)
+  # return list elements into global environment
[…]

### 2025-03-11 paper/Manuscript-Data-Package.Rmd
finish descriptive and start plot
finish descriptive statistics table and discussion; started creating dataset for plotting on the how to use session

@@
 # Load packages
-
+library(tidyverse)
+library(data.table)
+# library(summarytools)
+library(gt)
+library(gtsummary)
+# library(sf)
+# library(tmap)
+# library(tmaptools)
 ```
@@
 ```{r load-data, include = FALSE}
-
+load("sk_full.Rdata")
 ```
@@
 
-Below, we present some descriptive statistics from the travel time matrices contained in the {canaccessR} package.
+We now present some descriptive statistics from the travel time matrices contained in the {canaccessR} package. 
+The Table @tbl-table_1 below summarizes the travel time estimates.
+The travel time datasets contain, in total, 97,784,850 origin-destination pairs (observations) from population to employment locations for all the DA's in the sample and 18,519,897 pairs from population to groceries stores.
+Considering all areas combined, the travel times estimated to jobs have an average of 78, standard deviation of 25, 1st quantile of 59, median of 80, and 3rd quantile of 99.
+For travel times to groceries stores, these same statistics are the 76, 26, 56, 76, and 97, respectively.
 
 ```{r}
+#| label: tbl-table_1
+#| echo: false
+#| message: false
+
+sk_full %>% 
+  gt() %>% 
+  fmt_number(use_seps = TRUE, decimals = FALSE) |>  
+  tab_options( 
+    table.font.size = 10,
+    table.width = pct(100) 
+    )
 
@@
 
-This section presents some potential applications of the data package.
+This section exemplifies how the package can be used through visual representation.
+
+A more thorough example of the package's use can be found on the School of Cities' recent report on Canada's Urban Infrastructure Deficit.
+In its 11th Chapter, we use the travel time matrices to estimate accessibility metrics to jobs and groceries stores before and after the pandemic [@pargaDemocraticAccessOur2024].
+We then compare how changes affected groups differently according to their spatial distribution and income level, thus making explicit the connection of the package's information and matters of equity in transportation.
+The report is freely available for download at the State of Cities Summit [website](https://stateofcitiessummit.ca/report).
 

### 2025-03-11 paper/Manuscript-Data-Package.tex
finish descriptive and start plot
finish descriptive statistics table and discussion; started creating dataset for plotting on the how to use session

@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{booktabs}
+\usepackage{caption}
+\usepackage{longtable}
+\usepackage{colortbl}
+\usepackage{array}
+\usepackage{anyfontsize}
+\usepackage{multirow}
 
@@
 
-Below, we present some descriptive statistics from the travel time
-matrices contained in the \{canaccessR\} package.
+We now present some descriptive statistics from the travel time matrices
+contained in the \{canaccessR\} package. The Table \citet{tbl-table_1}
+below summarizes the travel time estimates. The travel time datasets
+contain, in total, 97,784,850 origin-destination pairs (observations)
+from population to employment locations for all the DA's in the sample
+and 18,519,897 pairs from population to groceries stores. Considering
+all areas combined, the travel times estimated to jobs have an average
+of 78, standard deviation of 25, 1st quantile of 59, median of 80, and
+3rd quantile of 99. For travel times to groceries stores, these same
+statistics are the 76, 26, 56, 76, and 97, respectively.
+
+\begingroup
+\setlength\LTleft{0\linewidth}
+\setlength\LTright{0\linewidth}\fontsize{7.5pt}{9.0pt}\selectfont
+\begin{longtable}{@{\extracolsep{\fill}}clrrrrrr}
+\toprule
+Study Region Name & Destination & Observations & Mean & Sd & P25 & P50 & P75 \\ 
+\midrule\addlinespace[2.5pt]
+All regions & Employment & 97,784,850 & 78 & 25 & 59 & 80 & 99 \\ 
+Toronto & Employment & 41,038,062 & 82 & 25 & 64 & 85 & 103 \\ 
+Montréal & Employment & 30,091,411 & 76 & 25 & 57 & 77 & 97 \\ 
+Vancouver & Employment & 12,254,478 & 76 & 25 & 58 & 77 & 96 \\ 
+Calgary & Employment & 3,022,244 & 75 & 23 & 59 & 75 & 91 \\ 
+Ottawa & Employment & 2,936,912 & 74 & 24 & 56 & 74 & 93 \\ 
+Edmonton & Employment & 2,057,486 & 70 & 23 & 53 & 70 & 87 \\ 
+Québec City & Employment & 1,473,702 & 72 & 25 & 54 & 72 & 93 \\ 
+Winnipeg & Employment & 1,574,205 & 62 & 22 & 46 & 61 & 76 \\ 
+Hamilton & Employment & 2,089,172 & 82 & 28 & 61 & 87 & 106 \\ 
+Waterloo & Employment & 552,826 & 71 & 27 & 49 & 69 & 93 \\ 
+London & Employment & 415,174 & 61 & 21 & 46 & 60 & 74 \\ 
+Halifax & Employment & 279,178 & 66 & 25 & 47 & 65 & 84 \\ 
+All regions & Groceries Stores & 18,519,897 & 76 & 26 & 56 & 76 & 97 \\ 
+Toronto & Groceries Stores & 8,512,874 & 80 & 25 & 61 & 82 & 101 \\ 
+Montréal & Groceries Stores & 2,993,965 & 73 & 26 & 53 & 74 & 94 \\ 
+Vancouver & Groceries Stores & 4,540,106 & 73 & 25 & 53 & 74 & 93 \\ 
+Calgary & Groceries Stores & 502,757 & 70 & 22 & 54 & 70 & 86 \\ 
+Ottawa & Groceries Stores & 619,791 & 72 & 24 & 55 & 72 & 91 \\ 
+Edmonton & Groceries Stores & 324,030 & 68 & 23 & 51 & 67 & 84 \\ 
+Québec City & Groceries Stores & 234,600 & 70 & 26 & 51 & 70 & 89 \\ 
+Winnipeg & Groceries Stores & 362,566 & 58 & 21 & 43 & 57 & 72 \\ 
+Hamilton & Groceries Stores & 281,118 & 84 & 29 & 60 & 91 & 110 \\ 
+Waterloo & Groceries Stores & 44,726 & 66 & 28 & 44 & 62 & 90 \\ 
+London & Groceries Stores & 52,617 & 55 & 19 & 42 & 55 & 67 \\ 
+Halifax & Groceries Stores & 50,747 & 63 & 26 & 42 & 63 & 83 \\ 
+\bottomrule
+\end{longtable}
+\endgroup
 

### 2025-03-11 paper/descriptive_stats_paper.R

@@ -1,99 +1,0 @@
-
-# library -----------------------------------------------------------------
-
-
-library(tidyverse)
-library(canaccessR)
-library(sf)
-library(data.table)
-
-
-# read data ---------------------------------------------------------------
-
-
-# c_regions <- c(
-#   "cal"
-#   , "edm"
-#   # , "ggh"
-#   # , "hal"
-#   # , "ldn"
-#   # , "mtl"
-#   # , "ott"
-#   # , "que"
-#   # , "van"
-#   # , "win"
-# )
-# 
-# c_dest_types <- c("emp", "grc")
-# 
-# 
-# 
-# vec_matrices <- map2_chr(
-#   .x = c_dest_types
-#   , .y = c_regions
-#   , ~paste0("travel_matrix_", .x, "_", .y)
-# )
-# 
-# l_teste <- purrr::map(
-#   .x = vec_matrices
-#   , ~data(.x)
-# )
-
-f_ttm <- function(){
-  # datasets from canaccessR package
-  d <- data(package = "canaccessR")
-  # get travel matrices names
-  vec_matrices <- d$results[, "Item"] %>% str_subset("travel_matrix")
-  
-  # call promised data
-  # https://stackoverflow.com/a/27710355
-  data(list = vec_matrices, package = "canaccessR")
-  
-  # separate ggh matrices (if necessary)
-  c_names <- travel_matrix_emp_ggh %>% 
-    pull(study_region_name) %>% 
-    unique() %>% 
-    droplevels.factor() %>% 
-    as.character()
-  
-  df_region <- c("tor", "ham", "wat")
-  
-  c_dest_types <- c("emp", "grc")
-  
-  # purrr into list
-  l_emp_ggh <- purrr::map(
-    .x = c_names
-    , ~dplyr::filter(.data = travel_matrix_emp_ggh, study_region_name == .x)
-  )
-  # rename list elements
-  names(l_emp_ggh) <- paste0("travel_matrix_emp_", df_region)
-  # return list elements into global environment
-  list2env(l_emp_ggh, globalenv())
-  
-  # purrr into list
-  l_grc_ggh <- purrr::map(
-    .x = c_names
-    , ~dplyr::filter(.data = travel_matrix_grc_ggh, study_region_name == .x)
-  )
-  # rename list elements
-  names(l_grc_ggh) <- paste0("travel_matrix_grc_", df_region)
-  # return list elements into global environment
[…]

### 2025-03-11 paper/paper_tables_figures.R

@@ -1,0 +1,255 @@
+
+# library -----------------------------------------------------------------
+
+
+library(tidyverse)
+library(canaccessR)
+library(sf)
+library(data.table)
+library(summarytools)
+library(skimr)
+
+
+# read data ---------------------------------------------------------------
+
+
+f_ttm <- function(){
+  # datasets from canaccessR package
+  d <- data(package = "canaccessR")
+  # get travel matrices names
+  vec_matrices <- d$results[, "Item"] %>% str_subset("travel_matrix")
+  
+  # call promised data
+  # https://stackoverflow.com/a/27710355
+  data(list = vec_matrices, package = "canaccessR")
+  
+  df_emp <- data.table::rbindlist(mget(ls(pattern = "travel_matrix_emp")))
+  
+  df_grc <- data.table::rbindlist(mget(ls(pattern = "travel_matrix_grc")))
+  
+  # # separate ggh matrices (if necessary)
+  # c_names <- travel_matrix_emp_ggh %>% 
+  #   pull(study_region_name) %>% 
+  #   unique() %>% 
+  #   droplevels.factor() %>% 
+  #   as.character()
+  # 
+  # df_region <- c("tor", "ham", "wat")
+  # 
+  # # c_dest_types <- c("emp", "grc")
+  # 
+  # # purrr into list
+  # l_emp_ggh <- purrr::map(
+  #   .x = c_names
+  #   , ~dplyr::filter(.data = travel_matrix_emp_ggh, study_region_name == .x)
+  # )
+  # # rename list elements
+  # names(l_emp_ggh) <- paste0("travel_matrix_emp_", df_region)
+  # # return list elements into global environment
+  # list2env(l_emp_ggh, globalenv())
+  # 
+  # # purrr into list
+  # l_grc_ggh <- purrr::map(
+  #   .x = c_names
+  #   , ~dplyr::filter(.data = travel_matrix_grc_ggh, study_region_name == .x)
+  # )
+  # # rename list elements
+  # names(l_grc_ggh) <- paste0("travel_matrix_grc_", df_region)
+  # # return list elements into global environment
+  # list2env(l_grc_ggh, globalenv())
+  # 
+  # rm(l_emp_ggh, l_grc_ggh)
+  
+  # combine data into list
+  # https://stackoverflow.com/a/61699751
+  # ttm_list <- mget(ls(pattern = "travel"))
+  
[…]

### 2025-03-11 paper/Manuscript-Data-Package.Rmd
create plot and finish draft
Create travel time map figure; include figure and brief description at the first version of the draft; finish first version of the draft.

@@
     keep_tex: yes
+    fig_caption: yes
+header-includes: \usepackage{float}
 editor_options: 
@@
 The travel time datasets contain, in total, 97,784,850 origin-destination pairs (observations) from population to employment locations for all the DA's in the sample and 18,519,897 pairs from population to groceries stores.
-Considering all areas combined, the travel times estimated to jobs have an average of 78, standard deviation of 25, 1st quantile of 59, median of 80, and 3rd quantile of 99.
-For travel times to groceries stores, these same statistics are the 76, 26, 56, 76, and 97, respectively.
+Considering all areas combined, the mean travel time to jobs was 80 and 76 for groceries stores.
 
-```{r}
+```{r, fig.pos="H"}
 #| label: tbl-table_1
@@
 This section exemplifies how the package can be used through visual representation.
+In Figure @fig-travel_time_emp_grc_plot, we present the spatial representation of the travel time matrices for the metropolitan region of Montréal.
+In it, we see the median travel time from each DA to employment (left) and to groceries stores (right).
+The plot shows that moving away from the city core increases the necessary travel time by public transit to reach employment locations and groceries stores.
+
+```{r, fig.pos="H"}
+#| label: fig-travel_time_emp_grc_plot
+#| fig-cap: "Estimated median travel time (minutes) per Dissemination Area to jobs (left) and groceries stores (right).
+Public transit travel times are calculated using {r5r} (Pereira et al., 2021).
+DA and planning boundaries of the Montréal metropolitan area (Statistics Canada, 2016)."
+#| out-width: 100%
+#| echo: false
+knitr::include_graphics("../figures/patch_tt_emp_grc.jpg")
+```
 

### 2025-03-11 paper/Manuscript-Data-Package.tex
create plot and finish draft
Create travel time map figure; include figure and brief description at the first version of the draft; finish first version of the draft.

@@
 
+\usepackage{float}
 \usepackage{booktabs}
@@
 and 18,519,897 pairs from population to groceries stores. Considering
-all areas combined, the travel times estimated to jobs have an average
-of 78, standard deviation of 25, 1st quantile of 59, median of 80, and
-3rd quantile of 99. For travel times to groceries stores, these same
-statistics are the 76, 26, 56, 76, and 97, respectively.
+all areas combined, the mean travel time to jobs was 80 and 76 for
+groceries stores.
 
@@
 
-This section presents some potential applications of the data package.
+This section exemplifies how the package can be used through visual
+representation. In Figure \citet{fig-travel_time_emp_grc_plot}, we
+present the spatial representation of the travel time matrices for the
+metropolitan region of Montréal. In it, we see the median travel time
+from each DA to employment (left) and to groceries stores (right). The
+plot shows that moving away from the city core increases the necessary
+travel time by public transit to reach employment locations and
+groceries stores.
+
+\begin{figure}[H]
+\includegraphics[width=1\linewidth]{../figures/patch_tt_emp_grc} \caption{Estimated median travel time (minutes) per Dissemination Area to jobs (left) and groceries stores (right).
+Public transit travel times are calculated using {r5r} (Pereira et al., 2021).
+DA and planning boundaries of the Montréal metropolitan area (Statistics Canada, 2016).}\label{fig:fig-travel_time_emp_grc_plot}
+\end{figure}
+
+A more thorough example of the package's use can be found on the School
+of Cities' recent report on Canada's Urban Infrastructure Deficit. In
+its 11th Chapter, we use the travel time matrices to estimate
+accessibility metrics to jobs and groceries stores before and after the
+pandemic \citep{pargaDemocraticAccessOur2024}. We then compare how
+changes affected groups differently according to their spatial
+distribution and income level, thus making explicit the connection of
+the package's information and matters of equity in transportation. The
+report is freely available for download at the State of Cities Summit
+\href{https://stateofcitiessummit.ca/report}{website}.
 

### 2025-03-11 paper/paper_tables_figures.R

@@ -8,6 +8,10 @@ library(sf)
 library(data.table)
 library(summarytools)
 library(skimr)
+library(stringi)
+library(hrbrthemes)
+library(tmap)
+library(patchwork)
 
 
 # read data ---------------------------------------------------------------
@@ -65,7 +69,7 @@ f_ttm <- function(){
   # ttm_list <- mget(ls(pattern = "travel"))
   
   rm(list = ls(pattern = "travel"))
-  rm(d)
+  # rm(d)
   gc()
 }
 
@@ -176,6 +180,79 @@ f_skim <- function(){
 
 
 
+# area and mean tt --------------------------------------------------------
+
+vec_cma <- d$results[, "Item"] %>% str_subset("census_data_cma")
+
+# call promised data
+# https://stackoverflow.com/a/27710355
+data(list = vec_cma, package = "canaccessR")
+
+# union vancouver
+census_data_cma_van <- census_data_cma_van %>% 
+  group_by(study_region_name) %>% 
+  summarise() %>% 
+  sf::st_union()
+
+# separate ggh
+data("census_data_da_ggh")
+
+census_data_cma_tor <- census_data_da_ggh %>% 
+  select(CMA_UID, geometry) %>% 
+  filter(CMA_UID %in% c("35535","35532")) %>% 
+  sf::st_union()
+
+census_data_cma_ham <- census_data_da_ggh %>% 
+  select(CMA_UID, geometry) %>% 
+  filter(CMA_UID %in% c("35537")) %>% 
+  sf::st_union()
+
+census_data_cma_wat <- census_data_da_ggh %>% 
+  select(CMA_UID, geometry) %>% 
+  filter(CMA_UID %in% c("35541")) %>% 
+  sf::st_union()
+
+
+rm(census_data_da_ggh)
+
+l_cma <- list(mget(ls(pattern = "census_data_cma"))) %>% 
+  unlist(recursive=FALSE)
+l_cma <- within(l_cma, rm(census_data_cma_ggh))
+rm(list = ls(pattern = "census_data_cma"))
+
+# compute areas
+vec_area <- map_vec(
+  .x = l_cma
+  , ~ sf::st_area(.x)
+)
+
+df_area <- data.table(
+  name = str_extract(string = names(vec_area), pattern = "(?<=census_data_cma_).*")
+  , vec_area
+)
+
+
+vec_lower <- sk_full %>% 
[…]

### 2025-03-18 paper/Manuscript-Data-Package.Rmd
AS edits to the draft

@@
 The objective of this paper is to describe the {canaccesR} open data package.
-Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023.
-The results are stored at the Dissemination Areas (DA) [^1] level, yielding origin-destination pairs containing information on travel time by public transit, population, and total employment within each metropolitan area.
-These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} package [@pereiraR5rRapidRealistic2021].
-We used public transit schedule, transport network, population, and business location data from different sources to estimate these TTM's datasets.
+Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023, representing approximately 55% of the Canadian population.
+The results are provided at the Dissemination Areas (DA) [^1] level, yielding origin-destination pairs containing associated travel time by public transit, population, total employment, mode share, and other relevant census variables and spatial shape boundaries for each metropolitan area.
+This data package was created by leveraging expertise in data science, R programming, and transportation analysis.
+It includes .Rmd notebooks written for computing travel time matrices for large sets of origin-destination pairs using the {r5r} R package [@pereiraR5rRapidRealistic2021].
+Overall, {canaccesR} offers `r data(package = "canaccessR")$results[, "Item"] |> length()` complementary objects ready for temporal and spatial analysis.
+The package is an analysis-read product based on a fusion of data sources, including public transit schedules (GTFS files), road and transit networks, census data, and filtered business location data.
 
@@
 
-Estimating accessibility - the potential offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012] - requires specialized datasets and technical expertise.
+TTMs are a core piece of information required for estimating spatial interaction.
+Accessibility -the _potential_ spatial interaction offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012]- is an example of a measure that requires TTMs.
 Recent efforts following an open-source and transparent philosophy have been made to disseminate useful data and information on transportation in the Canadian context [@soukhovTTS2016RDataSet2023].
@@
 
-The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2016 Canadian Census [@governmentofcanada2016CensusPopulation2016] [^2].
-These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
+The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2016 Canadian Census [@governmentofcanada2016CensusPopulation2016] [^2].These locations are the surrunding Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax areas.
 We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
 
-[^2]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter.
-We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.
+[^2]: We included Oshawa as part of the Greater Toronto Area (GTA) because of its proximity.
+Similarly, we included Abbotsford-Mission as part of the Vancouver metropolitan area due to its proximity to a transit station on the region’s West Coast Express commuter rail line.
 
@@
 
-Using the {r5r} package, we estimated public transit travel times for two destination types, grocery stores and jobs. 
+Using the {r5r} R package, we estimated public transit travel times for two destination types, grocery stores and jobs.
 For each amenity type, we chose a likely travel time and day of the week.
@@
 <!-- ## Travel time matrices -->
+```{r}
+data(package = "canaccessR")$results[, "Item"]
+grep(paste("^", "census_data_da_", sep=""), data(package = "canaccessR")$results[, "Item"])
+ data("census_data_cma_cal")
+```
 
-The package contains the following contents: travel time matrices, socio-economic and demographic data disaggregated at the DA level, the CMA areas' boundaries and backgrounds for plotting the data spatially, and aggregated statistics.
+Specifically, the package contains the following contents: `r grep(paste("^", "travel_matrix_grc_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` data.frame objects containing the calculated public travel times from DA centroids to grocery stores and another `r grep(paste("^", "travel_matrix_emp_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` for DA centroids to DA centroids.
+Notably, the {travel_matrix_grc_ggh} data.frame, is named after the acronym for the Greater Golden Horseshoe area, which includes the CMA regions of Toronto, Hamilton, and Kitchener-Cambridge-Waterloo.
+Hence, the `r grep(paste("^", "travel_matrix_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` travel time data.frames represent public transit travel times for all 12 CMAs across two sets of destinations.
+Next, `r grep(paste("^", "census_data_da_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` sf objects represent census data for each DA, including dwelling counts, population by age bracket, single-parent-headed households, low-income prevalence, official language knowledge, housing- quality, ownership and affordability variables, visible minority, newcomer- and immigration- related variables, educational attainment, and commuting mode shares.
+Furthermore, `r grep(paste("^", "census_data_cma_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` and `r grep(paste("^", "region_background_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` sf objects represent the CMA areas' boundaries and backgrounds for plotting the data spatially, respectively.
+Finally, `r grep("statistics", data(package = "canaccessR")$results[, "Item"]) |> length()` data.frames contain aggregated population and transit statistics at the CMA level.
 
-The main contents of the {canaccessR} package are the travel time matrices.
-<!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  -->
-These matrices (datasets) comprise the estimated travel time by transit from and to each origin and destination Dissemination Area pairs.
+<!--AS: Joao, I rephrased the below into one paragraph, what do you think?  -->
+
+<!-- The main contents of the {canaccessR} package are the travel time matrices. -->
+<!-- <!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  --> -->
+<!-- These matrices (datasets) comprise the estimated travel time by transit from and to each origin and destination Dissemination Area pairs.
 The datasets also feature the total population in the origin DA and the total employment in the destination DA.
@@
 These information can be turned into spatial data using the socio-economic and demographic datasets.
+-->
 
-[^4]: The information from Toronto, Hamilton, and Waterloo are aggregated at the Greater Golden Horseshoe travel time matrices, making it 20 datasets in total.
+<!-- [^4]: The information from Toronto, Hamilton, and Waterloo are aggregated at the Greater Golden Horseshoe travel time matrices, making it 20 datasets in total.  -->
 
-The socio-economic and demographic data contain other information disaggregated at the DA level for each CMA. 
-These refer to total population by age groups, number of dwellings, number of individuals below the Low Income Measure, etc.
-The aggregated statistics refer to population aggregates (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale.
+<!-- The socio-economic and demographic data contain other information disaggregated at the DA level for each CMA.  -->
+<!-- These refer to total population by age groups, number of dwellings, number of individuals below the Low Income Measure, etc. -->
+<!-- The aggregated statistics refer to population aggregates (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale. -->
 
 We now present some descriptive statistics from the travel time matrices contained in the {canaccessR} package.
-The Table @tbl-table_1 below summarizes the travel time estimates.
-The travel time datasets contain, in total, 97,784,850 origin-destination pairs (observations) from population to employment locations for all the DA's in the sample and 18,519,897 pairs from population to groceries stores.
+Table @tbl-table_1 summarizes the travel time estimates.
+The travel time objects contain, in total, 97,784,850 origin-destination pairs (observations) from population to employment locations for all the DA's in the sample and 18,519,897 pairs from population to groceries stores.
 Considering all areas combined, the mean travel time to jobs was 80 and 76 for groceries stores.
@@
 ```
+<!--AS: Where is the 0-30 bin for grocery stores? It would be helpful to change the plot so it reads '0-30' '30-60', '60-90', 90-120min -->
 
@@
 
-The author(s) declared no potential conflicts of interest with respect
-to the research, authorship, and/or publication of this article.
+The author(s) declared no potential conflicts of interest with respect to the research, authorship, and/or publication of this article.
 
@@
 
-The author(s) disclosed receipt of the following financial support for
-the research, authorship, and/or publication of this article: This work
-was supported by the Social Sciences and Humanities Research Council of
-Canada (*More description about the funding source after the review
-process*).
+The author(s) disclosed receipt of the following financial support for the research, authorship, and/or publication of this article: This work was supported by the Social Sciences and Humanities Research Council of Canada (*More description about the funding source after the review process*).
 

### 2025-03-18 paper/Manuscript-Data-Package.tex
AS edits to the draft

@@
 matrices (TTM) estimates to employment and groceries stores from the 12
-largest Canadian metropolitan areas in 2019 and 2023. The results are
-stored at the Dissemination Areas (DA) \footnote{Dissemination Areas are
-  the smallest publicly available spatial unit provided by Statistics
-  Canada \citep{governmentofcanadaDictionaryCensusPopulation2021a}.}
-level, yielding origin-destination pairs containing information on
-travel time by public transit, population, and total employment within
-each metropolitan area. These estimates were created by leveraging
-expertise in data science, computer programming, and transportation
-accessibility using the \{r5r\} package
-\citep{pereiraR5rRapidRealistic2021}. We used public transit schedule,
-transport network, population, and business location data from different
-sources to estimate these TTM's datasets.
+largest Canadian metropolitan areas in 2019 and 2023, representing
+approximately 55\% of the Canadian population. The results are provided
+at the Dissemination Areas (DA) \footnote{Dissemination Areas are the
+  smallest publicly available spatial unit provided by Statistics Canada
+  \citep{governmentofcanadaDictionaryCensusPopulation2021a}.} level,
+yielding origin-destination pairs containing associated travel time by
+public transit, population, total employment, mode share, and other
+relevant census variables and spatial shape boundaries for each
+metropolitan area. This data package was created by leveraging expertise
+in data science, R programming, and transportation analysis. It includes
+.Rmd notebooks written for computing travel time matrices for large sets
+of origin-destination pairs using the \{r5r\} R package
+\citep{pereiraR5rRapidRealistic2021}. Overall, \{canaccesR\} offers 52
+complementary objects ready for temporal and spatial analysis. The
+package is an analysis-read product based on a fusion of data sources,
+including public transit schedules (GTFS files), road and transit
+networks, census data, and filtered business location data.
 
-Estimating accessibility - the potential offered by the transportation
-system to reach destinations
-\citep{paezMeasuringAccessibilityPositive2012} - requires specialized
-datasets and technical expertise. Recent efforts following an
-open-source and transparent philosophy have been made to disseminate
-useful data and information on transportation in the Canadian context
+TTMs are a core piece of information required for estimating spatial
+interaction. Accessibility -the \emph{potential} spatial interaction
+offered by the transportation system to reach destinations
+\citep{paezMeasuringAccessibilityPositive2012}- is an example of a
+measure that requires TTMs. Recent efforts following an open-source and
+transparent philosophy have been made to disseminate useful data and
+information on transportation in the Canadian context
 \citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
@@
 Canadian Census \citep{governmentofcanada2016CensusPopulation2016}
-\footnote{We included Oshawa, Ontario, as part of the Greater Toronto
-  Area (GTA) because of the former's proximity to the latter. We also
-  included Abbotsford-Mission, British Columbia, as part of the
-  Vancouver metropolitan area because of the former's proximity to a
-  transit station on the region's West Coast Express commuter rail line.}.
-These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau,
-Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
-Kitchener-Cambridge-Waterloo, London, and Halifax. We used four main
-data sources to construct the \{canaccesR\} data package: General
+\footnote{We included Oshawa as part of the Greater Toronto Area (GTA)
+  because of its proximity. Similarly, we included Abbotsford-Mission as
+  part of the Vancouver metropolitan area due to its proximity to a
+  transit station on the region's West Coast Express commuter rail line.}.These
+locations are the surrunding Toronto, Montreal, Vancouver,
+Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
+Kitchener-Cambridge-Waterloo, London, and Halifax areas. We used four
+main data sources to construct the \{canaccesR\} data package: General
 Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
@@
 
-Using the \{r5r\} package, we estimated public transit travel times for
-two destination types, grocery stores and jobs. For each amenity type,
-we chose a likely travel time and day of the week. We set a 15 minutes
-time window and the maximum trip duration to 120 minutes. The estimated
-times are the median of the 15 minute time window. For groceries stores,
-we set the departure date to a weekend afternoon and the departure time
-to between 12:00 PM to 12:15 PM on April 20, 2019 and April 22, 2023.
-For employment, we ran the analysis on a typical weekday morning
-rush-hour commute, more specifically 8:00 to 8:15 AM departure on
-Tuesday, April 16, 2019 and Tuesday, April 18, 2023 \footnote{The one
+Using the \{r5r\} R package, we estimated public transit travel times
+for two destination types, grocery stores and jobs. For each amenity
+type, we chose a likely travel time and day of the week. We set a 15
+minutes time window and the maximum trip duration to 120 minutes. The
+estimated times are the median of the 15 minute time window. For
+groceries stores, we set the departure date to a weekend afternoon and
+the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and
+April 22, 2023. For employment, we ran the analysis on a typical weekday
+morning rush-hour commute, more specifically 8:00 to 8:15 AM departure
+on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 \footnote{The one
   exception is Quebec City, where the routing for 2019 occurs on a
@@
 
-The package contains the following contents: travel time matrices,
-socio-economic and demographic data disaggregated at the DA level, the
-CMA areas' boundaries and backgrounds for plotting the data spatially,
-and aggregated statistics.
+\begin{verbatim}
+ [1] "census_data_cma_cal"   "census_data_cma_edm"   "census_data_cma_ggh"  
+ [4] "census_data_cma_hal"   "census_data_cma_ldn"   "census_data_cma_mtl"  
+ [7] "census_data_cma_ott"   "census_data_cma_que"   "census_data_cma_van"  
+[10] "census_data_cma_win"   "census_data_da_cal"    "census_data_da_edm"   
+[13] "census_data_da_ggh"    "census_data_da_hal"    "census_data_da_ldn"   
+[16] "census_data_da_mtl"    "census_data_da_ott"    "census_data_da_que"   
+[19] "census_data_da_van"    "census_data_da_win"    "population_statistics"
+[22] "region_background_cal" "region_background_edm" "region_background_ggh"
+[25] "region_background_hal" "region_background_ldn" "region_background_mtl"
+[28] "region_background_ott" "region_background_que" "region_background_van"
+[31] "region_background_win" "transit_statistics"    "travel_matrix_emp_cal"
+[34] "travel_matrix_emp_edm" "travel_matrix_emp_ggh" "travel_matrix_emp_hal"
+[37] "travel_matrix_emp_ldn" "travel_matrix_emp_mtl" "travel_matrix_emp_ott"
+[40] "travel_matrix_emp_que" "travel_matrix_emp_van" "travel_matrix_emp_win"
+[43] "travel_matrix_grc_cal" "travel_matrix_grc_edm" "travel_matrix_grc_ggh"
+[46] "travel_matrix_grc_hal" "travel_matrix_grc_ldn" "travel_matrix_grc_mtl"
+[49] "travel_matrix_grc_ott" "travel_matrix_grc_que" "travel_matrix_grc_van"
+[52] "travel_matrix_grc_win"
+\end{verbatim}
 
-The main contents of the \{canaccessR\} package are the travel time
-matrices. These matrices (datasets) comprise the estimated travel time
-by transit from and to each origin and destination Dissemination Area
-pairs. The datasets also feature the total population in the origin DA
-and the total employment in the destination DA. Besides these
-information, the matrices contain the unique origin and destination DA
-codes, region and name identifiers, the year in which the travel times
-refers to, and the date and time of departure for a trip that originates
-at the Dissemination Area of origin. There are two matrices for each
-metropolitan area, one containing the travel times to jobs and the other
-to groceries stores \footnote{The information from Toronto, Hamilton,
-  and Waterloo are aggregated at the Greater Golden Horseshoe travel
-  time matrices, making it 20 datasets in total.}. These information can
-be turned into spatial data using the socio-economic and demographic
-datasets.
+\begin{verbatim}
+ [1] 11 12 13 14 15 16 17 18 19 20
+\end{verbatim}
 
-The socio-economic and demographic data contain other information
-disaggregated at the DA level for each CMA. These refer to total
-population by age groups, number of dwellings, number of individuals
-below the Low Income Measure, etc. The aggregated statistics refer to
-population aggregates (for the selected CMA's) and transit revenue and
-ridership data aggregated by regional and national scale.
+Specifically, the package contains the following contents: 10 data.frame
+objects containing the calculated public travel times from DA centroids
+to grocery stores and another 10 for DA centroids to DA centroids.
+Notably, the \{travel\_matrix\_grc\_ggh\} data.frame, is named after the
+acronym for the Greater Golden Horseshoe area, which includes the CMA
+regions of Toronto, Hamilton, and Kitchener-Cambridge-Waterloo. Hence,
+the 20 travel time data.frames represent public transit travel times for
+all 12 CMAs across two sets of destinations. Next, 10 sf objects
+represent census data for each DA, including dwelling counts, population
+by age bracket, single-parent-headed households, low-income prevalence,
+official language knowledge, housing- quality, ownership and
+affordability variables, visible minority, newcomer- and immigration-
+related variables, educational attainment, and commuting mode shares.
+Furthermore, 10 and 10 sf objects represent the CMA areas' boundaries
+and backgrounds for plotting the data spatially, respectively. Finally,
+2 data.frames contain aggregated population and transit statistics at
+the CMA level.
+
+--\textgreater{}
 
 We now present some descriptive statistics from the travel time matrices
-contained in the \{canaccessR\} package. The Table \citet{tbl-table_1}
-below summarizes the travel time estimates. The travel time datasets
-contain, in total, 97,784,850 origin-destination pairs (observations)
-from population to employment locations for all the DA's in the sample
-and 18,519,897 pairs from population to groceries stores. Considering
-all areas combined, the mean travel time to jobs was 80 and 76 for
-groceries stores.
+contained in the \{canaccessR\} package. Table \citet{tbl-table_1}
+summarizes the travel time estimates. The travel time objects contain,
+in total, 97,784,850 origin-destination pairs (observations) from
+population to employment locations for all the DA's in the sample and
+18,519,897 pairs from population to groceries stores. Considering all
+areas combined, the mean travel time to jobs was 80 and 76 for groceries
+stores.
 
-\begingroup
-\setlength\LTleft{0\linewidth}
-\setlength\LTright{0\linewidth}\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{longtable}{@{\extracolsep{\fill}}clrrrrrr}
+\begin{table}[!t]
+\fontsize{7.5pt}{9.0pt}\selectfont
+\begin{tabular*}{1\linewidth}{@{\extracolsep{\fill}}clrrrrrr}
 \toprule
@@
 \bottomrule
-\end{longtable}
-\endgroup
+\end{tabular*}
+\end{table}
 

### 2025-03-18 paper/Manuscript-Data-Package.Rmd
Forgot to fix a small typo, AS edit to draft done.

@@
 We filtered the locations within the DMTI dataset using the grocery stores code from the North American Industry Classification System (NAICS) and the Standard Industrial Classification (SIC).
-{**THE CODE FOR THIS ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE transit_death_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A PROBLEM?**}
+<!-- {**THE CODE FOR THIS ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE transit_death_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A PROBLEM?**} -->
 
@@
 In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
-We aggregated all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
+We aggregated all the resulting travel time matrices at the DA level, which comprise the fundamental unit of analysis in data package.
 
@@
 <!-- ## Travel time matrices -->
-```{r}
-data(package = "canaccessR")$results[, "Item"]
-grep(paste("^", "census_data_da_", sep=""), data(package = "canaccessR")$results[, "Item"])
- data("census_data_cma_cal")
-```
-
 Specifically, the package contains the following contents: `r grep(paste("^", "travel_matrix_grc_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` data.frame objects containing the calculated public travel times from DA centroids to grocery stores and another `r grep(paste("^", "travel_matrix_emp_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` for DA centroids to DA centroids.
@@
 <!-- The main contents of the {canaccessR} package are the travel time matrices. -->
-<!-- <!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  --> -->
+<!-- <!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  --> 
 <!-- These matrices (datasets) comprise the estimated travel time by transit from and to each origin and destination Dissemination Area pairs.

### 2025-03-18 paper/Manuscript-Data-Package.tex
Forgot to fix a small typo, AS edit to draft done.

@@
 the North American Industry Classification System (NAICS) and the
-Standard Industrial Classification (SIC). \{\textbf{THE CODE FOR THIS
-ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE
-transit\_death\_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A
-PROBLEM?}\}
+Standard Industrial Classification (SIC).
 
@@
 destination. We aggregated all the resulting travel time matrices at the
-Dissemination Area (DA) level, which comprise the fundamental unit of
-analysis in data package.
+DA level, which comprise the fundamental unit of analysis in data
+package.
 
 \section{\{canaccessR\}'s contents}\label{canaccessrs-contents}
-
-\begin{verbatim}
- [1] "census_data_cma_cal"   "census_data_cma_edm"   "census_data_cma_ggh"  
- [4] "census_data_cma_hal"   "census_data_cma_ldn"   "census_data_cma_mtl"  
- [7] "census_data_cma_ott"   "census_data_cma_que"   "census_data_cma_van"  
-[10] "census_data_cma_win"   "census_data_da_cal"    "census_data_da_edm"   
-[13] "census_data_da_ggh"    "census_data_da_hal"    "census_data_da_ldn"   
-[16] "census_data_da_mtl"    "census_data_da_ott"    "census_data_da_que"   
-[19] "census_data_da_van"    "census_data_da_win"    "population_statistics"
-[22] "region_background_cal" "region_background_edm" "region_background_ggh"
-[25] "region_background_hal" "region_background_ldn" "region_background_mtl"
-[28] "region_background_ott" "region_background_que" "region_background_van"
-[31] "region_background_win" "transit_statistics"    "travel_matrix_emp_cal"
-[34] "travel_matrix_emp_edm" "travel_matrix_emp_ggh" "travel_matrix_emp_hal"
-[37] "travel_matrix_emp_ldn" "travel_matrix_emp_mtl" "travel_matrix_emp_ott"
-[40] "travel_matrix_emp_que" "travel_matrix_emp_van" "travel_matrix_emp_win"
-[43] "travel_matrix_grc_cal" "travel_matrix_grc_edm" "travel_matrix_grc_ggh"
-[46] "travel_matrix_grc_hal" "travel_matrix_grc_ldn" "travel_matrix_grc_mtl"
-[49] "travel_matrix_grc_ott" "travel_matrix_grc_que" "travel_matrix_grc_van"
-[52] "travel_matrix_grc_win"
-\end{verbatim}
-
-\begin{verbatim}
- [1] 11 12 13 14 15 16 17 18 19 20
-\end{verbatim}
 
@@
 the CMA level.
-
---\textgreater{}
 

### 2025-04-07 paper/Manuscript-Data-Package.Rmd
fixed spelling errors
package name (canaccessR); data-based to data-driven; OPDs to ODP

@@
 abstract: 
-  "In this paper, we describe the {canaccesR} package, an open data product (ODP) created in R that contains public transit travel time estimates to employment locations and grocery stores across Canada's 12 largest metropolitan areas.
+  "In this paper, we describe the {canaccessR} package, an open data product (ODP) created in R that contains public transit travel time estimates to employment locations and grocery stores across Canada's 12 largest metropolitan areas.
 We calculate travel time matrices (TTM) from and to each Dissemination Area (DA) within these regions for the years 2019 and 2023.
@@
 Moreover, these estimations can be used as inputs in equity and inequalities assessments, through the comparisons of within and across accessibility levels found throughout Canada's largest metropolitan areas.
-Consequently, we expect to contribute to informed and data-based decision making in transportation by disseminating these data.
+Consequently, we expect to contribute to informed and data-driven decision making in transportation by disseminating these data.
 We hope that these datasets can substantiate future improvements in policy-making that may lead to greater justice in the country's urban transportation systems.
 The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities).
-Finally, as an ODPs, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."
+Finally, as an ODP, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."
 
@@
 
-The objective of this paper is to describe the {canaccesR} open data package.
+The objective of this paper is to describe the {canaccessR} open data package.
 Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023, representing approximately 55% of the Canadian population.
@@
 It includes .Rmd notebooks written for computing travel time matrices for large sets of origin-destination pairs using the {r5r} R package [@pereiraR5rRapidRealistic2021].
-Overall, {canaccesR} offers `r data(package = "canaccessR")$results[, "Item"] |> length()` complementary objects ready for temporal and spatial analysis.
+Overall, {canaccessR} offers `r data(package = "canaccessR")$results[, "Item"] |> length()` complementary objects ready for temporal and spatial analysis.
 The package is an analysis-read product based on a fusion of data sources, including public transit schedules (GTFS files), road and transit networks, census data, and filtered business location data.
@@
 Within this context, we expect to help filling this gap by the processing raw into user-ready data and making them publicly available to advance knowledge on the field.
-Our main contribution is to provide analysis-ready data for Canada's largest cities on the topic of transportation accessibility, thus making urban analytics in the country more accessible and contributing to future research and data-based decision making.
+Our main contribution is to provide analysis-ready data for Canada's largest cities on the topic of transportation accessibility, thus making urban analytics in the country more accessible and contributing to future research and data-driven decision making.
 
@@
 First, the datasets allow for static assessment of the level of public transit accessibility across the country's largest cities before and after the COVID-19 pandemic.
-In other words, {canaccesR} makes it easier for those interested in comparing cities regarding their level of public transit accessibility to essential destinations (such as employment centers and groceries stores) to do so.
+In other words, {canaccessR} makes it easier for those interested in comparing cities regarding their level of public transit accessibility to essential destinations (such as employment centers and groceries stores) to do so.
 Second, the temporal and spatial characters of the datasets made available here allow researchers to evaluate accessibility changes through time and across space within the largest Canadian urban areas.
@@
 We present some basic descriptive statistics of these datasets, and elucidate how one can use them in accessibility analysis.
-Finally, we conclude by explaining how we expect {canaccesR} to contribute to the urban analytics and science community.
+Finally, we conclude by explaining how we expect {canaccessR} to contribute to the urban analytics and science community.
 
@@
 The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2016 Canadian Census [@governmentofcanada2016CensusPopulation2016] [^2].These locations are the surrunding Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax areas.
-We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
+We used four main data sources to construct the {canaccessR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
 
@@
 In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
-We aggregated all the resulting travel time matrices at the DA level, which comprise the fundamental unit of analysis in data package.
+We aggregated all the resulting travel time matrices at the DA level, which comprise the fundamental spatial unit of analysis in data package.
 
@@
 
-# How to use {canaccessR}
+# {canaccessR} usage
 
-This section exemplifies how the package can be used through visual representation.
-In Figure @fig-travel_time_emp_grc_plot, we present the spatial representation of the travel time matrices for the metropolitan region of Montréal.
+This section presents an usage example of the package through visual representation of its data.
+In Figure @fig-travel_time_emp_grc_plot, we show the spatial distribution of the travel time matrices for the metropolitan region of Montréal.
 In it, we see the median travel time from each DA to employment (left) and to groceries stores (right).
@@
 #| fig-cap: "Estimated median travel time (minutes) per Dissemination Area to jobs (left) and groceries stores (right).
-Public transit travel times are calculated using {r5r} (Pereira et al., 2021).
-DA and planning boundaries of the Montréal metropolitan area (Statistics Canada, 2016)."
+Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016)."
 #| out-width: 100%
@@
 
-In this paper, we describe the {canaccesR} data package, created using the {r5r} package and transit schedule, street network, employment, and population data.
+In this paper, we describe the {canaccessR} data package, created using the {r5r} package and transit schedule, street network, employment, and population data.
 The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.
@@
 
-<!-- In conclusion, we argue that the {canaccesR} package adds  -->
+<!-- In conclusion, we argue that the {canaccessR} package adds  -->
 
@@
 <!-- We expect the package to have two main contribu -->
-<!-- The main contribution from the {canaccesR} package is the  -->
+<!-- The main contribution from the {canaccessR} package is the  -->
 

### 2025-04-07 paper/Manuscript-Data-Package.tex
fixed spelling errors
package name (canaccessR); data-based to data-driven; OPDs to ODP

@@
 \begin{abstract}
-In this paper, we describe the \{canaccesR\} package, an open data
+In this paper, we describe the \{canaccessR\} package, an open data
 product (ODP) created in R that contains public transit travel time
@@
 levels found throughout Canada's largest metropolitan areas.
-Consequently, we expect to contribute to informed and data-based
+Consequently, we expect to contribute to informed and data-driven
 decision making in transportation by disseminating these data. We hope
@@
 may undergo expansions in the future by adding TTM's for other
-destinations (e.g., schools, healthcare facilities). Finally, as an
-ODPs, the \{canaccess\} package allows for open exploration, use, and
+destinations (e.g., schools, healthcare facilities). Finally, as an ODP,
+the \{canaccess\} package allows for open exploration, use, and
 contribution by users through its GitHub repository.
@@
 
-The objective of this paper is to describe the \{canaccesR\} open data
+The objective of this paper is to describe the \{canaccessR\} open data
 package. Its main contents are a set of public transit travel time
@@
 of origin-destination pairs using the \{r5r\} R package
-\citep{pereiraR5rRapidRealistic2021}. Overall, \{canaccesR\} offers 52
+\citep{pereiraR5rRapidRealistic2021}. Overall, \{canaccessR\} offers 52
 complementary objects ready for temporal and spatial analysis. The
@@
 thus making urban analytics in the country more accessible and
-contributing to future research and data-based decision making.
+contributing to future research and data-driven decision making.
 
@@
 public transit accessibility across the country's largest cities before
-and after the COVID-19 pandemic. In other words, \{canaccesR\} makes it
+and after the COVID-19 pandemic. In other words, \{canaccessR\} makes it
 easier for those interested in comparing cities regarding their level of
@@
 elucidate how one can use them in accessibility analysis. Finally, we
-conclude by explaining how we expect \{canaccesR\} to contribute to the
+conclude by explaining how we expect \{canaccessR\} to contribute to the
 urban analytics and science community.
@@
 Kitchener-Cambridge-Waterloo, London, and Halifax areas. We used four
-main data sources to construct the \{canaccesR\} data package: General
+main data sources to construct the \{canaccessR\} data package: General
 Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
@@
 destination. We aggregated all the resulting travel time matrices at the
-DA level, which comprise the fundamental unit of analysis in data
-package.
+DA level, which comprise the fundamental spatial unit of analysis in
+data package.
 
@@
 
-\begin{table}[!t]
-\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{tabular*}{1\linewidth}{@{\extracolsep{\fill}}clrrrrrr}
+\begingroup
+\setlength\LTleft{0\linewidth}
+\setlength\LTright{0\linewidth}\fontsize{7.5pt}{9.0pt}\selectfont
+\begin{longtable}{@{\extracolsep{\fill}}clrrrrrr}
 \toprule
@@
 \bottomrule
-\end{tabular*}
-\end{table}
+\end{longtable}
+\endgroup
 
-\section{How to use \{canaccessR\}}\label{how-to-use-canaccessr}
+\section{\{canaccessR\} usage}\label{canaccessr-usage}
 
-This section exemplifies how the package can be used through visual
-representation. In Figure \citet{fig-travel_time_emp_grc_plot}, we
-present the spatial representation of the travel time matrices for the
-metropolitan region of Montréal. In it, we see the median travel time
-from each DA to employment (left) and to groceries stores (right). The
-plot shows that moving away from the city core increases the necessary
-travel time by public transit to reach employment locations and
-groceries stores.
+This section presents an usage example of the package through visual
+representation of its data. In Figure
+\citet{fig-travel_time_emp_grc_plot}, we show the spatial distribution
+of the travel time matrices for the metropolitan region of Montréal. In
+it, we see the median travel time from each DA to employment (left) and
+to groceries stores (right). The plot shows that moving away from the
+city core increases the necessary travel time by public transit to reach
+employment locations and groceries stores.
 
@@
 \includegraphics[width=1\linewidth]{../figures/patch_tt_emp_grc} \caption{Estimated median travel time (minutes) per Dissemination Area to jobs (left) and groceries stores (right).
-Public transit travel times are calculated using {r5r} (Pereira et al., 2021).
-DA and planning boundaries of the Montréal metropolitan area (Statistics Canada, 2016).}\label{fig:fig-travel_time_emp_grc_plot}
+Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016).}\label{fig:fig-travel_time_emp_grc_plot}
 \end{figure}
@@
 
-In this paper, we describe the \{canaccesR\} data package, created using
-the \{r5r\} package and transit schedule, street network, employment,
-and population data. The package's main contents refers to the
-ready-to-use travel time matrices for public transit to reach employment
-and groceries stores in Canada's 12 largest urban areas. We expect the
-contents of the package to be used in transportation accessibility
-evaluations within and across those regions. Moreover, these datasets
-can be used in further equity assessments that evaluate the distribution
-of accessibility across space and between social groups. Furthermore, in
-the spirit of open data products
+In this paper, we describe the \{canaccessR\} data package, created
+using the \{r5r\} package and transit schedule, street network,
+employment, and population data. The package's main contents refers to
+the ready-to-use travel time matrices for public transit to reach
+employment and groceries stores in Canada's 12 largest urban areas. We
+expect the contents of the package to be used in transportation
+accessibility evaluations within and across those regions. Moreover,
+these datasets can be used in further equity assessments that evaluate
+the distribution of accessibility across space and between social
+groups. Furthermore, in the spirit of open data products
 \citep{arribas-belOpenDataProductsA2021}, the package can be expanded

### 2025-04-07 paper/Manuscript-Data-Package.Rmd
minor updates
italics on data.frames and sfs

@@
 output:
+  word_document: default
   rticles::sage_article:
@@
 <!-- ## Travel time matrices -->
-Specifically, the package contains the following contents: `r grep(paste("^", "travel_matrix_grc_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` data.frame objects containing the calculated public travel times from DA centroids to grocery stores and another `r grep(paste("^", "travel_matrix_emp_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` for DA centroids to DA centroids.
-Notably, the {travel_matrix_grc_ggh} data.frame, is named after the acronym for the Greater Golden Horseshoe area, which includes the CMA regions of Toronto, Hamilton, and Kitchener-Cambridge-Waterloo.
-Hence, the `r grep(paste("^", "travel_matrix_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` travel time data.frames represent public transit travel times for all 12 CMAs across two sets of destinations.
-Next, `r grep(paste("^", "census_data_da_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` sf objects represent census data for each DA, including dwelling counts, population by age bracket, single-parent-headed households, low-income prevalence, official language knowledge, housing- quality, ownership and affordability variables, visible minority, newcomer- and immigration- related variables, educational attainment, and commuting mode shares.
-Furthermore, `r grep(paste("^", "census_data_cma_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` and `r grep(paste("^", "region_background_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` sf objects represent the CMA areas' boundaries and backgrounds for plotting the data spatially, respectively.
-Finally, `r grep("statistics", data(package = "canaccessR")$results[, "Item"]) |> length()` data.frames contain aggregated population and transit statistics at the CMA level.
+Specifically, the package contains the following contents: `r grep(paste("^", "travel_matrix_grc_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` *data.frame* objects containing the calculated public travel times from DA centroids to grocery stores and another `r grep(paste("^", "travel_matrix_emp_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` for DA centroids to DA centroids.
+Notably, the {travel_matrix_grc_ggh} *data.frame* is named after the acronym for the Greater Golden Horseshoe area, which includes the CMA regions of Toronto, Hamilton, and Kitchener-Cambridge-Waterloo.
+The `r grep(paste("^", "travel_matrix_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` travel time *data.frame*s represent public transit travel times for all 12 CMAs across two sets of destinations.
+Next, `r grep(paste("^", "census_data_da_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` *sf* objects represent census data for each DA, including dwelling counts, population by age bracket, single-parent-headed households, low-income prevalence, official language knowledge, housing- quality, ownership and affordability variables, visible minority, newcomer- and immigration- related variables, educational attainment, and commuting mode shares.
+Furthermore, `r grep(paste("^", "census_data_cma_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` and `r grep(paste("^", "region_background_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` *sf* objects represent the CMA areas' boundaries and backgrounds for plotting the data spatially, respectively.
+Finally, `r grep("statistics", data(package = "canaccessR")$results[, "Item"]) |> length()` *data.frame*s contain aggregated population and transit statistics at the CMA level.
 
@@
 The travel time objects contain, in total, 97,784,850 origin-destination pairs (observations) from population to employment locations for all the DA's in the sample and 18,519,897 pairs from population to groceries stores.
-Considering all areas combined, the mean travel time to jobs was 80 and 76 for groceries stores.
+Considering all areas combined, the mean travel time to jobs was 78 and 76 to groceries stores.
 
@@
 #| label: tbl-table_1
+#| tbl-cap: "Descriptive statistics of the travel time matrices from the {canaccessR} package."
 #| echo: false

### 2025-04-23 paper/Manuscript-Data-Package.Rmd

@@
 Moreover, these estimations can be used as inputs in equity and inequalities assessments, through the comparisons of within and across accessibility levels found throughout Canada's largest metropolitan areas.
-Consequently, we expect to contribute to informed and data-driven decision making in transportation by disseminating these data.
+Consequently, we expect to contribute to informed and data-driven decision-making in transportation by disseminating these data.
 We hope that these datasets can substantiate future improvements in policy-making that may lead to greater justice in the country's urban transportation systems.
-The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities).
+The package is still in its initial phase and may undergo expansions in the future by adding TTMs for other destinations (e.g., schools, healthcare facilities).
 Finally, as an ODP, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."
@@
 The objective of this paper is to describe the {canaccessR} open data package.
-Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023, representing approximately 55% of the Canadian population.
+Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and grocery stores from the 12 largest Canadian metropolitan areas in 2019 and 2023, representing approximately 55% of the Canadian population.
 The results are provided at the Dissemination Areas (DA) [^1] level, yielding origin-destination pairs containing associated travel time by public transit, population, total employment, mode share, and other relevant census variables and spatial shape boundaries for each metropolitan area.
@@
 Recent efforts following an open-source and transparent philosophy have been made to disseminate useful data and information on transportation in the Canadian context [@soukhovTTS2016RDataSet2023].
-However, despite these initiatives, pre-processed and available data that allow for the ease estimation of accessibility indicators are still scarce.
-Within this context, we expect to help filling this gap by the processing raw into user-ready data and making them publicly available to advance knowledge on the field.
-Our main contribution is to provide analysis-ready data for Canada's largest cities on the topic of transportation accessibility, thus making urban analytics in the country more accessible and contributing to future research and data-driven decision making.
+However, despite these initiatives, readily available pre-processed data that facilitate the estimation of accessibility indicators remain scarce.
+Within this context, we aim to help fill this gap by processing raw data into user-ready formats and making them publicly available to advance knowledge in the field.
+Our main contribution is to provide analysis-ready transportation accessibility data for Canada's largest cities, thus making urban analytics in the country more accessible and contributing to future research and data-driven decision-making.
 
@@
 First, the datasets allow for static assessment of the level of public transit accessibility across the country's largest cities before and after the COVID-19 pandemic.
-In other words, {canaccessR} makes it easier for those interested in comparing cities regarding their level of public transit accessibility to essential destinations (such as employment centers and groceries stores) to do so.
+In other words, {canaccessR} makes it easier for those interested in comparing cities regarding their level of public transit accessibility to essential destinations (such as employment centers and grocery stores) to do so.
 Second, the temporal and spatial characters of the datasets made available here allow researchers to evaluate accessibility changes through time and across space within the largest Canadian urban areas.
 Third, as is now common practice in transportation accessibility research, used as inputs, these estimates can substantiate broader investigations on transportation justice and equity [@higginsChangesAccessibilityEmergency2021; @humbertoHowTranslateJustice2023; @pereiraGeographicAccessCOVID192021].
-For example, the TTM estimates allow for evaluating the evolution of public transit's accessibility by income or spatial distribution across all Dissemination Areas (DA's) of each of the 12 cities in the sample [@pargaDemocraticAccessOur2024].
+For example, the TTM estimates allow for evaluating the evolution of public transit's accessibility by income or spatial distribution across all Dissemination Areas (DAs) of each of the 12 cities in the sample [@pargaDemocraticAccessOur2024].
 In other words, the package's contents can be used from straightforward assessments of accessibility in Canadian urban areas to more theoretically and morally complex evaluations of justice in the country's urban transportation system.
 
-Besides this introduction, we organize this paper as follows.
-The next section contains a description of the data sources we used to construct the data package.
+Besides this introduction, we organize this paper as follows: The next section contains a description of the data sources we used to construct the data package.
 Then, we recount the data processing necessary to create the package.
 Next, we go through the main contents of the data package, i.e., the travel time matrices estimated through our analysis.
-We present some basic descriptive statistics of these datasets, and elucidate how one can use them in accessibility analysis.
+We present some basic descriptive statistics of these datasets and elucidate how one can use them in accessibility analysis.
 Finally, we conclude by explaining how we expect {canaccessR} to contribute to the urban analytics and science community.
@@
 
-The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2016 Canadian Census [@governmentofcanada2016CensusPopulation2016] [^2].These locations are the surrunding Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax areas.
+The locations included in the data package comprise the 12 largest (population-wise) census metropolitan areas (CMAs) based on the 2016 Canadian Census [@governmentofcanada2016CensusPopulation2016] [^2].
+These locations are the surrounding Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax areas.
 We used four main data sources to construct the {canaccessR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.
@@
 <!-- ## General Transit Feed Specification -->
-We manually collected and processed the GTFS files from all transit agencies within the selected CMA's to use their information on the public transit schedule in 2019 and 2023.
+We manually collected and processed the GTFS files from all transit agencies within the selected CMAs to use their information on the public transit schedule in 2019 and 2023.
 <!-- ## OpenStreetMaps -->
@@
 <!-- ## Census Data -->
-We collected data from the 2016 Canadian Census using the {cancensus} package [@vonbergmannCancensusPackageAccess2022] and used its information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016].
+We collected data from the 2016 Canadian Census using the {cancensus} package [@vonbergmannCancensusPackageAccess2022] and used its information on the spatial distribution of the population and the number of workplace locations (employment) across the CMAs [@governmentofcanada2016CensusPopulation2016].
 <!-- ## DMTI’s Enhanced Points of Interest -->
-Finally, we gathered and cleaned the 2023 DMTI’s Enhanced Points of Interest dataset to obtain the location of the groceries stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015].
+Finally, we gathered and cleaned the 2023 DMTI’s Enhanced Points of Interest dataset to obtain the location of the grocery stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015].
 We filtered the locations within the DMTI dataset using the grocery stores code from the North American Industry Classification System (NAICS) and the Standard Industrial Classification (SIC).
@@
 For each amenity type, we chose a likely travel time and day of the week.
-We set a 15 minutes time window and the maximum trip duration to 120 minutes.
+We set a 15-minute time window and the maximum trip duration to 120 minutes.
 The estimated times are the median of the 15 minute time window.
-For groceries stores, we set the departure date to a weekend afternoon and the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and April 22, 2023.
-For employment, we ran the analysis on a typical weekday morning rush-hour commute, more specifically 8:00 to 8:15 AM departure on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 [^3].
-In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
-We aggregated all the resulting travel time matrices at the DA level, which comprise the fundamental spatial unit of analysis in data package.
+For grocery stores, we set the departure date to a weekend afternoon and the departure time to between 12:00 PM to 12:15 PM on April 20, 2019, and April 22, 2023.
+For employment, we ran the analysis on a typical weekday morning rush-hour commute, more specifically 8:00 to 8:15 AM departure on Tuesday, April 16, 2019 and Tuesday, April 18, 2023.[^3] In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
+We aggregated all the resulting travel time matrices at the DA level, which comprise the fundamental spatial unit of analysis in the data package.
 
@@
 The `r grep(paste("^", "travel_matrix_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` travel time *data.frame*s represent public transit travel times for all 12 CMAs across two sets of destinations.
-Next, `r grep(paste("^", "census_data_da_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` *sf* objects represent census data for each DA, including dwelling counts, population by age bracket, single-parent-headed households, low-income prevalence, official language knowledge, housing- quality, ownership and affordability variables, visible minority, newcomer- and immigration- related variables, educational attainment, and commuting mode shares.
+Next, `r grep(paste("^", "census_data_da_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` *sf* objects represent census data for each DA, including dwelling counts, population by age bracket, single-parent-headed households, low-income prevalence, official language knowledge, housing quality, ownership and affordability variables, visible minorities, newcomer- and immigration-related variables, educational attainment, and commuting mode shares.
 Furthermore, `r grep(paste("^", "census_data_cma_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` and `r grep(paste("^", "region_background_", sep=""), data(package = "canaccessR")$results[, "Item"]) |> length()` *sf* objects represent the CMA areas' boundaries and backgrounds for plotting the data spatially, respectively.
@@
 Besides these information, the matrices contain the unique origin and destination DA codes, region and name identifiers, the year in which the travel times refers to, and the date and time of departure for a trip that originates at the Dissemination Area of origin.
-There are two matrices for each metropolitan area, one containing the travel times to jobs and the other to groceries stores [^4].
+There are two matrices for each metropolitan area, one containing the travel times to jobs and the other to grocery stores [^4].
 These information can be turned into spatial data using the socio-economic and demographic datasets.
@@
 <!-- These refer to total population by age groups, number of dwellings, number of individuals below the Low Income Measure, etc. -->
-<!-- The aggregated statistics refer to population aggregates (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale. -->
+<!-- The aggregated statistics refer to population aggregates (for the selected CMAs) and transit revenue and ridership data aggregated by regional and national scale. -->
 
@@
 Table @tbl-table_1 summarizes the travel time estimates.
-The travel time objects contain, in total, 97,784,850 origin-destination pairs (observations) from population to employment locations for all the DA's in the sample and 18,519,897 pairs from population to groceries stores.
-Considering all areas combined, the mean travel time to jobs was 78 and 76 to groceries stores.
+The travel time objects contain, in total, 97,784,850 origin-destination pairs (observations) from population to employment locations for all the DAs in the sample and 18,519,897 pairs from population to grocery stores.
+Considering all areas combined, the mean travel time to jobs was 78 and 76 to grocery stores.
 
@@
 
-This section presents an usage example of the package through visual representation of its data.
+This section presents a usage example of the package through a visual representation of its data.
 In Figure @fig-travel_time_emp_grc_plot, we show the spatial distribution of the travel time matrices for the metropolitan region of Montréal.
-In it, we see the median travel time from each DA to employment (left) and to groceries stores (right).
-The plot shows that moving away from the city core increases the necessary travel time by public transit to reach employment locations and groceries stores.
+In it, we see the median travel time from each DA to employment (left) and to grocery stores (right).
+The plot shows that moving away from the city core increases the necessary travel time by public transit to reach employment locations and grocery stores.
 
@@
 #| label: fig-travel_time_emp_grc_plot
-#| fig-cap: "Estimated median travel time (minutes) per Dissemination Area to jobs (left) and groceries stores (right).
+#| fig-cap: "Estimated median travel time (minutes) per Dissemination Area to jobs (left) and grocery stores (right).
 Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016)."
@@
 
-A more thorough example of the package's use can be found on the School of Cities' recent report on Canada's Urban Infrastructure Deficit.
-In its 11th Chapter, we use the travel time matrices to estimate accessibility metrics to jobs and groceries stores before and after the pandemic [@pargaDemocraticAccessOur2024].
-We then compare how changes affected groups differently according to their spatial distribution and income level, thus making explicit the connection of the package's information and matters of equity in transportation.
+A more thorough example of the package's use can be found in the School of Cities' recent report on Canada's Urban Infrastructure Deficit.
+In its 11th Chapter, we use the travel time matrices to estimate accessibility metrics to jobs and grocery stores before and after the pandemic [@pargaDemocraticAccessOur2024].
+We then compare how changes affected groups differently according to their spatial distribution and income level, thus making explicit the connection between the package's information and matters of equity in transportation.
 The report is freely available for download at the State of Cities Summit [website](https://stateofcitiessummit.ca/report).
@@
 In this paper, we describe the {canaccessR} data package, created using the {r5r} package and transit schedule, street network, employment, and population data.
-The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.
+The package's main contents refer to the ready-to-use travel time matrices for public transit to reach employment and grocery stores in Canada's 12 largest urban areas.
 We expect the contents of the package to be used in transportation accessibility evaluations within and across those regions.
@@
 Furthermore, in the spirit of open data products [@arribas-belOpenDataProductsA2021], the package can be expanded through collaboration with other researchers by, for example, including travel time matrices to other essential destinations within the DMTI's dataset (*e.g.*, schools, healthcare, etc.).
-In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [@martensTransportJusticeDesigning2016], and the inherent connection between public transit and the "right to the city" [@cogginRightTransportMoving2015].
+In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [@martensTransportJusticeDesigning2016] and the inherent connection between public transit and the "right to the city" [@cogginRightTransportMoving2015].
 
@@
 
-The {canaccessR} data package can be found and installed on its Github [respository](https://github.com/paezha/canaccessR).
+The {canaccessR} data package can be found and installed on its Github [repository](https://github.com/paezha/canaccessR).
 

### 2025-04-23 paper/Manuscript-Data-Package.tex

@@
 
-Specifically, the package contains the following contents: 10 data.frame
-objects containing the calculated public travel times from DA centroids
-to grocery stores and another 10 for DA centroids to DA centroids.
-Notably, the \{travel\_matrix\_grc\_ggh\} data.frame, is named after the
-acronym for the Greater Golden Horseshoe area, which includes the CMA
-regions of Toronto, Hamilton, and Kitchener-Cambridge-Waterloo. Hence,
-the 20 travel time data.frames represent public transit travel times for
-all 12 CMAs across two sets of destinations. Next, 10 sf objects
-represent census data for each DA, including dwelling counts, population
-by age bracket, single-parent-headed households, low-income prevalence,
-official language knowledge, housing- quality, ownership and
-affordability variables, visible minority, newcomer- and immigration-
-related variables, educational attainment, and commuting mode shares.
-Furthermore, 10 and 10 sf objects represent the CMA areas' boundaries
-and backgrounds for plotting the data spatially, respectively. Finally,
-2 data.frames contain aggregated population and transit statistics at
-the CMA level.
+Specifically, the package contains the following contents: 10
+\emph{data.frame} objects containing the calculated public travel times
+from DA centroids to grocery stores and another 10 for DA centroids to
+DA centroids. Notably, the \{travel\_matrix\_grc\_ggh\}
+\emph{data.frame} is named after the acronym for the Greater Golden
+Horseshoe area, which includes the CMA regions of Toronto, Hamilton, and
+Kitchener-Cambridge-Waterloo. The 20 travel time \emph{data.frame}s
+represent public transit travel times for all 12 CMAs across two sets of
+destinations. Next, 10 \emph{sf} objects represent census data for each
+DA, including dwelling counts, population by age bracket,
+single-parent-headed households, low-income prevalence, official
+language knowledge, housing- quality, ownership and affordability
+variables, visible minority, newcomer- and immigration- related
+variables, educational attainment, and commuting mode shares.
+Furthermore, 10 and 10 \emph{sf} objects represent the CMA areas'
+boundaries and backgrounds for plotting the data spatially,
+respectively. Finally, 2 \emph{data.frame}s contain aggregated
+population and transit statistics at the CMA level.
 
@@
 18,519,897 pairs from population to groceries stores. Considering all
-areas combined, the mean travel time to jobs was 80 and 76 for groceries
+areas combined, the mean travel time to jobs was 78 and 76 to groceries
 stores.

### 2025-05-01 paper/Manuscript-Data-Package-cover.Rmd
Version of paper submitted to EPB

---
title: "canaccessR: An open data product for transit accessibility analysis in Canada’s largest metropolitan areas."
author:
  - name: João Pedro Figueira Amorim Parga
    email: joao.parga@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-4105-5927
    footnote: 1
  - name: Anastasia Soukhov
    email: soukhoa@mcmaster.ca
    affil-id: 2
    orcid: 0000-0003-4371-4831
  - name: Robert Nutifafa Arku
    email: rn.arku@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-2018-886X
  - name: Christopher Higgins
    email: cd.higgins@utoronto.ca
    affil-id: 3
    orcid: 0000-0002-3551-7750
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affil-id: 2
    orcid: 0000-0001-6912-9919
affiliations:
  - id: 1
    name: University of Toronto
  - id: 2
    name: McMaster University
  - id: 3
    name: University of Toronto Scarborough
# address:
#   - code: CODE
#     address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "In this paper, we describe the {canaccessR} package, an open data product (ODP) created in R that contains public transit travel time estimates to employment locations and grocery stores across Canada's 12 largest metropolitan areas. We calculate travel time matrices (TTM) from and to each Dissemination Area (DA) within these regions for the years 2019 and 2023. We add value to the urban analytics community by processing and integrating raw data, and disseminating user-ready data in the domain of transportation accessibility in Canada. To do so, we use the {r5r} R package, General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data. This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility levels to these two essential destinations within these urban areas. Moreover, travel time matrices are computed from DA centroid to DA centroid, which means that they can be adapted for use in applications with any type of destination that is aggregated at the DA level. Finally, as an ODP, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."

keywords: Public transit accessibility; open data products (ODPs); R data package; travel time matrices.
classoption:
  - Royal
  - times
bibliography: bibfile
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
    fig_caption: yes
header-includes: \usepackage{float}
editor_options: 
  markdown: 
    wrap: 72
---

### 2025-05-01 paper/Manuscript-Data-Package-cover.tex
Version of paper submitted to EPB

--- 2025-04-07 paper/Manuscript-Data-Package.tex
@@
 \usepackage{float}
-\usepackage{booktabs}
-\usepackage{caption}
-\usepackage{longtable}
-\usepackage{colortbl}
-\usepackage{array}
-\usepackage{anyfontsize}
-\usepackage{multirow}
 
@@
 
-\title{canaccessR: An open data product for analyzing transportation
-accessibility to employment and grocery stores in Canada's largest
-metropolitan areas.}
+\title{canaccessR: An open data product for transit accessibility
+analysis in Canada's largest metropolitan areas.}
 
@@
 these two essential destinations within these urban areas. Moreover,
-these estimations can be used as inputs in equity and inequalities
-assessments, through the comparisons of within and across accessibility
-levels found throughout Canada's largest metropolitan areas.
-Consequently, we expect to contribute to informed and data-driven
-decision making in transportation by disseminating these data. We hope
-that these datasets can substantiate future improvements in
-policy-making that may lead to greater justice in the country's urban
-transportation systems. The package is still in its initial phase and
-may undergo expansions in the future by adding TTM's for other
-destinations (e.g., schools, healthcare facilities). Finally, as an ODP,
-the \{canaccess\} package allows for open exploration, use, and
-contribution by users through its GitHub repository.
+travel time matrices are computed from DA centroid to DA centroid, which
+means that they can be adapted for use in applications with any type of
+destination that is aggregated at the DA level. Finally, as an ODP, the
+\{canaccess\} package allows for open exploration, use, and contribution
+by users through its GitHub repository.
 \end{abstract}
@@
 
-\section{Introduction}\label{introduction}
-
-The objective of this paper is to describe the \{canaccessR\} open data
-package. Its main contents are a set of public transit travel time
-matrices (TTM) estimates to employment and groceries stores from the 12
-largest Canadian metropolitan areas in 2019 and 2023, representing
-approximately 55\% of the Canadian population. The results are provided
-at the Dissemination Areas (DA) \footnote{Dissemination Areas are the
-  smallest publicly available spatial unit provided by Statistics Canada
-  \citep{governmentofcanadaDictionaryCensusPopulation2021a}.} level,
-yielding origin-destination pairs containing associated travel time by
-public transit, population, total employment, mode share, and other
-relevant census variables and spatial shape boundaries for each
-metropolitan area. This data package was created by leveraging expertise
-in data science, R programming, and transportation analysis. It includes
-.Rmd notebooks written for computing travel time matrices for large sets
-of origin-destination pairs using the \{r5r\} R package
-\citep{pereiraR5rRapidRealistic2021}. Overall, \{canaccessR\} offers 52
-complementary objects ready for temporal and spatial analysis. The
-package is an analysis-read product based on a fusion of data sources,
-including public transit schedules (GTFS files), road and transit
-networks, census data, and filtered business location data.
-
-TTMs are a core piece of information required for estimating spatial
-interaction. Accessibility -the \emph{potential} spatial interaction
-offered by the transportation system to reach destinations
-\citep{paezMeasuringAccessibilityPositive2012}- is an example of a
-measure that requires TTMs. Recent efforts following an open-source and
-transparent philosophy have been made to disseminate useful data and
-information on transportation in the Canadian context
-\citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
-pre-processed and available data that allow for the ease estimation of
-accessibility indicators are still scarce. Within this context, we
-expect to help filling this gap by the processing raw into user-ready
-data and making them publicly available to advance knowledge on the
-field. Our main contribution is to provide analysis-ready data for
-Canada's largest cities on the topic of transportation accessibility,
-thus making urban analytics in the country more accessible and
-contributing to future research and data-driven decision making.
-
-The package's main audiences are Canadian researchers in urban planning
-and transportation and transportation system agencies. We anticipate
-three primary uses for the open data product (ODP) described in this
-paper. First, the datasets allow for static assessment of the level of
-public transit accessibility across the country's largest cities before
-and after the COVID-19 pandemic. In other words, \{canaccessR\} makes it
-easier for those interested in comparing cities regarding their level of
-public transit accessibility to essential destinations (such as
-employment centers and groceries stores) to do so. Second, the temporal
-and spatial characters of the datasets made available here allow
-researchers to evaluate accessibility changes through time and across
-space within the largest Canadian urban areas. Third, as is now common
-practice in transportation accessibility research, used as inputs, these
-estimates can substantiate broader investigations on transportation
-justice and equity
-\citep{higginsChangesAccessibilityEmergency2021, humbertoHowTranslateJustice2023, pereiraGeographicAccessCOVID192021}.
-For example, the TTM estimates allow for evaluating the evolution of
-public transit's accessibility by income or spatial distribution across
-all Dissemination Areas (DA's) of each of the 12 cities in the sample
-\citep{pargaDemocraticAccessOur2024}. In other words, the package's
-contents can be used from straightforward assessments of accessibility
-in Canadian urban areas to more theoretically and morally complex
-evaluations of justice in the country's urban transportation system.
-
-Besides this introduction, we organize this paper as follows. The next
-section contains a description of the data sources we used to construct
-the data package. Then, we recount the data processing necessary to
-create the package. Next, we go through the main contents of the data
-package, i.e., the travel time matrices estimated through our analysis.
-We present some basic descriptive statistics of these datasets, and
-elucidate how one can use them in accessibility analysis. Finally, we
-conclude by explaining how we expect \{canaccessR\} to contribute to the
-urban analytics and science community.
-
-\section{Data and methods}\label{data-and-methods}
-
-\subsection{Raw data sources}\label{raw-data-sources}
-
-The locations included in the data package comprise the 12 largest
-(population-wise) Census metropolitan areas (CMA's) based on the 2016
-Canadian Census \citep{governmentofcanada2016CensusPopulation2016}
-\footnote{We included Oshawa as part of the Greater Toronto Area (GTA)
-  because of its proximity. Similarly, we included Abbotsford-Mission as
-  part of the Vancouver metropolitan area due to its proximity to a
-  transit station on the region's West Coast Express commuter rail line.}.These
-locations are the surrunding Toronto, Montreal, Vancouver,
-Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
-Kitchener-Cambridge-Waterloo, London, and Halifax areas. We used four
-main data sources to construct the \{canaccessR\} data package: General
-Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
-Points of Interest, and Statistics Canada Census data.
-
-We manually collected and processed the GTFS files from all transit
-agencies within the selected CMA's to use their information on the
-public transit schedule in 2019 and 2023. The OpenStreetMap data for the
-selected areas were collected through the \{osmextract\} package
-\citep{gilardiOsmextractDownloadImport2025}. We used OSM data from 2019
-and 2023, which provided information on the areas' transit network in
-two points in time. We collected data from the 2016 Canadian Census
-using the \{cancensus\} package
-\citep{vonbergmannCancensusPackageAccess2022} and used its information
-on the spatial distribution of the population and the number of
-workplace locations (employment) across the CMA's
-\citep{governmentofcanada2016CensusPopulation2016}. Finally, we gathered
-and cleaned the 2023 DMTI's Enhanced Points of Interest dataset to
-obtain the location of the groceries stores within every urban area
-selected \citep{dmtispatialincEnhancedPointsInterest2015}. We filtered
-the locations within the DMTI dataset using the grocery stores code from
-the North American Industry Classification System (NAICS) and the
-Standard Industrial Classification (SIC).
-
-\subsection{Methods: travel time matrices
-processing}\label{methods-travel-time-matrices-processing}
-
-Using the \{r5r\} R package, we estimated public transit travel times
-for two destination types, grocery stores and jobs. For each amenity
-type, we chose a likely travel time and day of the week. We set a 15
-minutes time window and the maximum trip duration to 120 minutes. The
-estimated times are the median of the 15 minute time window. For
-groceries stores, we set the departure date to a weekend afternoon and
-the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and
-April 22, 2023. For employment, we ran the analysis on a typical weekday
-morning rush-hour commute, more specifically 8:00 to 8:15 AM departure
-on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 \footnote{The one
-  exception is Quebec City, where the routing for 2019 occurs on a
-  Saturday and Tuesday in June (instead of April) due to the GTFS data
-  unavailability.}. In both cases, we assumed that walking was the mode
-of travel from origin to transit stop and from transit stop to
-destination. We aggregated all the resulting travel time matrices at the
-DA level, which comprise the fundamental spatial unit of analysis in
-data package.
-
-\section{\{canaccessR\}'s contents}\label{canaccessrs-contents}
-
-Specifically, the package contains the following contents: 10 data.frame
-objects containing the calculated public travel times from DA centroids
-to grocery stores and another 10 for DA centroids to DA centroids.
-Notably, the \{travel\_matrix\_grc\_ggh\} data.frame, is named after the
-acronym for the Greater Golden Horseshoe area, which includes the CMA
-regions of Toronto, Hamilton, and Kitchener-Cambridge-Waterloo. Hence,
-the 20 travel time data.frames represent public transit travel times for
-all 12 CMAs across two sets of destinations. Next, 10 sf objects
-represent census data for each DA, including dwelling counts, population
-by age bracket, single-parent-headed households, low-income prevalence,
-official language knowledge, housing- quality, ownership and
-affordability variables, visible minority, newcomer- and immigration-
-related variables, educational attainment, and commuting mode shares.
-Furthermore, 10 and 10 sf objects represent the CMA areas' boundaries
-and backgrounds for plotting the data spatially, respectively. Finally,
-2 data.frames contain aggregated population and transit statistics at
-the CMA level.
-
-We now present some descriptive statistics from the travel time matrices
-contained in the \{canaccessR\} package. Table \citet{tbl-table_1}
-summarizes the travel time estimates. The travel time objects contain,
-in total, 97,784,850 origin-destination pairs (observations) from
-population to employment locations for all the DA's in the sample and
-18,519,897 pairs from population to groceries stores. Considering all
-areas combined, the mean travel time to jobs was 80 and 76 for groceries
-stores.
-
-\begingroup
-\setlength\LTleft{0\linewidth}
-\setlength\LTright{0\linewidth}\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{longtable}{@{\extracolsep{\fill}}clrrrrrr}
-\toprule
-Study Region Name & Destination & Observations & Mean & Sd & P25 & P50 & P75 \\ 
-\midrule\addlinespace[2.5pt]
-All regions & Employment & 97,784,850 & 78 & 25 & 59 & 80 & 99 \\ 
-Toronto & Employment & 41,038,062 & 82 & 25 & 64 & 85 & 103 \\ 
-Montréal & Employment & 30,091,411 & 76 & 25 & 57 & 77 & 97 \\ 
-Vancouver & Employment & 12,254,478 & 76 & 25 & 58 & 77 & 96 \\ 
-Calgary & Employment & 3,022,244 & 75 & 23 & 59 & 75 & 91 \\ 
-Ottawa & Employment & 2,936,912 & 74 & 24 & 56 & 74 & 93 \\ 
-Edmonton & Employment & 2,057,486 & 70 & 23 & 53 & 70 & 87 \\ 
-Québec City & Employment & 1,473,702 & 72 & 25 & 54 & 72 & 93 \\ 
-Winnipeg & Employment & 1,574,205 & 62 & 22 & 46 & 61 & 76 \\ 
-Hamilton & Employment & 2,089,172 & 82 & 28 & 61 & 87 & 106 \\ 
-Waterloo & Employment & 552,826 & 71 & 27 & 49 & 69 & 93 \\ 
-London & Employment & 415,174 & 61 & 21 & 46 & 60 & 74 \\ 
-Halifax & Employment & 279,178 & 66 & 25 & 47 & 65 & 84 \\ 
-All regions & Groceries Stores & 18,519,897 & 76 & 26 & 56 & 76 & 97 \\ 
-Toronto & Groceries Stores & 8,512,874 & 80 & 25 & 61 & 82 & 101 \\ 
-Montréal & Groceries Stores & 2,993,965 & 73 & 26 & 53 & 74 & 94 \\ 
-Vancouver & Groceries Stores & 4,540,106 & 73 & 25 & 53 & 74 & 93 \\ 
-Calgary & Groceries Stores & 502,757 & 70 & 22 & 54 & 70 & 86 \\ 
-Ottawa & Groceries Stores & 619,791 & 72 & 24 & 55 & 72 & 91 \\ 
-Edmonton & Groceries Stores & 324,030 & 68 & 23 & 51 & 67 & 84 \\ 
-Québec City & Groceries Stores & 234,600 & 70 & 26 & 51 & 70 & 89 \\ 
-Winnipeg & Groceries Stores & 362,566 & 58 & 21 & 43 & 57 & 72 \\ 
-Hamilton & Groceries Stores & 281,118 & 84 & 29 & 60 & 91 & 110 \\ 
-Waterloo & Groceries Stores & 44,726 & 66 & 28 & 44 & 62 & 90 \\ 
-London & Groceries Stores & 52,617 & 55 & 19 & 42 & 55 & 67 \\ 
-Halifax & Groceries Stores & 50,747 & 63 & 26 & 42 & 63 & 83 \\ 
-\bottomrule
-\end{longtable}
-\endgroup
-
-\section{\{canaccessR\} usage}\label{canaccessr-usage}
-
-This section presents an usage example of the package through visual
-representation of its data. In Figure
-\citet{fig-travel_time_emp_grc_plot}, we show the spatial distribution
-of the travel time matrices for the metropolitan region of Montréal. In
-it, we see the median travel time from each DA to employment (left) and
-to groceries stores (right). The plot shows that moving away from the
-city core increases the necessary travel time by public transit to reach
-employment locations and groceries stores.
-
-\begin{figure}[H]
-\includegraphics[width=1\linewidth]{../figures/patch_tt_emp_grc} \caption{Estimated median travel time (minutes) per Dissemination Area to jobs (left) and groceries stores (right).
-Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016).}\label{fig:fig-travel_time_emp_grc_plot}
-\end{figure}
-
-A more thorough example of the package's use can be found on the School
-of Cities' recent report on Canada's Urban Infrastructure Deficit. In
-its 11th Chapter, we use the travel time matrices to estimate
-accessibility metrics to jobs and groceries stores before and after the
-pandemic \citep{pargaDemocraticAccessOur2024}. We then compare how
-changes affected groups differently according to their spatial
-distribution and income level, thus making explicit the connection of
-the package's information and matters of equity in transportation. The
-report is freely available for download at the State of Cities Summit
-\href{https://stateofcitiessummit.ca/report}{website}.
-
-\section{Concluding remarks}\label{concluding-remarks}
-
-In this paper, we describe the \{canaccessR\} data package, created
-using the \{r5r\} package and transit schedule, street network,
-employment, and population data. The package's main contents refers to
-the ready-to-use travel time matrices for public transit to reach
-employment and groceries stores in Canada's 12 largest urban areas. We
-expect the contents of the package to be used in transportation
-accessibility evaluations within and across those regions. Moreover,
-these datasets can be used in further equity assessments that evaluate
-the distribution of accessibility across space and between social
-groups. Furthermore, in the spirit of open data products
-\citep{arribas-belOpenDataProductsA2021}, the package can be expanded
-through collaboration with other researchers by, for example, including
-travel time matrices to other essential destinations within the DMTI's
-dataset (\emph{e.g.}, schools, healthcare, etc.). In other words, we
-hope that by making these datasets publicly available, future analysis
-can contribute to making Canada's transportation system more just and
-fair, considering accessibility's as the main social good of
-transportation \citep{martensTransportJusticeDesigning2016}, and the
-inherent connection between public transit and the ``right to the city''
-\citep{cogginRightTransportMoving2015}.
-
-\section{Declaration of Conflicting
-Interests}\label{declaration-of-conflicting-interests}
-
-The author(s) declared no potential conflicts of interest with respect
-to the research, authorship, and/or publication of this article.
-
-\section{Funding}\label{funding}
-
-The author(s) disclosed receipt of the following financial support for
-the research, authorship, and/or publication of this article: This work
-was supported by the Social Sciences and Humanities Research Council of
-Canada (\emph{More description about the funding source after the review
-process}).
-
-\section{ORCID}\label{orcid}
-
-\begin{itemize}
-\tightlist
-\item
-  name: João Pedro Figueira Amorim Parga orcid: 0000-0002-4105-5927
-\item
-  name: Anastasia Soukhov orcid: 0000-0003-4371-4831
-\item
-  name: Robert Nutifafa Arku orcid: 0000-0002-2018-886X
-\item
-  name: Christopher Higgins orcid: 0000-0002-3551-7750
-\item
-  name: Antonio Páez orcid: 0000-0001-6912-9919
-\end{itemize}
-
-\section{Data availability statement}\label{data-availability-statement}
-
-The \{canaccessR\} data package can be found and installed on its Github
-\href{https://github.com/paezha/canaccessR}{respository}.
-
 \bibliographystyle{sageh}
-\bibliography{bibfile.bib}
+\bibliography{bibfile}
 

### 2025-05-01 paper/Manuscript-Data-Package.Rmd
Version of paper submitted to EPB

@@
 ---
-title: "canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada’s largest metropolitan areas."
+title: "canaccessR: An open data product for transit accessibility analysis in Canada’s largest metropolitan areas."
 author:
-  - name: João Pedro Figueira Amorim Parga
-    email: joao.parga@mail.utoronto.ca
+  - name: Author One
+    email: author1@mail.ca
     affil-id: 1
-    orcid: 0000-0002-4105-5927
+    orcid: 0000-0000-0000-000
     footnote: 1
-  - name: Anastasia Soukhov
-    email: soukhoa@mcmaster.ca
+  - name: Author Two
+    email: author2@mail.ca
     affil-id: 2
-    orcid: 0000-0003-4371-4831
-  - name: Robert Nutifafa Arku
-    email: rn.arku@mail.utoronto.ca
-    affil-id: 1
-    orcid: 0000-0002-2018-886X
-  - name: Christopher Higgins
-    email: cd.higgins@utoronto.ca
-    affil-id: 3
-    orcid: 0000-0002-3551-7750
-  - name: Antonio Páez
-    email: paezha@mcmaster.ca
-    affil-id: 2
-    orcid: 0000-0001-6912-9919
+    orcid: 0000-0000-0000-0000
 affiliations:
   - id: 1
-    name: University of Toronto
+    name: University
   - id: 2
-    name: McMaster University
-  - id: 3
-    name: University of Toronto Scarborough
+    name: College
 # address:
@@
 This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility levels to these two essential destinations within these urban areas.
-Moreover, these estimations can be used as inputs in equity and inequalities assessments, through the comparisons of within and across accessibility levels found throughout Canada's largest metropolitan areas.
-Consequently, we expect to contribute to informed and data-driven decision-making in transportation by disseminating these data.
-We hope that these datasets can substantiate future improvements in policy-making that may lead to greater justice in the country's urban transportation systems.
-The package is still in its initial phase and may undergo expansions in the future by adding TTMs for other destinations (e.g., schools, healthcare facilities).
+Moreover, travel time matrices are computed from DA centroid to DA centroid, which means that they can be adapted for use in applications with any type of destination that is aggregated at the DA level.
 Finally, as an ODP, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."
@@
   - times
-bibliography: bibfile.bib
+bibliography: bibfile
 bibliographystyle: sageh
 output:
-  word_document: default
-  rticles::sage_article:
-    keep_tex: yes
-    fig_caption: yes
+  #rticles::sage_article:
+  #  keep_tex: yes
+  #  fig_caption: yes
+  word_document
 header-includes: \usepackage{float}
@@
 # Load packages
-library(tidyverse)
+library(canaccessR)
 library(data.table)
-# library(summarytools)
+library(glue)
 library(gt)
 library(gtsummary)
+# library(summarytools)
 # library(sf)
+library(tidyverse)
 # library(tmap)
@@
 
+```{r number-origin-destination-pairs, eval=FALSE}
+metro_regions <- c("cal", "edm", "ggh", "hal", "ldn", "mtl", "ott", "que", "van", "win")
+
+n <- 0
+for(i in metro_regions){
+  eval(str2lang(glue("data(travel_matrix_emp_{i})")))
+  eval(str2lang(glue("n <- n + nrow(travel_matrix_emp_{i})")))
+  eval(str2lang(glue("rm(travel_matrix_emp_{i})")))
+}
+
+# There are 97,784,850 origin destination pairs
+```
+
 # Introduction
@@
 The objective of this paper is to describe the {canaccessR} open data package.
-Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and grocery stores from the 12 largest Canadian metropolitan areas in 2019 and 2023, representing approximately 55% of the Canadian population.
-The results are provided at the Dissemination Areas (DA) [^1] level, yielding origin-destination pairs containing associated travel time by public transit, population, total employment, mode share, and other relevant census variables and spatial shape boundaries for each metropolitan area.
+Open data products are a core element of reproducible research in the spatial sciences [@paez_open_2021], and facilitate access while adding value to data [@arribas-bel2021].
+The main contents of {canaccessR} are a set of public transit travel time matrices (TTM) estimates to Dissemination Areas (DA) [^1] as a set of standard Census zones, as well as to grocery stores as points.
+In addition,  employment data at the level of DAs are also furnished.
+These matrices are provided for the 12 largest Canadian metropolitan areas in 2019 and 2023, representing approximately 55% of the Canadian population.
+The matrices include travel time for origin-destination pairs containing associated travel time by public transit, population, total employment, mode share, and other relevant census variables and spatial shape boundaries for each metropolitan area.
 This data package was created by leveraging expertise in data science, R programming, and transportation analysis.
-It includes .Rmd notebooks written for computing travel time matrices for large sets of origin-destination pairs using the {r5r} R package [@pereiraR5rRapidRealistic2021].
-Overall, {canaccessR} offers `r data(package = "canaccessR")$results[, "Item"] |> length()` complementary objects ready for temporal and spatial analysis.
-The package is an analysis-read product based on a fusion of data sources, including public transit schedules (GTFS files), road and transit networks, census data, and filtered business location data.
+It includes .Rmd notebooks written for computing travel time matrices for large sets of destination pairs using the {r5r} R package [@pereiraR5rRapidRealistic2021].
+Overall, {canaccessR} offers `r data(package = "canaccessR")$results[, "Item"] |> length()` complementary objects ready for temporal and spatial analysis and, between groceries and DAS, travel times for almost 120 million origin-destination pairs.
+The package is an analysis-ready product based on a fusion of data sources, including public transit schedules (GTFS files), road and transit networks, census data, and filtered business location data.
 
@@
 
-TTMs are a core piece of information required for estimating spatial interaction.
-Accessibility -the _potential_ spatial interaction offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012]- is an example of a measure that requires TTMs.
+TTMs are a core piece of information required for many tasks in transit and transportation planning.
+As an example, TTMs are essential for the analysis of accessibility--the _potential_ for spatial interaction offered by the transportation system [@paezMeasuringAccessibilityPositive2012].
 Recent efforts following an open-source and transparent philosophy have been made to disseminate useful data and information on transportation in the Canadian context [@soukhovTTS2016RDataSet2023].
 However, despite these initiatives, readily available pre-processed data that facilitate the estimation of accessibility indicators remain scarce.
-Within this context, we aim to help fill this gap by processing raw data into user-ready formats and making them publicly available to advance knowledge in the field.
-Our main contribution is to provide analysis-ready transportation accessibility data for Canada's largest cities, thus making urban analytics in the country more accessible and contributing to future research and data-driven decision-making.
+Within this context, we aim to help fill this gap by processing raw data into analysis-ready formats and making them publicly available to advance knowledge in the field.
+Our main contribution is to provide easily accessible transportation accessibility data for Canada's largest cities, thus making urban analytics in the country simpler and more affordable, thus contributing to future research and data-driven decision-making.
 
-The package's main audiences are Canadian researchers in urban planning and transportation and transportation system agencies.
+The package's main audiences are Canadian researchers in urban planning and transportation, as well as transit agencies.
 We anticipate three primary uses for the open data product (ODP) described in this paper.
-First, the datasets allow for static assessment of the level of public transit accessibility across the country's largest cities before and after the COVID-19 pandemic.
+First, the data sets allow for static assessment of the level of public transit accessibility across the country's largest cities before and after the COVID-19 pandemic.
 In other words, {canaccessR} makes it easier for those interested in comparing cities regarding their level of public transit accessibility to essential destinations (such as employment centers and grocery stores) to do so.
-Second, the temporal and spatial characters of the datasets made available here allow researchers to evaluate accessibility changes through time and across space within the largest Canadian urban areas.
+Second, the temporal and spatial characters of the data sets made available here allow researchers to evaluate accessibility changes through time and across space within the largest Canadian urban areas.
 Third, as is now common practice in transportation accessibility research, used as inputs, these estimates can substantiate broader investigations on transportation justice and equity [@higginsChangesAccessibilityEmergency2021; @humbertoHowTranslateJustice2023; @pereiraGeographicAccessCOVID192021].
-For example, the TTM estimates allow for evaluating the evolution of public transit's accessibility by income or spatial distribution across all Dissemination Areas (DAs) of each of the 12 cities in the sample [@pargaDemocraticAccessOur2024].
+As an example, the TTM estimates allow for the evaluation of the evolution of public transit's accessibility by income or spatial distribution across all Dissemination Areas (DAs) of each of the 12 cities in the sample [@pargaDemocraticAccessOur2024].
 In other words, the package's contents can be used from straightforward assessments of accessibility in Canadian urban areas to more theoretically and morally complex evaluations of justice in the country's urban transportation system.
 
-Besides this introduction, we organize this paper as follows: The next section contains a description of the data sources we used to construct the data package.
+After this introduction, the paper is organizes as follows: the following section contains a description of the data sources we used to construct the data package.
 Then, we recount the data processing necessary to create the package.
 Next, we go through the main contents of the data package, i.e., the travel time matrices estimated through our analysis.
-We present some basic descriptive statistics of these datasets and elucidate how one can use them in accessibility analysis.
-Finally, we conclude by explaining how we expect {canaccessR} to contribute to the urban analytics and science community.
+We present some basic descriptive statistics of these data sets and elucidate potential uses.
+Finally, we conclude by explaining how we expect {canaccessR} to contribute to the urban analytics and city science community.
 
@@
 
-Using the {r5r} R package, we estimated public transit travel times for two destination types, grocery stores and jobs.
-For each amenity type, we chose a likely travel time and day of the week.
+We used the {r5r} R package to estimate public transit travel times for two destination types, grocery stores (points) and DA centroids (zones).
+For each we chose a likely travel time and day of the week.
 We set a 15-minute time window and the maximum trip duration to 120 minutes.
@@
 For grocery stores, we set the departure date to a weekend afternoon and the departure time to between 12:00 PM to 12:15 PM on April 20, 2019, and April 22, 2023.
-For employment, we ran the analysis on a typical weekday morning rush-hour commute, more specifically 8:00 to 8:15 AM departure on Tuesday, April 16, 2019 and Tuesday, April 18, 2023.[^3] In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
+For DAs, we ran the analysis on a typical weekday morning rush-hour commute, more specifically 8:00 to 8:15 AM departure on Tuesday, April 16, 2019 and Tuesday, April 18, 2023.[^3] In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
 We aggregated all the resulting travel time matrices at the DA level, which comprise the fundamental spatial unit of analysis in the data package.
@@
 Finally, `r grep("statistics", data(package = "canaccessR")$results[, "Item"]) |> length()` *data.frame*s contain aggregated population and transit statistics at the CMA level.
-
-<!--AS: Joao, I rephrased the below into one paragraph, what do you think?  -->
-
-<!-- The main contents of the {canaccessR} package are the travel time matrices. -->
-<!-- <!-- These matrices are data frames with origin-destination pairs, travel time by public transit, and population and employment by Census Dissemination Area.  --> 
-<!-- These matrices (datasets) comprise the estimated travel time by transit from and to each origin and destination Dissemination Area pairs.
-The datasets also feature the total population in the origin DA and the total employment in the destination DA.
-Besides these information, the matrices contain the unique origin and destination DA codes, region and name identifiers, the year in which the travel times refers to, and the date and time of departure for a trip that originates at the Dissemination Area of origin.
-There are two matrices for each metropolitan area, one containing the travel times to jobs and the other to grocery stores [^4].
-These information can be turned into spatial data using the socio-economic and demographic datasets.
--->
-
-<!-- [^4]: The information from Toronto, Hamilton, and Waterloo are aggregated at the Greater Golden Horseshoe travel time matrices, making it 20 datasets in total.  -->
-
-<!-- The socio-economic and demographic data contain other information disaggregated at the DA level for each CMA.  -->
-<!-- These refer to total population by age groups, number of dwellings, number of individuals below the Low Income Measure, etc. -->
-<!-- The aggregated statistics refer to population aggregates (for the selected CMAs) and transit revenue and ridership data aggregated by regional and national scale. -->
 
@@
 In it, we see the median travel time from each DA to employment (left) and to grocery stores (right).
-The plot shows that moving away from the city core increases the necessary travel time by public transit to reach employment locations and grocery stores.
+The plot shows that moving away from the city core increases the necessary travel time by public transit to reach both employment locations and grocery stores.
+This result is expected for employment, given the typical concentration of this type of opportunity in downtown centers, but it also points out at a distribution of grocery stores that also requires increased travel for non-central city residents.
 
@@
 The report is freely available for download at the State of Cities Summit [website](https://stateofcitiessummit.ca/report).
+
+Since a set of travel time matrices (those labeled as employment) were computed from DA centroid to DA centroid, users of the package can easily incorporate any other type of opportunity aggregated at the DA level to expand the range of possible applications of the data package.
 
@@
 
+<!-- 
 # Funding
@@
     orcid: 0000-0001-6912-9919
+-->
 
@@
 
+For peer-review purposes the package is shared as a source package. Once anonymity can be lifted, the address of the repository will be published.
+
+The package can be installed locally in R from file `canaccessR_0.0.1.tar.gz` in this way:
+
+```
+install.packages("path_to_file/canaccessR_0.0.1.tar.gz", 
+  repos = NULL, 
+  type="source")
+```
+
+The file is available for download using this link:
+
+https://drive.google.com/drive/folders/1CEm-jHyp1VBB27Mq4UWslANhHDKxUOju?usp=drive_link
+
+<!--
 The {canaccessR} data package can be found and installed on its Github [repository](https://github.com/paezha/canaccessR).
+-->
 

### 2025-05-01 paper/Manuscript-Data-Package.tex
Version of paper submitted to EPB

@@
 
-\author{João Pedro Figueira Amorim Parga\affilnum{}, Anastasia
-Soukhov\affilnum{}, Robert Nutifafa Arku\affilnum{}, Christopher
-Higgins\affilnum{}, Antonio Páez\affilnum{}}
+\author{Author One\affilnum{}, Author Two\affilnum{}}
 
@@
 The objective of this paper is to describe the \{canaccessR\} open data
-package. Its main contents are a set of public transit travel time
-matrices (TTM) estimates to employment and groceries stores from the 12
-largest Canadian metropolitan areas in 2019 and 2023, representing
-approximately 55\% of the Canadian population. The results are provided
-at the Dissemination Areas (DA) \footnote{Dissemination Areas are the
-  smallest publicly available spatial unit provided by Statistics Canada
-  \citep{governmentofcanadaDictionaryCensusPopulation2021a}.} level,
-yielding origin-destination pairs containing associated travel time by
-public transit, population, total employment, mode share, and other
-relevant census variables and spatial shape boundaries for each
-metropolitan area. This data package was created by leveraging expertise
-in data science, R programming, and transportation analysis. It includes
-.Rmd notebooks written for computing travel time matrices for large sets
-of origin-destination pairs using the \{r5r\} R package
-\citep{pereiraR5rRapidRealistic2021}. Overall, \{canaccessR\} offers 52
-complementary objects ready for temporal and spatial analysis. The
-package is an analysis-read product based on a fusion of data sources,
-including public transit schedules (GTFS files), road and transit
-networks, census data, and filtered business location data.
+package. Open data products are a core element of reproducible research
+in the spatial sciences \citep{paez_open_2021}, and facilitate access
+while adding value to data \citep{arribas-bel2021}. The main contents of
+\{canaccessR\} are a set of public transit travel time matrices (TTM)
+estimates to Dissemination Areas (DA) \footnote{Dissemination Areas are
+  the smallest publicly available spatial unit provided by Statistics
+  Canada \citep{governmentofcanadaDictionaryCensusPopulation2021a}.} as
+a set of standard Census zones, as well as to grocery stores as points.
+In addition, employment data at the level of DAs are also furnished.
+These matrices are provided for the 12 largest Canadian metropolitan
+areas in 2019 and 2023, representing approximately 55\% of the Canadian
+population. The matrices include travel time for origin-destination
+pairs containing associated travel time by public transit, population,
+total employment, mode share, and other relevant census variables and
+spatial shape boundaries for each metropolitan area. This data package
+was created by leveraging expertise in data science, R programming, and
+transportation analysis. It includes .Rmd notebooks written for
+computing travel time matrices for large sets of destination pairs using
+the \{r5r\} R package \citep{pereiraR5rRapidRealistic2021}. Overall,
+\{canaccessR\} offers 52 complementary objects ready for temporal and
+spatial analysis and, between groceries and DAS, travel times for almost
+120 million origin-destination pairs. The package is an analysis-ready
+product based on a fusion of data sources, including public transit
+schedules (GTFS files), road and transit networks, census data, and
+filtered business location data.
 
-TTMs are a core piece of information required for estimating spatial
-interaction. Accessibility -the \emph{potential} spatial interaction
-offered by the transportation system to reach destinations
-\citep{paezMeasuringAccessibilityPositive2012}- is an example of a
-measure that requires TTMs. Recent efforts following an open-source and
-transparent philosophy have been made to disseminate useful data and
-information on transportation in the Canadian context
+TTMs are a core piece of information required for many tasks in transit
+and transportation planning. As an example, TTMs are essential for the
+analysis of accessibility--the \emph{potential} for spatial interaction
+offered by the transportation system
+\citep{paezMeasuringAccessibilityPositive2012}. Recent efforts following
+an open-source and transparent philosophy have been made to disseminate
+useful data and information on transportation in the Canadian context
 \citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
-pre-processed and available data that allow for the ease estimation of
-accessibility indicators are still scarce. Within this context, we
-expect to help filling this gap by the processing raw into user-ready
-data and making them publicly available to advance knowledge on the
-field. Our main contribution is to provide analysis-ready data for
-Canada's largest cities on the topic of transportation accessibility,
-thus making urban analytics in the country more accessible and
-contributing to future research and data-driven decision making.
+readily available pre-processed data that facilitate the estimation of
+accessibility indicators remain scarce. Within this context, we aim to
+help fill this gap by processing raw data into analysis-ready formats
+and making them publicly available to advance knowledge in the field.
+Our main contribution is to provide easily accessible transportation
+accessibility data for Canada's largest cities, thus making urban
+analytics in the country simpler and more affordable, thus contributing
+to future research and data-driven decision-making.
 
 The package's main audiences are Canadian researchers in urban planning
-and transportation and transportation system agencies. We anticipate
-three primary uses for the open data product (ODP) described in this
-paper. First, the datasets allow for static assessment of the level of
-public transit accessibility across the country's largest cities before
-and after the COVID-19 pandemic. In other words, \{canaccessR\} makes it
+and transportation, as well as transit agencies. We anticipate three
+primary uses for the open data product (ODP) described in this paper.
+First, the data sets allow for static assessment of the level of public
+transit accessibility across the country's largest cities before and
+after the COVID-19 pandemic. In other words, \{canaccessR\} makes it
 easier for those interested in comparing cities regarding their level of
 public transit accessibility to essential destinations (such as
-employment centers and groceries stores) to do so. Second, the temporal
-and spatial characters of the datasets made available here allow
+employment centers and grocery stores) to do so. Second, the temporal
+and spatial characters of the data sets made available here allow
 researchers to evaluate accessibility changes through time and across
@@
 \citep{higginsChangesAccessibilityEmergency2021, humbertoHowTranslateJustice2023, pereiraGeographicAccessCOVID192021}.
-For example, the TTM estimates allow for evaluating the evolution of
-public transit's accessibility by income or spatial distribution across
-all Dissemination Areas (DA's) of each of the 12 cities in the sample
-\citep{pargaDemocraticAccessOur2024}. In other words, the package's
-contents can be used from straightforward assessments of accessibility
-in Canadian urban areas to more theoretically and morally complex
-evaluations of justice in the country's urban transportation system.
+As an example, the TTM estimates allow for the evaluation of the
+evolution of public transit's accessibility by income or spatial
+distribution across all Dissemination Areas (DAs) of each of the 12
+cities in the sample \citep{pargaDemocraticAccessOur2024}. In other
+words, the package's contents can be used from straightforward
+assessments of accessibility in Canadian urban areas to more
+theoretically and morally complex evaluations of justice in the
+country's urban transportation system.
 
-Besides this introduction, we organize this paper as follows. The next
-section contains a description of the data sources we used to construct
-the data package. Then, we recount the data processing necessary to
-create the package. Next, we go through the main contents of the data
-package, i.e., the travel time matrices estimated through our analysis.
-We present some basic descriptive statistics of these datasets, and
-elucidate how one can use them in accessibility analysis. Finally, we
-conclude by explaining how we expect \{canaccessR\} to contribute to the
-urban analytics and science community.
+After this introduction, the paper is organizes as follows: the
+following section contains a description of the data sources we used to
+construct the data package. Then, we recount the data processing
+necessary to create the package. Next, we go through the main contents
+of the data package, i.e., the travel time matrices estimated through
+our analysis. We present some basic descriptive statistics of these data
+sets and elucidate potential uses. Finally, we conclude by explaining
+how we expect \{canaccessR\} to contribute to the urban analytics and
+city science community.
 
@@
 The locations included in the data package comprise the 12 largest
-(population-wise) Census metropolitan areas (CMA's) based on the 2016
+(population-wise) census metropolitan areas (CMAs) based on the 2016
 Canadian Census \citep{governmentofcanada2016CensusPopulation2016}
@@
   part of the Vancouver metropolitan area due to its proximity to a
-  transit station on the region's West Coast Express commuter rail line.}.These
-locations are the surrunding Toronto, Montreal, Vancouver,
+  transit station on the region's West Coast Express commuter rail line.}.
+These locations are the surrounding Toronto, Montreal, Vancouver,
 Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
@@
 We manually collected and processed the GTFS files from all transit
-agencies within the selected CMA's to use their information on the
-public transit schedule in 2019 and 2023. The OpenStreetMap data for the
+agencies within the selected CMAs to use their information on the public
+transit schedule in 2019 and 2023. The OpenStreetMap data for the
 selected areas were collected through the \{osmextract\} package
@@
 on the spatial distribution of the population and the number of
-workplace locations (employment) across the CMA's
+workplace locations (employment) across the CMAs
 \citep{governmentofcanada2016CensusPopulation2016}. Finally, we gathered
 and cleaned the 2023 DMTI's Enhanced Points of Interest dataset to
-obtain the location of the groceries stores within every urban area
+obtain the location of the grocery stores within every urban area
 selected \citep{dmtispatialincEnhancedPointsInterest2015}. We filtered
@@
 
-Using the \{r5r\} R package, we estimated public transit travel times
-for two destination types, grocery stores and jobs. For each amenity
-type, we chose a likely travel time and day of the week. We set a 15
-minutes time window and the maximum trip duration to 120 minutes. The
-estimated times are the median of the 15 minute time window. For
-groceries stores, we set the departure date to a weekend afternoon and
-the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and
-April 22, 2023. For employment, we ran the analysis on a typical weekday
-morning rush-hour commute, more specifically 8:00 to 8:15 AM departure
-on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 \footnote{The one
-  exception is Quebec City, where the routing for 2019 occurs on a
-  Saturday and Tuesday in June (instead of April) due to the GTFS data
-  unavailability.}. In both cases, we assumed that walking was the mode
-of travel from origin to transit stop and from transit stop to
-destination. We aggregated all the resulting travel time matrices at the
-DA level, which comprise the fundamental spatial unit of analysis in
-data package.
+We used the \{r5r\} R package to estimate public transit travel times
+for two destination types, grocery stores (points) and DA centroids
+(zones). For each we chose a likely travel time and day of the week. We
+set a 15-minute time window and the maximum trip duration to 120
+minutes. The estimated times are the median of the 15 minute time
+window. For grocery stores, we set the departure date to a weekend
+afternoon and the departure time to between 12:00 PM to 12:15 PM on
+April 20, 2019, and April 22, 2023. For DAs, we ran the analysis on a
+typical weekday morning rush-hour commute, more specifically 8:00 to
+8:15 AM departure on Tuesday, April 16, 2019 and Tuesday, April 18,
+2023.\footnote{The one exception is Quebec City, where the routing for
+  2019 occurs on a Saturday and Tuesday in June (instead of April) due
+  to the GTFS data unavailability.} In both cases, we assumed that
+walking was the mode of travel from origin to transit stop and from
+transit stop to destination. We aggregated all the resulting travel time
+matrices at the DA level, which comprise the fundamental spatial unit of
+analysis in the data package.
 
@@
 single-parent-headed households, low-income prevalence, official
-language knowledge, housing- quality, ownership and affordability
-variables, visible minority, newcomer- and immigration- related
+language knowledge, housing quality, ownership and affordability
+variables, visible minorities, newcomer- and immigration-related
 variables, educational attainment, and commuting mode shares.
@@
 in total, 97,784,850 origin-destination pairs (observations) from
-population to employment locations for all the DA's in the sample and
-18,519,897 pairs from population to groceries stores. Considering all
-areas combined, the mean travel time to jobs was 78 and 76 to groceries
+population to employment locations for all the DAs in the sample and
+18,519,897 pairs from population to grocery stores. Considering all
+areas combined, the mean travel time to jobs was 78 and 76 to grocery
 stores.
 
-\begingroup
-\setlength\LTleft{0\linewidth}
-\setlength\LTright{0\linewidth}\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{longtable}{@{\extracolsep{\fill}}clrrrrrr}
+\begin{table}[t]
+\fontsize{7.5pt}{9.0pt}\selectfont
+\begin{tabular*}{1\linewidth}{@{\extracolsep{\fill}}clrrrrrr}
 \toprule
@@
 
-This section presents an usage example of the package through visual
+This section presents a usage example of the package through a visual
 representation of its data. In Figure
@@
 it, we see the median travel time from each DA to employment (left) and
-to groceries stores (right). The plot shows that moving away from the
-city core increases the necessary travel time by public transit to reach
-employment locations and groceries stores.
+to grocery stores (right). The plot shows that moving away from the city
+core increases the necessary travel time by public transit to reach both
+employment locations and grocery stores. This result is expected for
+employment, given the typical concentration of this type of opportunity
+in downtown centers, but it also points out at a distribution of grocery
+stores that also requires increased travel for non-central city
+residents.
 
 \begin{figure}[H]
-\includegraphics[width=1\linewidth]{../figures/patch_tt_emp_grc} \caption{Estimated median travel time (minutes) per Dissemination Area to jobs (left) and groceries stores (right).
+\includegraphics[width=1\linewidth]{../figures/patch_tt_emp_grc} \caption{Estimated median travel time (minutes) per Dissemination Area to jobs (left) and grocery stores (right).
 Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016).}\label{fig:fig-travel_time_emp_grc_plot}
@@
 
-A more thorough example of the package's use can be found on the School
+A more thorough example of the package's use can be found in the School
 of Cities' recent report on Canada's Urban Infrastructure Deficit. In
 its 11th Chapter, we use the travel time matrices to estimate
-accessibility metrics to jobs and groceries stores before and after the
+accessibility metrics to jobs and grocery stores before and after the
 pandemic \citep{pargaDemocraticAccessOur2024}. We then compare how
 changes affected groups differently according to their spatial
-distribution and income level, thus making explicit the connection of
-the package's information and matters of equity in transportation. The
-report is freely available for download at the State of Cities Summit
-\href{https://stateofcitiessummit.ca/report}{website}.
+distribution and income level, thus making explicit the connection
+between the package's information and matters of equity in
+transportation. The report is freely available for download at the State
+of Cities Summit \href{https://stateofcitiessummit.ca/report}{website}.
+
+Since a set of travel time matrices (those labeled as employment) were
+computed from DA centroid to DA centroid, users of the package can
+easily incorporate any other type of opportunity aggregated at the DA
+level to expand the range of possible applications of the data package.
 
@@
 using the \{r5r\} package and transit schedule, street network,
-employment, and population data. The package's main contents refers to
+employment, and population data. The package's main contents refer to
 the ready-to-use travel time matrices for public transit to reach
-employment and groceries stores in Canada's 12 largest urban areas. We
+employment and grocery stores in Canada's 12 largest urban areas. We
 expect the contents of the package to be used in transportation
@@
 fair, considering accessibility's as the main social good of
-transportation \citep{martensTransportJusticeDesigning2016}, and the
+transportation \citep{martensTransportJusticeDesigning2016} and the
 inherent connection between public transit and the ``right to the city''
@@
 
-\section{Funding}\label{funding}
-
-The author(s) disclosed receipt of the following financial support for
-the research, authorship, and/or publication of this article: This work
-was supported by the Social Sciences and Humanities Research Council of
-Canada (\emph{More description about the funding source after the review
-process}).
-
-\section{ORCID}\label{orcid}
-
-\begin{itemize}
-\tightlist
-\item
-  name: João Pedro Figueira Amorim Parga orcid: 0000-0002-4105-5927
-\item
-  name: Anastasia Soukhov orcid: 0000-0003-4371-4831
-\item
-  name: Robert Nutifafa Arku orcid: 0000-0002-2018-886X
-\item
-  name: Christopher Higgins orcid: 0000-0002-3551-7750
-\item
-  name: Antonio Páez orcid: 0000-0001-6912-9919
-\end{itemize}
-
 \section{Data availability statement}\label{data-availability-statement}
 
-The \{canaccessR\} data package can be found and installed on its Github
-\href{https://github.com/paezha/canaccessR}{respository}.
+For peer-review purposes the package is shared as a source package. Once
+anonymity can be lifted, the address of the repository will be
+published.
+
+The package can be installed locally in R from file
+\texttt{canaccessR\_0.0.1.tar.gz} in this way:
+
+\begin{verbatim}
+install.packages("path_to_file/canaccessR_0.0.1.tar.gz", 
+  repos = NULL, 
+  type="source")
+\end{verbatim}
+
+The file is available for download using this link:
+
+\url{https://drive.google.com/drive/folders/1CEm-jHyp1VBB27Mq4UWslANhHDKxUOju?usp=drive_link}
 
 \bibliographystyle{sageh}
-\bibliography{bibfile.bib}
+\bibliography{bibfile}
 

### 2025-05-03 paper/Manuscript-Data-Package-cover.Rmd
Fixed some errors for submission to EPB

@@
 author:
-  - name: João Pedro Figueira Amorim Parga
+  - name: João Figueroa Amorim PARGA
     email: joao.parga@mail.utoronto.ca
@@
     footnote: 1
-  - name: Anastasia Soukhov
+  - name: Anastasia SOUKHOV
     email: soukhoa@mcmaster.ca
@@
     orcid: 0000-0003-4371-4831
-  - name: Robert Nutifafa Arku
+  - name: Robert Nutifafa ARKU
     email: rn.arku@mail.utoronto.ca
@@
     orcid: 0000-0002-2018-886X
-  - name: Christopher Higgins
+  - name: Christopher HIGGINS
     email: cd.higgins@utoronto.ca
@@
     orcid: 0000-0002-3551-7750
-  - name: Antonio Páez
+  - name: Antonio PÁEZ
     email: paezha@mcmaster.ca

### 2025-05-03 paper/Manuscript-Data-Package-cover.tex
Fixed some errors for submission to EPB

@@
 
-\author{João Pedro Figueira Amorim Parga\affilnum{}, Anastasia
-Soukhov\affilnum{}, Robert Nutifafa Arku\affilnum{}, Christopher
-Higgins\affilnum{}, Antonio Páez\affilnum{}}
+\author{João Figueroa Amorim PARGA\affilnum{}, Anastasia
+SOUKHOV\affilnum{}, Robert Nutifafa ARKU\affilnum{}, Christopher
+HIGGINS\affilnum{}, Antonio PÁEZ\affilnum{}}
 

### 2025-05-03 paper/Manuscript-Data-Package.Rmd
Fixed some errors for submission to EPB

@@
   - times
-bibliography: bibfile
+bibliography: bibfile.bib
 bibliographystyle: sageh
 output:
-  #rticles::sage_article:
-  #  keep_tex: yes
-  #  fig_caption: yes
-  word_document
+  rticles::sage_article:
+   keep_tex: yes
+   fig_caption: yes
+#  word_document
 header-includes: \usepackage{float}
@@
 library(glue)
-library(gt)
+#library(gt)
 library(gtsummary)
+library(kableExtra)
 # library(summarytools)
@@
 We now present some descriptive statistics from the travel time matrices contained in the {canaccessR} package.
-Table @tbl-table_1 summarizes the travel time estimates.
+Table \ref{tab:table_1} summarizes the travel time estimates.
 The travel time objects contain, in total, 97,784,850 origin-destination pairs (observations) from population to employment locations for all the DAs in the sample and 18,519,897 pairs from population to grocery stores.
@@
 ```{r, fig.pos="H"}
-#| label: tbl-table_1
-#| tbl-cap: "Descriptive statistics of the travel time matrices from the {canaccessR} package."
 #| echo: false
@@
 sk_full %>%
-  gt() %>% 
-  fmt_number(use_seps = TRUE, decimals = FALSE) |>  
-  tab_options( 
-    table.font.size = 10,
-    table.width = pct(100) 
-    )
+  kableExtra::kable("latex",
+                    booktabs = TRUE,
+                    digits = 2,
+                    caption = "\\label{tab:table_1}Descriptive statistics of the travel time matrices from the {canaccessR} package.") %>%
+kable_styling(latex_options = c("striped", "scale_down"))
 
@@
 This section presents a usage example of the package through a visual representation of its data.
-In Figure @fig-travel_time_emp_grc_plot, we show the spatial distribution of the travel time matrices for the metropolitan region of Montréal.
+In Figure \ref{fig:fig-travel_time_emp_grc_plot}, we show the spatial distribution of the travel time matrices for the metropolitan region of Montréal.
 In it, we see the median travel time from each DA to employment (left) and to grocery stores (right).
@@
 
-```{r, fig.pos="H"}
-#| label: fig-travel_time_emp_grc_plot
-#| fig-cap: "Estimated median travel time (minutes) per Dissemination Area to jobs (left) and grocery stores (right).
-Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016)."
+```{r, fig.pos="H", fig.cap={"\\label{fig:fig-travel_time_emp_grc_plot}Estimated median travel time (minutes) per Dissemination Area to jobs (left) and grocery stores (right).
+Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016)."}}
+##| label: fig-travel_time_emp_grc_plot
+##| fig-cap: "Estimated median travel time (minutes) per Dissemination Area to jobs (left) and grocery stores (right).
+Public transit travel times are #calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics #Canada, 2016)."
 #| out-width: 100%
@@
 A more thorough example of the package's use can be found in the School of Cities' recent report on Canada's Urban Infrastructure Deficit.
-In its 11th Chapter, we use the travel time matrices to estimate accessibility metrics to jobs and grocery stores before and after the pandemic [@pargaDemocraticAccessOur2024].
-We then compare how changes affected groups differently according to their spatial distribution and income level, thus making explicit the connection between the package's information and matters of equity in transportation.
+In its 11th Chapter, the travel time matrices in the package are used to estimate accessibility metrics to jobs and grocery stores before and after the pandemic [@pargaDemocraticAccessOur2024].
+This work also compares how changes affected groups differently according to their spatial distribution and income level, thus making explicit the connection between the package's information and matters of equity in transportation.
 The report is freely available for download at the State of Cities Summit [website](https://stateofcitiessummit.ca/report).

### 2025-05-03 paper/Manuscript-Data-Package.tex
Fixed some errors for submission to EPB

@@
 \usepackage{booktabs}
-\usepackage{caption}
 \usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
 \usepackage{colortbl}
-\usepackage{array}
-\usepackage{anyfontsize}
-\usepackage{multirow}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
@@
 We now present some descriptive statistics from the travel time matrices
-contained in the \{canaccessR\} package. Table \citet{tbl-table_1}
+contained in the \{canaccessR\} package. Table \ref{tab:table_1}
 summarizes the travel time estimates. The travel time objects contain,
@@
 
-\begin{table}[t]
-\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{tabular*}{1\linewidth}{@{\extracolsep{\fill}}clrrrrrr}
+\begin{table}
+\centering
+\caption{\label{tab:unnamed-chunk-1}\label{tab:table_1}Descriptive statistics of the travel time matrices from the {canaccessR} package.}
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\begin{tabular}[t]{llrrrrrr}
 \toprule
-Study Region Name & Destination & Observations & Mean & Sd & P25 & P50 & P75 \\ 
-\midrule\addlinespace[2.5pt]
-All regions & Employment & 97,784,850 & 78 & 25 & 59 & 80 & 99 \\ 
-Toronto & Employment & 41,038,062 & 82 & 25 & 64 & 85 & 103 \\ 
-Montréal & Employment & 30,091,411 & 76 & 25 & 57 & 77 & 97 \\ 
-Vancouver & Employment & 12,254,478 & 76 & 25 & 58 & 77 & 96 \\ 
-Calgary & Employment & 3,022,244 & 75 & 23 & 59 & 75 & 91 \\ 
-Ottawa & Employment & 2,936,912 & 74 & 24 & 56 & 74 & 93 \\ 
-Edmonton & Employment & 2,057,486 & 70 & 23 & 53 & 70 & 87 \\ 
-Québec City & Employment & 1,473,702 & 72 & 25 & 54 & 72 & 93 \\ 
-Winnipeg & Employment & 1,574,205 & 62 & 22 & 46 & 61 & 76 \\ 
-Hamilton & Employment & 2,089,172 & 82 & 28 & 61 & 87 & 106 \\ 
-Waterloo & Employment & 552,826 & 71 & 27 & 49 & 69 & 93 \\ 
-London & Employment & 415,174 & 61 & 21 & 46 & 60 & 74 \\ 
-Halifax & Employment & 279,178 & 66 & 25 & 47 & 65 & 84 \\ 
-All regions & Groceries Stores & 18,519,897 & 76 & 26 & 56 & 76 & 97 \\ 
-Toronto & Groceries Stores & 8,512,874 & 80 & 25 & 61 & 82 & 101 \\ 
-Montréal & Groceries Stores & 2,993,965 & 73 & 26 & 53 & 74 & 94 \\ 
-Vancouver & Groceries Stores & 4,540,106 & 73 & 25 & 53 & 74 & 93 \\ 
-Calgary & Groceries Stores & 502,757 & 70 & 22 & 54 & 70 & 86 \\ 
-Ottawa & Groceries Stores & 619,791 & 72 & 24 & 55 & 72 & 91 \\ 
-Edmonton & Groceries Stores & 324,030 & 68 & 23 & 51 & 67 & 84 \\ 
-Québec City & Groceries Stores & 234,600 & 70 & 26 & 51 & 70 & 89 \\ 
-Winnipeg & Groceries Stores & 362,566 & 58 & 21 & 43 & 57 & 72 \\ 
-Hamilton & Groceries Stores & 281,118 & 84 & 29 & 60 & 91 & 110 \\ 
-Waterloo & Groceries Stores & 44,726 & 66 & 28 & 44 & 62 & 90 \\ 
-London & Groceries Stores & 52,617 & 55 & 19 & 42 & 55 & 67 \\ 
-Halifax & Groceries Stores & 50,747 & 63 & 26 & 42 & 63 & 83 \\ 
+Study Region Name & Destination & Observations & Mean & Sd & P25 & P50 & P75\\
+\midrule
+\cellcolor{gray!10}{All regions} & \cellcolor{gray!10}{Employment} & \cellcolor{gray!10}{97784850} & \cellcolor{gray!10}{78.06} & \cellcolor{gray!10}{25.18} & \cellcolor{gray!10}{59} & \cellcolor{gray!10}{80} & \cellcolor{gray!10}{99}\\
+Toronto & Employment & 41038062 & 81.99 & 24.74 & 64 & 85 & 103\\
+\cellcolor{gray!10}{Montréal} & \cellcolor{gray!10}{Employment} & \cellcolor{gray!10}{30091411} & \cellcolor{gray!10}{76.18} & \cellcolor{gray!10}{25.26} & \cellcolor{gray!10}{57} & \cellcolor{gray!10}{77} & \cellcolor{gray!10}{97}\\
+Vancouver & Employment & 12254478 & 75.90 & 24.87 & 58 & 77 & 96\\
+\cellcolor{gray!10}{Calgary} & \cellcolor{gray!10}{Employment} & \cellcolor{gray!10}{3022244} & \cellcolor{gray!10}{74.57} & \cellcolor{gray!10}{22.60} & \cellcolor{gray!10}{59} & \cellcolor{gray!10}{75} & \cellcolor{gray!10}{91}\\
+\addlinespace
+Ottawa & Employment & 2936912 & 74.11 & 24.16 & 56 & 74 & 93\\
+\cellcolor{gray!10}{Edmonton} & \cellcolor{gray!10}{Employment} & \cellcolor{gray!10}{2057486} & \cellcolor{gray!10}{70.11} & \cellcolor{gray!10}{23.37} & \cellcolor{gray!10}{53} & \cellcolor{gray!10}{70} & \cellcolor{gray!10}{87}\\
+Québec City & Employment & 1473702 & 72.46 & 25.37 & 54 & 72 & 93\\
+\cellcolor{gray!10}{Winnipeg} & \cellcolor{gray!10}{Employment} & \cellcolor{gray!10}{1574205} & \cellcolor{gray!10}{61.75} & \cellcolor{gray!10}{21.58} & \cellcolor{gray!10}{46} & \cellcolor{gray!10}{61} & \cellcolor{gray!10}{76}\\
+Hamilton & Employment & 2089172 & 82.00 & 27.56 & 61 & 87 & 106\\
+\addlinespace
+\cellcolor{gray!10}{Waterloo} & \cellcolor{gray!10}{Employment} & \cellcolor{gray!10}{552826} & \cellcolor{gray!10}{70.77} & \cellcolor{gray!10}{26.62} & \cellcolor{gray!10}{49} & \cellcolor{gray!10}{69} & \cellcolor{gray!10}{93}\\
+London & Employment & 415174 & 60.99 & 20.98 & 46 & 60 & 74\\
+\cellcolor{gray!10}{Halifax} & \cellcolor{gray!10}{Employment} & \cellcolor{gray!10}{279178} & \cellcolor{gray!10}{65.96} & \cellcolor{gray!10}{25.45} & \cellcolor{gray!10}{47} & \cellcolor{gray!10}{65} & \cellcolor{gray!10}{84}\\
+All regions & Groceries Stores & 18519897 & 75.54 & 25.66 & 56 & 76 & 97\\
+\cellcolor{gray!10}{Toronto} & \cellcolor{gray!10}{Groceries Stores} & \cellcolor{gray!10}{8512874} & \cellcolor{gray!10}{79.56} & \cellcolor{gray!10}{25.34} & \cellcolor{gray!10}{61} & \cellcolor{gray!10}{82} & \cellcolor{gray!10}{101}\\
+\addlinespace
+Montréal & Groceries Stores & 2993965 & 73.14 & 25.68 & 53 & 74 & 94\\
+\cellcolor{gray!10}{Vancouver} & \cellcolor{gray!10}{Groceries Stores} & \cellcolor{gray!10}{4540106} & \cellcolor{gray!10}{72.84} & \cellcolor{gray!10}{25.37} & \cellcolor{gray!10}{53} & \cellcolor{gray!10}{74} & \cellcolor{gray!10}{93}\\
+Calgary & Groceries Stores & 502757 & 70.03 & 22.25 & 54 & 70 & 86\\
+\cellcolor{gray!10}{Ottawa} & \cellcolor{gray!10}{Groceries Stores} & \cellcolor{gray!10}{619791} & \cellcolor{gray!10}{72.19} & \cellcolor{gray!10}{24.41} & \cellcolor{gray!10}{55} & \cellcolor{gray!10}{72} & \cellcolor{gray!10}{91}\\
+Edmonton & Groceries Stores & 324030 & 68.07 & 23.41 & 51 & 67 & 84\\
+\addlinespace
+\cellcolor{gray!10}{Québec City} & \cellcolor{gray!10}{Groceries Stores} & \cellcolor{gray!10}{234600} & \cellcolor{gray!10}{69.66} & \cellcolor{gray!10}{25.79} & \cellcolor{gray!10}{51} & \cellcolor{gray!10}{70} & \cellcolor{gray!10}{89}\\
+Winnipeg & Groceries Stores & 362566 & 58.42 & 20.96 & 43 & 57 & 72\\
+\cellcolor{gray!10}{Hamilton} & \cellcolor{gray!10}{Groceries Stores} & \cellcolor{gray!10}{281118} & \cellcolor{gray!10}{83.61} & \cellcolor{gray!10}{29.15} & \cellcolor{gray!10}{60} & \cellcolor{gray!10}{91} & \cellcolor{gray!10}{110}\\
+Waterloo & Groceries Stores & 44726 & 66.06 & 27.62 & 44 & 62 & 90\\
+\cellcolor{gray!10}{London} & \cellcolor{gray!10}{Groceries Stores} & \cellcolor{gray!10}{52617} & \cellcolor{gray!10}{55.48} & \cellcolor{gray!10}{19.26} & \cellcolor{gray!10}{42} & \cellcolor{gray!10}{55} & \cellcolor{gray!10}{67}\\
+\addlinespace
+Halifax & Groceries Stores & 50747 & 63.16 & 26.39 & 42 & 63 & 83\\
 \bottomrule
-\end{tabular*}
+\end{tabular}}
 \end{table}
@@
 representation of its data. In Figure
-\citet{fig-travel_time_emp_grc_plot}, we show the spatial distribution
+\ref{fig:fig-travel_time_emp_grc_plot}, we show the spatial distribution
 of the travel time matrices for the metropolitan region of Montréal. In
@@
 \begin{figure}[H]
-\includegraphics[width=1\linewidth]{../figures/patch_tt_emp_grc} \caption{Estimated median travel time (minutes) per Dissemination Area to jobs (left) and grocery stores (right).
-Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016).}\label{fig:fig-travel_time_emp_grc_plot}
+\includegraphics[width=1\linewidth]{../figures/patch_tt_emp_grc} \caption{\label{fig:fig-travel_time_emp_grc_plot}Estimated median travel time (minutes) per Dissemination Area to jobs (left) and grocery stores (right).
+Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016).}\label{fig:unnamed-chunk-2}
 \end{figure}
@@
 of Cities' recent report on Canada's Urban Infrastructure Deficit. In
-its 11th Chapter, we use the travel time matrices to estimate
-accessibility metrics to jobs and grocery stores before and after the
-pandemic \citep{pargaDemocraticAccessOur2024}. We then compare how
-changes affected groups differently according to their spatial
-distribution and income level, thus making explicit the connection
-between the package's information and matters of equity in
+its 11th Chapter, the travel time matrices in the package are used to
+estimate accessibility metrics to jobs and grocery stores before and
+after the pandemic \citep{pargaDemocraticAccessOur2024}. This work also
+compares how changes affected groups differently according to their
+spatial distribution and income level, thus making explicit the
+connection between the package's information and matters of equity in
 transportation. The report is freely available for download at the State
@@
 \bibliographystyle{sageh}
-\bibliography{bibfile}
+\bibliography{bibfile.bib}
 
